### 2014-04-04 paper.tex
adding parts

% v2-acmsmall-sample.tex, dated March 6 2012
% This is a sample file for ACM small trim journals
%
% Compilation using 'acmsmall.cls' - version 1.3 (March 2012), Aptara Inc.
% (c) 2010 Association for Computing Machinery (ACM)
%
% Questions/Suggestions/Feedback should be addressed to => "acmtexsupport@aptaracorp.com".
% Users can also go through the FAQs available on the journal's submission webpage.
%
% Steps to compile: latex, bibtex, latex latex
%
% For tracking purposes => this is v1.3 - March 2012

\documentclass[prodmode,acmtecs]{acmsmall} % Aptara syntax

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{mindmap,trees}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

% Metadata Information
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2010}
\acmMonth{3}

% Document starts
\begin{document}

% Page heads
\markboth{K. Maheshwari et al.}{A Survey of Scientific Workflow Management Systems}

% Title portion
\title{A Survey of Scientific Workflow Management Systems}
\author{KETAN MAHESHWARI
\affil{University of Nice, Sophia Antipolis}
JOHAN MONTAGNAT
\affil{CNRS}
}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

\begin{abstract}

In this work we conduct a literature study of the key features of the
contemporary scientific workflow environments. The criteria for study are
defined. An overview of studied scientific workflow environments is given
followed by the identification and description of the features offered by each
of the environment. A tabulated summary for each feature gives a high-level
view of the offered features. Finally, a taxonomy is proposed placing these
features into respective places in a hierarchy.

\end{abstract}

%\category{C.2.2}{Computer-Communication Networks}{Network Protocols}

%\terms{Design, Algorithms, Performance}

%\keywords{Wireless sensor networks, media access control, multi-channel, radio interference, time synchronization}

\acmformat{Ketan Maheshwari, Johan Montagnat
, 2014. A survey of scientific workflow management systems.}
% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

\begin{bottomstuff}
This work is supported by the ANR-CNRS, under
grant contract CNRS-0435060.

Author's addresses: Ketan Maheshwari, Argonne National Laboratory; Johan Montagnat, I3S, Sophia Antipolis
\end{bottomstuff}

\maketitle

%Workflows considered:
% MOTEUR2
% Taverna
% Swift
% Pegasus
% Triana
% Galaxy
% Makeflow
% Dryad
% Science pipes
% GWES
% Spark
% Linda
% DaGue

\section{Introduction}

\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
are, in principle similar to the traditional programming environments
(programming language, interpreter or compiler, and runtime) albeit with a
larger scale of data and a higher abstraction of computation. The similarity is
such that scripting languages are sometimes advocated for programming at this
level. However, traditional scripts do not deal with parallelism and dataflow
aspects of workflow paradigm that makes them a limited contenders. 
%
%Add more content here to beef up the context.
%
The goals of a workflow management system for scientific computing are as
follows:

\begin{enumerate}
\item \textbf{User Interaction.} Adaptation to new methods of computing is not
easy for domain science users. Such users often need special assistance in
order to translate their computation requirements into concrete computing
entities.  SWEs are an ideal platform enabling users to perform many tasks
conveniently.  These tasks include composing and marshalling of execution,
browsing and selecting data, visualizing results. Furthermore, dissemination of
methods and results to communities and collaborators are some of the activities
that SWEs make convenient for end-users.

\item \textbf{Resource Access and Utilization.} Computational platforms and
 access modalities change rapidly with upgradation of existing technologies and
adaptation of new techniques. Clouds are the latest examples of new techniques
of computational platform access while GPGPUs examples in upgradation of
existing processing environments. Additionally, with the advent of fast and
dedicated networks such as ESNet and federated visualization resources, remote
visualization is gaining prominence among scientific users. SWEs play important
role as platforms to give transparent and hassle free access to such novel and
multiple environments to users by hiding the technicalities of access methods.

\item \textbf{Repeatable Execution.} Scientific users often require a
repeatable execution of same workflow for reproduction of results and
demonstration of methods. SWEs can record and save workflows, associated
parameters, source data and historical results to aid in such activities.
Additionally, SWEs facilitate an incremental development of such methods
leading to augmentation of past efforts to current and future developments.

\item \textbf{Provenance and MetaData.} Historical data and metadata has been
crucial entities in many research fields. They help connecting results from
a multitude of research aspects. A provenance trace of computation can help
checkpoint methods and reproduce and modify partial computation. Automated
observations of deviations in scientific phenomena from provenance data is one
of the key triggers for new scientific discoveries. Fields of studies involving
genomic and protiens are examples of domains where provenance and metadata
plays key roles. SWEs can be integrated or extended with metadata and
provenance techniques rendering powerful computing traceability features.

\item \textbf{Sharing and Collaboration.} Workflow managers are often targetted
towards a community of users. For instance, the Galaxy workflow system is
targetted specifically for the bioinformatics users. Galaxy attracts many users
from the same community thus promoting sharing and collaboration via a common
computational platform. Another such example is Taverna which is geared towards
sharing and collaborative workflow oriented scientific computation enabling
users to discover new services introduced to a community.

\end{enumerate}

As a consequence to these design goals, the SWEs differ in many aspects with
respect to traditional programming environments. The following list identifies
the major differences between traditional and the SWE programming
paradigm:

\begin{enumerate}
%
\item \textbf{Scalability and Optimization.} SWEs are more likely to involve extensive usage of distributed resources than conventional programs. Owing to the large-scale experiments, for which scientific users turn to automation, their demands for computational power often exceeds that of a single personal computer. Such applications are often composed of multiple, distinct computational stages making them well-suited to the workflow paradigm. These properties also make them ideal candidates for a large-scale distributed computing environment. The case for suitability of SWEs on large-scale distributed infrastructures has been well-argued upon in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to ensure scalable and optimized performance is vital.
%
\item \textbf{Data Management.} The scale of data transfer is much higher in SWEs from one execution site to others as well as among storage resources while it is lower (ranging from variables to small data structures) in programs.
%
\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).} Building SWE's interfaces to distributed systems is a complex task. The distributed nature of resource utilization and asynchronous execution of workflow requires a set of interdependent parameters to be configured correctly. Further complexity is added because of complex dataflow requirements, interactions with diverse tools and linking with legacy application codes.
%
\item \textbf{Expressiveness.} Programming languages are inherently control flow oriented where data flow is a secondary aspect. In fact, the programmer must explicitly make sure that correct the inputs are available to instructions and are in the right format. A program proceeds as the execution of its statements from beginning to end and completes its execution in a predefined sequence. However, in SWEs, the flow is often determined by the data availability for execution. This model of computation is known as dataflow computing. A workflow thus does not follow a fixed order of execution and different execution sequences are possible which are dominated by dataflow. The unit of work in SWEs is at a higher level of granularity. A unit of execution is coarse-grained, and often termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE community. This notion of high-level granularity drives the depth to which a workflow can be expressed in a given language form. In the rest of this paper, this unit of work will be referred to as \textit{processor}. On the other hand, a unit of execution in most programming languages is a much simpler and finer executable statement that accomplishes a small computational task. This presents a whole new set of requirements for expressing the processors and how they should interact and operate upon data.
%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain experts to utilize the computational power. This makes it essential for them to have a friendly, and intuitive user-interface for novice and non-expert users. For this reason, SWE user interfaces are dominated by the visual paradigm of programming. On the other hand, traditional programs are most often written by expert programmers. Thus, they have a text-based composition environment albeit with visually assisting programming environments such as IDEs. However, they are seldom to the extent that a non-expert user can graphically compose programs.

\end{enumerate}

We concede that the above is not an exhaustive list of differences in features.
However, it is reasonably inclusive and we believe it covers the breadth of
aspects we cover in the current work. We do not include other interesting SWE
features such as provenance and interoperability in this analysis as they are
out of scope of the objectives of the current paper.

\section{Architecture of a Generic SWE}
Figure \ref{fig:wfms} shows a high-level architecture of a SWE. It shows key components and their interaction in a SWE. These components are the focus of interest for the present work. A \textit{user interface} interacts with the \textit{workflow engine} using commands and messages. The \textit{workflow language} becomes an intermediate representation for communication between user interface and the \textit{workflow engine}. The workflow engine acts as a compiler/interpreter and scheduler for the workflow tasks and interfaces with the underlying \textit{distributed resources}. It also monitors the workflow progress and collects the intermediate and final results.
%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing infrastructure such as clusters, supercomputers, clouds, grids. Such environments are in use today for production level work in different scientific domains worldwide. They do the basic task of linking inter-related processors of a multi-staged scientific computation. However, they differ in the features they provide based upon the scientific domain they offer service for, and the computational model they follow. Often, one of the main reasons for such difference is the specialized components and requirements for a particular domain. Owing to this fact, a given workflow system dominates a particular area of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora of SWEs suitable for general purpose workflow style computations.
% Ketanmark

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs, studying many existing systems for distributed computing. This taxonomy
is based, to a large extent on the capabilities of the workflow enactors. It
provides only a limited insight to the SWE aspects considered in the present
paper, especially the workflow languages expressiveness and SWEs interactions
with the distributed computing infrastructures. A more recent work by Goderis
and co-authors \cite{goderis-brooks-etal:2009} studies a topic more closely
related to the workflow language expressiveness: it compares the models of
computations of various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. The recent survey from Stratan et. al.
\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
account the latencies involved during the enactment mechanisms. A small survey
and analysis of file-access characteristics of Data-intensive workflows has
been done by Shibata et. al.~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in \cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,delaitre-kiss-etal:2005,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR DAGMan
\cite{dagman} to abstractions for parallel computations such as Petri nets
\cite{alt-hoheisel:2005}, meta-models \cite{nemo-glatard-etal:2007},
data-driven languages such as Scufl \cite{turi-missier-etal:2007} and scripting
languages such as SwiftScript \cite{swift}. Glatard's PhD thesis
\cite{glatard:2007} presents a taxonomy of SWE languages classifying them into
functional, graph, service-oriented and executable representations. A detailed
classification and taxonomy of parallelism patterns in Grid Workflows has been
presented by Pautasso and Alonso in \cite{pautasso1}. This classification is
more generic while the one presented in Glatard's thesis has emphasis on
service based workflows.  Each of these approaches can be defended through some
aspects well covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of applications logic for nonexpert users and scripting is
extensively used by programmers for prototyping, etc.

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA
DAG\footnote{http://graal.ens-lyon.fr/diet/workflow.html}
\cite{caron-desprez:2005}, engine is one example of a DAG generator engine. MA
DAG has evolved from a regular DAG generator to a more abstract DAG generator
laced with control structures. MA DAG benefits from the workflow scheduling
mechanisms included in the middleware. Due to the impossibility to produce
complete DAGs owing to the presence of control structures such as conditionals
and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
possible, is produced as soon as possible. In spite of the presence of many
different approaches to workflow languages, new application requirements
continue to pose workflow expressiveness challenges. The community is yet to
realize a truly generic workflow language that can express a wide variety of
flow patterns encountered in scientific domains.

\section{Objectives and Organization}
The present work analyzes contemporary SWEs used in the context of e-science to
find out the capabilities and key features in terms of the characteristics
described in the introduction part of this paper. The rest of this paper is
organized as follows: Section \ref{sec:overview} presents an overview of the
key SWEs studied in this work. Section \ref{sec:scal} discusses the
optimization features and resulting scalable performance as observed in the
SWEs studied. Section \ref{sec:data} studies the data management and
description aspects of the SWEs. Section \ref{sec:inter} presents a study of
SWEs approaches to interface the large-scale computing infrastructures. Section
\ref{sec:lang} details the language features and expressiveness of the SWEs
studied. Section \ref{sec:usable} discusses the usability aspects of the SWEs.
Section \ref{sec:sum} presents a summary and a classification of all the
aspects of SWEs studied in this work. Throughout the paper, we also present
tables summarizing the aspects of SWEs considered in respective sections. It
was not possible to obtain an exhaustive information about all the SWEs
considered in this work since we rely on the peer reviewed publications for
information on a given SWE. Some features, such as extendability of SWE
strongly depend on the extension points designed in the code; consequently, we
consider it present in only those SWEs that have a documented description or
demonstration of the given feature.

\section{Overview of the Workflow Environments} \label{sec:overview}
In this section we present an overview of the Workflow Environments studied.
The paragraphs headers are organized by the name of the workflow environment
followed by the language they chiefly support separated by a forward slash (/).
The cases where the workflow language is not specifically defined or conversely
where there are multiple engines supporting a generic workflow language are
denoted by an asterisk (*) or a hyphen (-) at the respective places.
\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{oinn-addis-etal:2004} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment engine. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.

A Taverna workflow consists of a collection of processors connected by data
links, which establish a dependency between the output(s) of a processor and
the input(s) of another. Apart from data links, the processors may also be
connected through coordination links that make a processor invoke when a
previous processor ends its execution.  The processors are of different kinds
depending on the application code to be invoked: typically, processors are Web
Services, with input and output \textit{ports} that correspond to the
operation's invocations and response sequences defined in the service`s WSDL
interface.  Java classes and local scripts can be used as workflow processors,
as long as they expose their signature as input and output ports as required by
the model.  The engine orchestrates the execution of the processors in a way
that is consistent with the dependencies, and manages the flow of data through
the processors' ports. The overall workflow execution is data-driven, with a
push model: a processor's execution is started as soon as all of its inputs are
available.

Taverna is a flexible and extendible workflow environment
\cite{missier-soiland-reyes-etal:2010,sroka2009a} in that extending the
engine's functionality is facilitated through a Service Provider Interface
(SPI). A third party developer intending to extend a functionality must write
its logic as an SPI and publish it as an extension SPI. This may manifest
itself in different ways depending upon the intended requirements of the
developer, examples being as a task in the service panel or internally within
the enactor as additional functionality.

The Taverna engine implements a sophisticated mechanism for enactment of
parallel workflows. These mechanisms manifest in the form of pipelines and an
advanced superpipelined enactment of workflows. In the case of pipelined
execution, a simple pipeline of tasks are overlapped into a back to back
execution for a single data set. Such a mechanism induces multiple downstream
workflow processors running in parallel for a dataset. In the case of
superpipelines, multiple such pipelines instances run for multi-datasets
providing a high degree of parallelism limited only by the availability of
underlying hardware resources.

Taverna is not natively interfaced with a Grid infrastructure but the same can
be achieved with the help of extending the engine through one of its SPIs.

\paragraph{MOTEUR/scufl} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
MOTEUR facilitates executing workflows consisting of remote web- and
grid-services. It exploits service parallelism at workflow level and data
parallelism for multiple datasets.

MOTEUR has sophisticated data management mechanisms within the workflow definition. It provides a diverse set of iterations strategies well adapted for different combinations of input datasets at the processor. An iteration strategy allows different combinations of the data arriving at a processor along its multiple inputs. Running legacy application code on grids has been eased by MOTEUR by providing a generic grid-service wrapper for these applications \cite{glatard-montagnat-etal:2008a}. Grid-services thus created can be readily deployed on the grids by the MOTEUR workflow enactor. MOTEUR workflows are written in the scufl language and are compatible with those written for the Taverna SWE. It also supports an extension of scufl language called the Gwendia language, developed at the same laboratory. 
%
\paragraph{Makeflow} Business Process Execution Language (BPEL) \cite{oasis:2007} is a workflow language pioneered by IBM Corporation and used predominantly to define business workflows. It is now a broad standard for business workflow applications. BPEL is an open language and commercial enactors are available that provide enactment functionality for the BPEL workflows. There has been attempts to use BPEL or its extended version in the scientific applications \cite{slominski:2007,bpel}. However, the language does not suit well to the data-intensive, grid workflows and has a limited adoption in the scientific community \cite{tan-missier-etal:2010,barga-gannon:2007,scherp-hoing-etal:2009,emmerich-butchart-etal:2005}. Apart from that, a lack of an open-source workflow engine with graphical interface also contributes to its limited acceptance in academic research. However, BPEL has been famously unsuitable for data intensive scientific applications  BPEL lacks different user-defined strategies to effect multiple invocations of each processor according to the application semantics. BPEL also lacks data streams and implicit parallelism forcing a user to explicitly define parallelism using one of its parallel processing constructs such as \texttt{foreach}.
%
\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based SWE developed at the Cardiff University. It consists of a workflow composition GUI, a framework to integrate workflow with the different execution environments and middlewares and a set of predefined services written as java components. A Triana workflow is expressed as an XML grammar similar to a WSDL specification for webservices \cite{taylor-wand-etal:2005}. A highlight of Triana is that it provides ready-to-use interface for a wide variety of middleware and distributed infrastructures including P2P, grids and web-services frameworks. A Triana workflow provides dataflow as well as control flow support. Triana allows conditional branching and looping in its workflows.
%
\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at the University of California. Kepler workflows follow an actor-director metaphor where an actor represents a unit of execution and the director represents the paradigm for executing the workflows consisting of the actors. Depending upon the types of directors, a Kepler workflow may exhibit a different type of execution behavior. A workflow designer can chose from among directors based upon the purpose of a workflow. Few of these include Process Network (PN) for a parallel execution at the workflow level, Synchronous Data Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its workflow representation. MoML is an extended XML syntax providing a rich vocabulary to compose Kepler workflows. A workflow processor is called an `actor' in Kepler. 
%
\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the University of North Carolina, Wilmington, facilitates a graphical composition and execution of DAG-based workflows. Like Kepler, GridNexus is also based on Ptolemy II framework. GridNexus workflows are expressed in an XML-based scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a customizable and extensible scripting language allowing users to write their own workflow processor types. GridNexus workflows are grid oriented and can be deployed on grid resources accessed through a persistent WS-RF (Web Services Resource Framework) interface.
%
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed computing framework that can map abstract workflows as DAGs to the concrete underlying resources using intelligent and dynamic scheduling policies. A Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in XML). DAX describes the logical workflow components, their inputs and outputs and their links and dependencies in a Pegasus workflow. These kind of workflows has been shown to enact optimally using Pegasus' smart memory and disk management techniques. Pegasus manager buffers and maintains a fix-sized window of executable tasks before accepting new tasks.
%

\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific workflow environment developed at the University of Chicago. A Swift workflow specified in the SwiftScript scripting language is translated to an intermediate XML-based representation and enacted by the Karajan workflow enactment engine. The SwiftScript language is proposed as a workflow language founded on a scripting approach programmers are familiar with. To make the language adapted to grids, the inventors of SwiftScript generalized the use of futures \cite{Halstead85multilisp:a} for every variables defined in the script. Futures are non-blocking assignment variables, using a proxy to ensure immediate execution of assignment and performing lazy blocking on variable value accesses only. The systematic use of futures for every variables in the language makes it completely data-driven: the execution progresses asynchronously as long as a data access is not blocking. The availability of data enables blocked thread to restart execution. The beauty of futures is to make this process completely hidden from the programmer who can use a scripting language he is familiar with without ever manipulating explicit parallel programming constructs. Swift facilitates a highly asynchronous workflow execution. Swift supports advanced workflow constructs such as parallel loops and conditionals. Furthermore Swift supports processing of explicitly defined arrays of data. Swift workflows take advantage of POSIX-style shared file system to reuse the data produced by other processors.
% Put this somewhere else, either in the analysis or may be in the conclusions and summary.
%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language developed as a part of the current work is similar in spirit to the SwiftScript. Similar to SwiftScript, gscript adopts futures to express asynchronous execution in a script-based language. A SwiftScript workflow contains statically typed datasets that can be specified using ``annotations'' in the script. Unlike SwiftScript, gscript does provide user-defined strategies in order to effect data combinations in the workflow. SwiftScript provides for fault-tolerance and exception handling at run-time which is lacking currently in gscript. The underlying datasets organization in swift is represented using an XDTM XML-representation. This means that effectively, a workflow is SwiftScript plus the XDTM data representation. A workflow is simpler and compact in gscript as its data description is embedded within a workflow representation. SwiftScript supports creation of nested-workflows in the form of compound procedures. Conditional and Loop execution is supported using the if and for/while statements in both the languages.
%
\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE) \cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck. ASKALON is described as a Scientific PSE capable of composing, hosting, executing, sharing and reusing scientific workflows written in the Abstract Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is capable of expressing scientific application components chains at a high-level of abstraction. A concrete workflow consists of the abstract workflow combined with the resources and target execution environment binding the workflow. The graphical interface of AGWL is based upon the UML (Unified Modeling Language) diagrams.
%
\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service (GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First institute. It implements a dynamic workflow composition and enactment model based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}. GWES enables key workflow development and deployment lifecycle phases including visual composition, data-staging, Grid middleware interfacing and execution monitoring. Currently, GWES is intensively being used in the field of Medical Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany. It has been successfully interfaced with the German D-Grid infrastructure.
%

\paragraph{The Galaxy Portal/JSON} The P-GRADE Portal \cite{kacsuk-sipos:2005} developed at the MTA SZTAKI Computation and Automation Institute, Budapest, provides an access to grid computational and storage resources from a web browser through a specialized workflow oriented grid portal. A workflow development environment interfaces the users through this web based portal. The portal interfaces the multiple grid computing middleware and storage resources as a backend component of the environment. A highlight of P-GRADE portal is that it provides access to multiple grids and scientific communities in a single environment leading to a highly collaborative inter-community, inter-grid research environment. An older version of P-GRADE used the condor-Dagman as workflow engine while the recent version uses its own in-house developed workflow engine called Xen.
%
\paragraph{DaGue/--} Bioflow \cite{bioflow} is a query based workflow language. It is motivated by the fact that bioinformatics applications are query based applications that benefit from the complex database queries or saved database `views'. A typical Bioflow workflow pipelines a sequence of filter queries in order to get the desired results. Bioflow is mainly utilized in the area of bioinformatics and querying genome databases. Additionally, a Bioflow workflow provides mechanisms for data manipulation and updation in the existing databases. It also includes advanced features such as parallel processing of a set of mutually independent database queries. The execution model is based on the popular database query language SQL (Structured Query Language). Processes can be defined in Bioflow which are processor counterparts in workflow paradigm and similar to stored-procedures in SQL jargon.

Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical representation of bioflow workflows. A Vizbuilder workflow links to an underlying bioflow engine and provides an interface to submit the bioflow workflows for execution.
%
\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative workflow language that is inspired by the functional programming paradigm. A Martlet workflow conveniently enables distributed parameter sweep applications wherein the language intuitively defines the computation and a matching with the distributed resources can be carried out at workflow execution time. List processing applications where a sequence of application components work over an array of data in a Multi-data multi-instruction style setup is well-suited to the Martlet's model of computation.
%
\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining system for data intensive applications developed at the University of Konstanz. It is a java-based platform with a Graphical Interface that is based on eclipse (\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists of processors belonging to categories such as readers, manipulators, learners, predictors, etc. Each processor has a configuration dialog associated with it that allows user to set its parameters. Processors are connected through data and/or control connections. A highlight of knime is support for shared- and distributed-memory workflow enactment and support for parallel workflow loops and subworkflows. The loops in Knime manifest themselves in the form of subworkflows (called metanodes in Knime) by a repeated enactment of the same subworkflow.
%
\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based script oriented workflow language used to encode parallel pipelines. StarFlow handles data as well as control flow using a combination of static, dynamic and user annotations analysis. A parallelization engine uses data and control dependencies in a workflow in order to build a parallel profile of the workflow. StarFlow has been shown to be successfully interface with the SunGrid \cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}. The key feature of StarFlow is its capability to execute workflows to maximum possible parallelism. While StarFlow provides a script-based compact and semiautomatic workflow composition environment, currently it lacks a sophisticated GUI for workflow composition.

\paragraph{Galaxy/--} Galaxy \cite{blankenberg-vonkuster-etal:2010} is a bioinformatics domain SWE based framework developed at the Penn State University, Pennsilvania. It provides a webportal based environment to build complex pipelines from an available repository of tools (\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided by Galaxy using their own Galaxy server or a web-server exposing the webservice descriptions through public WSDLs. One of the major use of Galaxy is to achieve a mix and match of data analysis tools in order to analyze a vast repository of genome based database.

% == ENACTOR SCALABILITY AND OPTIMIZATION ==
\section{Enactor Scalability and Optimization} \label{sec:scal}
%We have results comparing \MOTEUR and Taverna performance!
%For Pegasus: Scaling up workflow based applications.
A simple definition of scalability in the context of SWE is that the
performance of the workflow engine increases linearly with the increasing
number of similar tasks.  Here, similar is an important term that means the
nature of tasks should be the same in terms of processing and data handling.
Depending upon the lag or increase in the linearity we say a workflow enactor
scales up or down.  In the present section, we focus on the scalability aspects
of the core workflow engine. We do not consider the specific scalability issues
related to a distributed infrastructure such as network latencies and protocol
offsets. However, there have been efforts to thwart such issues at the workflow
level by combining workflow enactment techniques with those of distributed
computing.  One such example is the map/reduce framework \cite{mapreduce}.
Recently, the map/reduce framework, that aims at performing distributed
processing on large datasets, has gained a significant popularity in workflow
community. It is based on the master/worker metaphor of distributed processing.
It employs the commonly used \textit{map} and \textit{reduce} functions from
the functional programming languages. The map function chops a large chunk of
data into smaller pieces and distributes it to \textit{workers} which in turn
can also chop it to further smaller datasets forming a tree-like hierarchy. The
worker process the data and return the results back to their parents. The
reduce step collects the results from the workers and combines them to obtain
the final result. One of the main reason for the popularity of the map/reduce
framework for workflows is the recent attention to collections based datasets
and processing. There exists implementations of the map/reduce framework
written in different languages with enhancements and support services such as a
sophisticated query service. Hadoop (\textit{http://hadoop.apache.org/}) is one
such implementation which has already been interfaced by the Kepler
SWE~\cite{wang-crawl-etal:2009}.

Large-scale workflows with thousands of tasks often pose memory and disk
management difficulties leading to scalability issues.  However, the
performance of a workflow engine is dependent on several factors including the
type of tasks and approach for submitting tasks for execution.  There is not
much research done towards the enactor scalability baring a couple of studies
carried out by C. Stratan et. al. \cite{stratan-iosup-etal:2008} and Deelman
et. al. \cite{callaghan-deelman-etal:2010}.  One way to compare the scalability
of different workflow environment is to experiment with a similar set of
workflows on a given system. In addition, different SWEs interface to different
DCIs. This is a time consuming process.  In the current study we do not
directly compare the workflow engine scalability but we study the similarities
and differences in the techniques employed by workflow engines in order to
achieve scalability.

The software architecture of an enactor plays a key role in determining the
scalability for the workflow applications exhibiting a large number of tasks.
An architecture that takes into account the properties of the underlying
infrastructure could scale well in the face of growing number of workflow
tasks.

In the context of workflow enactors, scalability can be further classified in
terms of the computational and data transfer scalability. Computational
scalability could be expressed as the rate of increase in number of tasks
executed with respect to the increase in the elapsed time.

While the computational scalability of a large scale experiment depends upon
the underlying computing infrastructure, at the workflow enactor level, it
becomes important to be able to prepare and dispatch as many processor
instances as possible without loss in performance. Failing to do so for a
workflow engine can become a performance bottleneck and hamper the optimal
utilization of an available computing infrastructure.

Most workflow enactors exploit the parallelism at three-levels. First, at the
workflow-level wherein the independent branches of a workflow are enacted
concurrently. Second, at the data-level, where a same processor is instantiated
multiple times based upon the repeated operations on different datasets. Third,
the low-level parallelism that is encoded within the application components
independent of the workflow itself. At this level, the workflow enactor may
need to make a matching based upon the application type. For instance, a
workflow enactor may schedule the MPI-based processors on the resources that
indicate as possessing such capabilities.

In this section we analyze the scalability measures and techniques employed by
contemporary SWEs.

Illustrated below are some of the techniques used to achieve scalability in
various SWEs.

Taverna version 2.x introduced new measures in order to be able to submit a
large number of jobs simultaneously. It is able to dispatch large number of
concurrent instances of processors. An ability to implement superpipelined
\cite{hwang:1992} execution of workflow tasks leads to a maximum expression of
parallelism.

Within the Taverna engine, all the data involved in the execution is managed by
a data manager through an external database, and is addressed by the engine
only by \textit{reference}. This facilitates the management of data-intensive
workflows in a scalable way, as high volume data is only transferred into the
engine space when needed by processors that are unable to resolve external data
references on their own.

Taverna, Swift-Karajan and Pegasus all provide a means to limit the maximum
number of processor instances that can run concurrently through a builtin
parameter. In Taverna, this can be configured at a workflow level while in
Pegasus, this can be configured at the Condor \cite{thain-tannenbaum-etal:2005}
scheduler level. In Swift-Karajan system this parameter is regulated
automatically as part of a ``throttling parallelfor" construct. A similar
configurable parameter is provided by the Knime thread manager in order to
limit the size of the thread-pool employed to run a workflow.

Kepler optimizes the workflow execution through parallel and pipelined
execution. In addition to these, Kepler also provides dataflow analysis
optimization that involves data bypassing the processors in a workflow that
does not make use of that data. This is detected using a builtin Kepler
mechanism called ``static type propagation".

Another alternative is to optimize the enactment at runtime. This approach has
advantages including a better information about the actual data available. For
instance, Askalon does runtime branch prediction, parallel loop unrolling and
loop sequentialization.

The Pegasus workflow manager introduced subworkflows because of scalability
issues at the local host for a large number of nodes in a DAG. A hierarchical
representation helps it scale at the local planning level as well as at the
execution level on distributed resources.

Static planners in Pegasus translate a workflow graph into a location specific
DAGMan input file, adding stages for data staging, intersite transfer and data
registration. These planners can reduce processing of tasks for files that
already exist, select sites for jobs, and cluster jobs based on various
criteria. Pegasus performs graph transformation with the knowledge of the whole
workflow graph.

In Swift systems the structure of a workflow is constructed and expanded
dynamically as opposed to that of Pegasus' static planning. A static workflow
graph allows optimizations based on the global state of the workflow, whereas a
dynamic workflow structure allows adaptations to changing environments.
Depending upon the execution environment and access conditions to the available
resources, each approach can result in the overall optimization of a given
application.

Knime has a different execution strategies for shared memory Symmetric
Multiprocessing Systems (SMP) and distributed memory systems. In the case of
SMP, it uses the message passing library or MPI to distribute the independent
parts of a workflow to available processors. This approach results in efficient
use of available processors. However, for a large scale, the global memory
becomes a bottleneck hindering the scalability.

In the case of distributed memory and independent computational nodes connected
with highspeed network or the internet, Knime employs a Master/Worker metaphor
of computing. A Knime installation with GUI acts as a master node that
coordinates the bare Knime installations on the worker nodes. These worker
nodes are then sent the portions of a workflow by the master node. The master
node keeps track of the participating nodes by their unique IP addresses. At
the end of each execution, the master node collects the results from the worker
nodes.

\begin{table}[ht!]
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\textbf{SWE} / & \multicolumn{3}{c|}{\textbf{Parallelism}} & \multicolumn{3}{c|}{\textbf{Optimization}}\\
\cline{2-7}
\textbf{Language} & \textbf{data} & \textbf{workflow} & \textbf{pipeline} & \textbf{static} & \textbf{runtime} & \textbf{other}\\ \hline
Taverna/Scufl2 & \checkmark & \checkmark & \checkmark & -- & \checkmark & -- \\\hline
+/BPEL &foreach & \checkmark & -- &-- &\checkmark &-- \\\hline
Triana/- & \checkmark & \checkmark & \checkmark & -- & \checkmark &--\\\hline
MOTEUR/Scufl & \checkmark & \checkmark & \checkmark &-- &\checkmark &-- \\\hline
Kepler/MoML & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark &--\\\hline
GridNexus/GXPL & -- & \checkmark & -- &-- &\checkmark &-- \\\hline
Swift/SwiftScript & \checkmark & \checkmark & \checkmark & \checkmark &\checkmark &-- \\\hline
Askalon/AGWL &--& \checkmark & -- & \checkmark & \checkmark & -- \\\hline
P-GRADE/- &\checkmark &\checkmark &\checkmark &--&--&--\\\hline
Vizbuilder/bioflow &-- &\checkmark &-- &--&\checkmark &Query optimization\\\hline
Martlet/- & \checkmark & \checkmark & -- &-- &\checkmark&-- \\\hline
Knime/- & \checkmark &\checkmark &\checkmark & -- & \checkmark & -- \\\hline
StarFlow/- &\checkmark & \checkmark & \checkmark & \checkmark & \checkmark & -- \\\hline
Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark &AI-based techniques \\\hline
Galaxy/- &-- &-- &\checkmark &-- &-- &-- \\\hline
GWES/GWorkflowDL &-- &-- &\checkmark &-- &\checkmark &-- \\\hline
MOTEUR2/Gwendia~ & \checkmark &\checkmark & \checkmark &-- &\checkmark &-- \\\hline
\end{tabular}
%\caption{A summary of SWEs Scalability and optimization Mechanisms}
\label{tbl:scalability}
\end{center}
\end{table}

Shown in Table \ref{tbl:scalability} is a summary of the Scalability and optimization mechanisms of the studied SWEs. The scalability features considered are through three levels of parallelism in the form of data, workflow and pipeline. The optimization features considered are the presence of mechanisms by which static and runtime optimizations are achieved.
%
% == DATA MANAGEMENT ==
\section{Data Management and Description} \label{sec:data}
Many aspects of data handling present significant diversity among the workflow
managers. Data management approaches for scientific workflows can be classified
as resource and language level. Resource level strategies include data
management at the application level of abstraction where the data is required
to be distributed, available and securely remote-accessible. It also determines
how the data will be delivered to a resource that handles the processing and
returning the processed data or results to the next processors in the workflow.
Provenance, staging, and managing heterogeneous data structures (files, arrays
etc.) are functional requirements at this level.

Whereas the language level data management involves how the actual data items
are described and handled across a workflow execution. Decisions such as how
complex data structures are supported, transferred and delivered from workflow
input source or from one processor to another are determined at language level.
In most workflow systems, the transfer of data is achieved through `data-ports'
attached to the processors. These ports act as channels to collect and transfer
data across a workflow. The association of data description with the workflow
determines whether the data is loosely or strongly coupled with a workflow. A
loosely coupled workflow with data has advantage over the strongly coupled ones
in the sense that the workflow is independent of its data and can be invoked
with a different set of data.

A source of complexity for the workflow manager is the existence of different
formats of data for a single scientific experiment. These include the different
legacy formats of storage as well as the different ways the data is organized
including database, filesystems and remote links. Further, most remote data is
stored at locations secured behind different types of secure communication
protocols involving special access mechanisms.

In the rest of this section, we study the data handling mechanisms of SWEs.
Taverna's reference scheme makes sure the data produced and consumed is in the
form of references. This approach of data management is efficient because: 1)
It takes less time to transport references, and 2) Locality of references
ensures that the data is processed at a location that is local to multiple
processors operating on the same dataset.

In MOTEUR and Kepler workflow environments, data is described as an XML grammar
allowing for nested collections or multi-dimensional arrays of data. Kepler's
COMAD \cite{keplercomad} workflow enactment model is well suited to collection
oriented data like arrays and streams. Similar efforts exist elsewhere to model
the streams for workflow programming
\cite{blower-haines-etal:2005,herath-plale:2010}.

MOTEUR only uses references to data files while Pegasus groups small files into
a single compressed file for transfer thus saving transfer time. It optimizes
the workflow by annotating and reusing the data already available, if
appropriate. Pegasus is able to generate a workflow from a metadata description
of the desired data product with the aid of Artificial Intelligence planning
techniques. Grouping of tasks by Pegasus increases the data usage efficiency
since a group of tasks that require the same set of files for execution are
localized. Pegasus also manages the temporary storage space needed for data
processing at a particular processor location and commands the local Operating
System to allocate the space. A data provider abstraction allows data to be
referenced in the Trident SWE from any external entities in the form of typed
referenced data objects.

A majority of SWE support primitive, MIME based or XSD data types similar to
traditional programming datatypes. Specific support for complex and structured
data types are provided by the SWEs based upon their domain and application
requirements. For instance, Kepler supports a specialized byte data streams and
Bioflow supports a natively developed life-science database engine. Others such
as the SwiftScript support extensible structures such as filesystem mappers and
customized C language style structures.

Remote and secure data access is supported by SWEs by providing secured
protocols based accession to remote data repositories. Third party protocols
such as GridFTP, HTTP over TLS (HTTPS) and gLite LFC (LCG File Catalog) access
are commonly supported.

\begin{table}[ht!]
\begin{center}
\hspace*{-2.2cm}
% use packages: array
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{SWE /} & \multicolumn{3}{c|}{\textbf{Language Level}} & \multicolumn{2}{c|}{\textbf{Resource Level}}\\
\cline{2-6}
\multirow{2}{*}{\textbf{Language}} & \textbf{Data Coupling} & \textbf{Data Types} & \textbf{Data Structures} & \textbf{Type} & \textbf{Access}\\
& \textbf{with Workflow} & \textbf{Supported} & \textbf{Supported} & & \\
\hline
Taverna/Scufl2&Loose&MIME-types&multi-dim Arrays&Local+Remote&By Reference\\\hline
+/BPEL&Strong&XSD-Based&--&Grids, Clouds& Adhoc\\\hline
%
\multirow{2}{*}{Triana/-}&Strong&XML Schema-&Streams,&Local+&Through \\
&&Types&Java Objects&Remote&Adaptors\\\hline
%
\multirow{2}{*}{MOTEUR/Scufl}&Loose&MIME-&Lists&Local+&Services+\\
&&Types&&Remote&Middleware\\\hline
% 
\multirow{2}{*}{Kepler/}&Strong&Primitive&Collections,&Filesystem&Customized\\
MoML&&Types&User defined&&Actors\\\hline
% 
GridNexus/GXPL&Loose&XSD-Based&--&Grid Data&OGSA-DAI \\\hline
% 
\multirow{2}{*}{Swift/SwiftScript} &Strong&--&1-Dim Arrays,&Shared& POSIX-style\\
&&&Customised&FileSystem&\\\hline
% 
Askalon/AGWL&Loose&MIME-types&Arrays&Data Repository&SQL-based\\\hline
%
\multirow{2}{*}{P-GRADE/-}&Loose&--&--&Grid&Secure\\
&&&&File Catalogs&Protocols\\\hline
%
Vizbuilder/bioflow&Strong&SQL-types&--&Database&SQL,PL/SQL\\\hline
%
\multirow{2}{*}{Martlet/-}&Loose&URI&Lists&Distributed& By\\
&&&&Filesystems&Reference\\\hline
% 
\multirow{2}{*}{Knime/-}&Loose&String, Integer,&Filestreams,&FileSystems,&Datamining\\
&&Double&Arrays&Databases&Tools\\\hline
% 
StarFlow/-&Strong&Python-based&Python-based&--&--\\\hline
GWES/GWorkflowDL&Strong&java-based&java-based&--&--\\\hline
% 
\multirow{2}{*}{}Pegasus/&Loose&--&--&Distributed&Secure\\
DAX&&&&Filesystems&Protocols\\\hline
% 
Galaxy/-&Loose&--&File Streams&File Systems&--\\\hline
% 
\multirow{2}{*}{MOTEUR2/} &Loose&Integer,String,&multi-dim Arrays&Grid&By\\
Gwendia~ &&Double,URI&&Data&Reference\\\hline
\end{tabular}
%\caption{A summary of SWEs Data Management Mechanisms}
\label{tbl:datadesc}
\end{center}
\end{table}

Table \ref{tbl:datadesc} presents a summary of the data management mechanisms of the studied SWEs. Language and resource level data management mechanisms are highlighted. At the language level, data description as coupled with that of the workflow, data types and structures are considered for the SWE studied. At the resource level, the major resource types supported and their access mechanisms are highlighted. We distinguish the remote form of Grid resource types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those resource access. Other resource types such as data repositories or file catalogues are the resource types specific to a given SWE. For instance, Askalon uses a specifically configured data repository to access data and application binaries.
%Replication is one of the solution.
% == INTERFACE TO COMPUTING AND DATA INFRASTRUCTURES ==
\section{Interface to Distributed Computing Infrastructures (DCIs)} \label{sec:inter}
Workflow environments are external systems to large scale computing
infrastructures. They require an interface to connect to and use the
infrastructures. Provided the complex and distributed nature of grids, it is
not easy and practical to interact with such an infrastructure directly. Most
distributed systems have a software service layer through which it interacts
with the outside world. These interactions include accepting queries for
information and accepting requests and data for job executions. Such a layer is
typically called the \textit{middleware}. A workflow engine intending to
interact with a grid environment must provide interface to link to such
middleware layer.

Most grid computing systems today are batch oriented execution systems (with a
notable exception of GridRPC \cite{seymour-nakada-etal:2002} based systems). A
batch of ``jobs" are submitted to the system which in turn passes through some
execution queues before actually being scheduled for execution. Such systems
are prone to high execution latencies and unreliability issues. A workflow
manager must take these issues into account while interacting with such
systems.

At the resource management level, most grid computing environments today are
utilizing one or more components of the Globus Toolkit
\cite{foster-kesselman:1997} or the Condor distributed computing software
system~\cite{dagman}.

The Globus Toolkit is a software stack that provides useful functionalities for
grid computing operations. These functionalities are provided in the form of a
software stack that ranges from low-level data management tasks to the higher
level functionalities like authentication and metadata services. Any external
entity accessing the grid computing system must possess proper authentication
and sufficient authority to operate upon the infrastructure. The most popular
secure access system in place for today's grids is the public key
infrastructure (PKI) system through which a secure access to the grid is
guaranteed for a finite time period. A workflow manager must be capable of
handling the security protocols with the underlying execution infrastructure.
The duration of access to the system is also of critical importance as a
typical scientific workflow is bound to run for a considerably long time
period. The Condor software system provides scheduling mechanisms to access the
low-layer grid resources. A grid workflow enactor must provide interface to
access the services of these external scheduling and middleware systems.

\begin{table}
\begin{center}
%\hspace*{-1cm}
% use packages: array
\begin{tabular}{|l|l|l|}
\hline
\textbf{SWE/Language} & \textbf{DCI Interfacing Mechanisms} & \textbf{Currently Interfaced DCIs}\\
\hline
Taverna/Scufl2 & By Extension & None Inherently \\\hline
+/BPEL & Ad-hoc & None Inherently \\\hline
Triana/- & Provides Adapters & P2P Networks, NGS (ngs.org.uk)  \\\hline
MOTEUR/Scufl & Inherent & EGI (www.eu-egi.eu), g5k (www.grid5000.fr)\\\hline
Kepler/MoML & Using Directors & EarthGrid (http://seek.ecoinformatics.org) \\\hline
GridNexus/GXPL & Inherent & Generic Grids \\\hline
Swift/SwiftScript & Inherent & IBM/bluegene/P HPC Clusters \\\hline
Askalon/AGWL & Inherent & Austrian Grid (http://www.austriangrid.at/)\\\hline
P-GRADE/- & Inherent & NGS \\\hline
Vizbuilder/bioflow & DCI Independent & None Inherently \\\hline
Martlet/- & interfaces multiple middlewares & Generic Grids \\\hline
Knime/- & Extensions & SGE Clusters \\\hline
StarFlow/- &Interfacing External Schedulars & Generic Clouds \\\hline
GWES/GWorkflowDL&Globus WS-Gram& German D-Grid \\\hline
Pegasus/DAX & Using Condor & OSG (www.opensciencegrid.org) \\\hline
Galaxy/- & Extensions & None Inherently \\\hline
MOTEUR2/Gwendia~ & Inherent & EGI, g5k, HIPernet Cloud \\\hline
\end{tabular}
%\vspace*{1cm}
%\caption{A summary of SWEs DCI Interfacing Mechanisms}
\label{tbl:interfacedci}
\end{center}
\end{table}

Remotely located grid resources are accessed normally over the Internet links.
In such situations, long network latency or outright link failures are common.
Workflow managers are likely to face issues such as no response from the job
submitter or long delays. There must be some mechanism in place to handle such
scenarios. Depending upon the sophistication of the underlying infrastructure a
failover mechanism or logging information could be obtained for failures. Often
times workflow managers have their own failover mechanisms in place.

In the rest of this section, we study the approaches and techniques of SWEs to
interact with the Computing and Data Infrastructures. We also discuss briefly
the mechanisms for the mandatory security requirements.

MOTEUR provides seamless access to local and remote grid data resources by
closely coupling itself with the vBrowser~\cite{olabarriaga-deboer-etal:2006}
virtual resource browser environment. It has a built-in job management
mechanism that includes building job-specification, job submission data staging
and collection of results from the jobmanager. Additionally, MOTEUR2 is
interfaced with the HiPerNet cloud, a virtual platform manager that takes into
account, both computing and network resources
\cite{vicat-blancprimet-roca-etal:2009}.

In Taverna, a processor's invocation sequence can be customized
programmatically to enhance the processor's performance selectively. This
feature can be exploited by effectively interfacing a Taverna processor to Grid
middleware and create jobs and inputs/outputs definitions for that middleware.

In Taverna, a processor's invocation sequence can be customized
programmatically to enhance the processor's performance selectively. This
feature can be exploited by effectively interfacing a Taverna processor to Grid
middleware and create jobs and inputs/outputs definitions for that middleware.

Pegasus interfaces explicitly and transparently with the infrastructure in that
it maps the simplified workflow DAGs to the grid resources through the Condor
resource management service. Pegasus interacts with the available replica
management services in order to optimize the data access cost for the workflow
executions at runtime. It does cluster--group similar jobs together, in order
to ease the load on the scheduler and the dispatching of those jobs to the
worker nodes.

Swift/Karajan and Pegasus interface to a POSIX-style shared file system which
is one crucial aspect to their performance. %== may be shifted to data desc ==

Askalon does provide advanced resources mapping and reservations. These
features however are limited to the fact that not all underlying
infrastructures allow for a resource reservation scheme as offered by ASKALON.
GridNexus uses the OGSA-DAI (Open Grid Services Architecture Data Access and
Integration) services to access Globus grid data resources. It uses
web/grid-service clients in order to interact with the computations wrapped by
those services.

The data management services of Trident supports local storage as well as the
remote data storage services such as the Amazon S3
(\textit{http://aws.amazon.com/s3}) and SQL Server Data Services (SSDS)
(\textit{http://microsoft.com/sql/dataservices}) clouds.

Table \ref{tbl:interfacedci} presents a summary of the studied SWEs mechanisms for interfacing DCIs.
% == WORKFLOW LANGUAGE EXPRESSIVENESS ==
\section{Workflow Language Expressiveness} \label{sec:lang}
A workflow language is the medium of expression of the task a workflow is supposed to accomplish. A workflow language acts as an interface between the workflow enactor and an end-user. Apart from correctly expressing the flow of execution and data, a workflow language must be able to link with the application code in a transparent way.

Historically, low-level, general-purpose scripting languages have been used for
workflow descriptions. For example, batch files, UNIX shell scripts and
script-based languages like perl, python \cite{begeman-belikov-etal:2010} etc.
There also exist some domain specific scripting languages such as bioperl
(\textit{www.bioperl.org}). However, these languages are less suitable for
expression of high-level, sophisticated workflow constructs such as iterations,
while at the same time dealing with parallelism and array processing. They also
fall short of providing implicit parallelism and must be defined explicitly by
the user designing workflows. Description of remote data is another important
aspect that these languages are limited in expressing. Thus, in modern SWEs, it
is required to have a higher-level, specialized and dedicated workflow
language.

Applications and in-silico scientific experiments exhibit diverse patterns of
data and/or control flow. Apart from this, operations and combinations of data
are required to correctly and efficiently express a scientific workflow. This
has led to a requirement of highly expressive workflow languages. Most tasks
could be accomplished by simplified and minimal constructs in a language,
however, they lead to unintuitive and inefficient workflow design. A
sophisticated language not only increases the readability and maintainability
of a workflow, it also increases the chances of its efficient deployment and
execution in the target computational environment.

Description of data and dataflow in a workflow such that the application code
is invoked correctly and in accordance with the experiment has been a
challenging task for large experiments. This is partly because of the large
amount and different type of parallelization opportunities and partly because
of the way the components of an application deal with the data.

A workflow language must be able to express the invocation that is required for
tools to deploy the application components. For example, an experiment may be a
combination of remotely executable services and deployable legacy application
binaries. Along with the application binaries, a workflow language must be
capable of handling the different expressions for different kinds of data
transfer methods and protocols. A common way of invoking legacy applications is
through a traditional Command Line Interface. It is thus desirable to have a
way to wrap such command lines in order to provide a suitable interface to such
a legacy
application~\cite{rojasbalderrama-montagnat-etal:2010,maheshwari:2007}.

%
\begin{table}
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{SWE/Language} &\textbf{Control Structures}& \textbf{Array Programming} & \textbf{Iterators}\\
\hline
Taverna/Scufl2 &Conditionals and Loops& \checkmark & \checkmark \\\hline
+/BPEL &Conditionals and Loops& -- & -- \\\hline
Triana/- &Conditionals and Loops& \checkmark & -- \\\hline
MOTEUR/Scufl &Conditionals& -- & \checkmark \\\hline
Kepler/MoML &Conditionals and Loops & \checkmark & limited \\\hline
GridNexus/GXPL &Conditionals and Loops& -- & -- \\\hline
Swift/SwiftScript & Conditionals and Loops & 1-dim & limited \\\hline
Askalon/AGWL & Conditionals and Loops & -- & -- \\\hline
P-GRADE/- & Conditionals & -- & \checkmark \\\hline
Vizbuilder/bioflow & Conditionals and Loops & -- & -- \\\hline
Martlet/- & Conditionals and Loops & -- & -- \\\hline
Knime/- & Loops & -- & \checkmark \\\hline
StarFlow/- & Conditionals and Loops & -- & -- \\\hline
GWES/GWorkflowDL & Conditionals& -- & -- \\\hline
Pegasus/DAX & -- & \checkmark & -- \\\hline
Galaxy/- & -- & -- & -- \\\hline
MOTEUR2/Gwendia~ & Conditionals and Loops & \checkmark & \checkmark \\\hline
\end{tabular}
%\caption{A summary of SWEs Expressiveness Capabilities}
\label{tbl:expressive}
\end{center}
\end{table}

\paragraph{Dataflow and Functional Languages} Functional and Dataflow languages
and their derivatives have been a viable choice for workflow languages
paradigm. In the dataflow paradigm, a workflow processor will `fire' when the
data sufficient for its firing is available. Dataflow programming paradigm
particularly suites the workflow approach of programming. There are many
research works advocating a dataflow oriented approach to solve problems
involving a flow of data among computing components, predominantly in a
Directed Acyclic Graph form \cite{dataflow,lee-messerschmitt:1987}. Early
research on dataflow languages as visual language \cite{sutherland:1966}
reinforce the notion of the suitability of dataflow paradigm for workflows.

Functional programming complement the dataflow paradigm by providing
appropriate expression and semantics to a dataflow. Several features of
functional programming are shared by the dataflow programming approach. One of
the key hinderance of functional programming is the `side-effects' produced
when the values of variables are mutated more than once during a single
execution. Dataflow and functional programming approach converge when these
`side-effects' are eliminated and a functional program can express pure
dataflow under such conditions. 
% Modify and write ==> Functional languages, such as SCALA and Haskell, offer
% an implicit way of exposing parallelism, making it easier to identify data
% dependencies so that the runtime system can implicitly schedule independent
% work (work that respects the dependency graph).
In this section we study the features of contemporary workflow languages that
make them expressive of representing complex data and control flow situations
and enable them to map these expression onto the underlying computing
infrastructure.

Pegasus uses DAX which is an XML-based language to describe DAGs. Applications
using Pegasus use higher level languages such as Python or Java to describe
workflows and then map them onto DAX specification. Such transformations face
two issues: 1. DAGs are simple to transform but can get complicated if
hierarchical DAGs are used. 2. DAGs representations need sophisticated
translators. All features in low-level languages such as loops are not
supported by DAGs. This must be taken care of by the translators.

Kepler, GridNexus and Triana support loops, conditionals and sub-workflows as
advanced constructs of their workflow language. These constructs are manifested
in the GUI as configurable components that get their mapping in the underlying
language by a mapper software component.

ASKALON's AGWL (Abstract Grid Workflow Language) provides an abstract XML-based
representation of dataflows. It does not directly support control flow, however
control flow can be achieved indirectly using `pseudo datatokens'. It is
modularized and supports sub-workflows. AGWL provides conditionals, iterators,
and explicit parallelism. AGWL provides sophisticated workflow constructs
including sub-workflows, parallel and sequential execution constructs, loops,
conditionals and iteration strategies.

The GWorkflowDL employed by GWES is an XML-based representation of Workflows
which are based on the theory of high-level petri nets. This theoreticaly
well-found language provides basis for high expressiveness to GWES workflows.
It leads to the expressions of workflows at three different levels of
abstractions from user-intentions at high-level to the executable services
instances at the low-level.
%Swift is dynamic
%
%A paragraph saying the most generic language approaches of contemporary
%workflow environments
Both Askalon and Pegasus enable users to define abstract workflows to be
concretized at the later stage of workflow development. This feature
facilitates users to have the flexibility to generate concrete workflows based
upon the state and availability of resources at the time of enactment.

The graphical representation is the only workflow language in Knime. It
provides an interesting feature of metanodes that acts as a self-contained
subworkflow. A metanode is a processor that act as a container for other
processors. Thus a branch of a workflow in Knime can be configured into a
single processor in the form of metanode. A single such processor might be
enacted multiple times giving a form of loop to a Knime workflow.

Table \ref{tbl:expressive} shows a summary of language expressiveness capabilities of the SWEs studied. We consider the control structures: conditionals and loops, support for arrays and iterators \cite{montagnat-glatard-etal:2006} as the expressive features of SWEs.

\section{Usability} \label{sec:usable}
A workflow environment influences the end-user comfort and convenience towards the goal of efficiently organizing and deploying a scientific experiment involving heavy usage of computational resources. However, there is no single factor that determines the usability of a SWE. A combination of several qualitative and quantitative factors play a role in determining how usable a given SWE is.
\begin{table}[ht!]
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{SWE/Language} & \textbf{Workflow Composition} & \textbf{Reusable} & \textbf{Extendable}\\
\hline
Taverna/Scufl2 & GUI & \checkmark & \checkmark\\\hline
+/BPEL & GUI & \checkmark & \checkmark \\\hline
Triana/- & GUI & \checkmark & \checkmark \\\hline
MOTEUR/Scufl & GUI & \checkmark & -- \\\hline
Kepler/MoML & GUI & \checkmark & \checkmark \\\hline
GridNexus/GXPL & GUI/Text & -- & -- \\\hline
Swift/SwiftScript & Text & -- & -- \\\hline
Askalon/AGWL & GUI & \checkmark & \checkmark \\\hline
P-GRADE/- & Web Portal & \checkmark & -- \\\hline
Vizbuilder/bioflow & GUI/Text & \checkmark & -- \\\hline
Martlet/- & Text & -- & -- \\\hline
Knime/- & GUI & \checkmark & -- \\\hline
StarFlow/- & Text & \checkmark & -- \\\hline
GWES/GWorkflowDL & GUI & \checkmark & \checkmark \\\hline
Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
Galaxy/- & Web Portal & -- & -- \\\hline
\multirow{2}{*}{MOTEUR2/} & GUI/Text & \checkmark & --\\
Gwendia~/gscript & & & \\\hline
\end{tabular}
%\caption{A summary of SWEs Usability Analysis}
\label{tbl:usability}
\end{center}
\end{table}

In this section, the following aspects of grid workflow usability are mainly
considered: a) Ease and convenience for workflow composition. This largely
rests on the available composition interfaces such as a GUI. b) Workflow
reusability. c) Extensibility by being able to be linked to external
applications or the ease of customizing the existing codebase of the SWE. Since
legacy applications are not developed keeping workflow enactment in mind, it
largely depends upon the extendability of the workflow environment to adapt to
the application's requirements. We discuss these features in more details in
the rest of this section and summarize them in table \ref{tbl:usability} for
the SWEs studied.

A visual workflow development environment is most prominent among the workflow
community. The visual paradigm of workflow development is convenient as it
provides an intuitive ability to the user to design workflows by visualizing
the flow of data and the processors involved in the data processing at each
stage. It also facilitates users to visualize the opportunities to exploit
parallelism in the graph.

Another dimension of usability is from the point of view of the computational
infrastructure and its middleware. A SWE must optimally expose and make
available all the features of the underlying middleware components.

On the other side of the workflow development paradigm are the script-based
workflow development capability. A script-based workflow development has its
own advantages and can lead to the development of highly expressive, compact
and portable workflows rapidly in the hands of expert or well-experienced
end-users. An ideal situation is a workflow manager allowing both visual as
well as scripting development of scientific workflows.

There are two key aspects to highly usable SWE, expressibility and simplicity.
From the end-user point of view, these are often contradictory and a trade-off
is required to attain a balance between the degree to which each aspect is
contained and come in the way of user comfort. User comfort, time to design and
construct a workflow are of particular importance in the case of SWEs as the
users are not necessarily technical experts.

This section explores the above mentioned aspects and features seen in the
SWEs.

The above mentioned aspects also play a role in workflow reusability.
Scientific procedures are recurring in nature and offer opportunities to reuse
fully or partially , the existing workflows~\cite{xiaorong-madey:2007}. In
order to be able to reuse an existing workflow, it must be easily available and
clearly described as to what are its functions.

Furthermore, other operational factors contribute significantly to a SWE's
overall usability. An example is as reported by Ramakrishnan et. al. in
\cite{ramakrishnan-nurmi-etal:2009} wherein factors such as virtual
reservations, coordinated use of resources and fault tolerance combined
together considerably enhance workflow usability. Among these, the fault
tolerance mechanisms are the most challenging to implement and realize at the
workflow level. This is owing to the fact that workflows interacting with many
diverse entities have a large number of, and unforeseen fault conditions. A
small number of SWEs provide some kind of fault tolerance mechanisms. For
instance MOTEUR provides a `retry' mechanism on job fault while SwiftScript and
BPEL provide an exception handling mechanism in order to notify users of
occurrence of known fault conditions.

Some interesting efforts have been made in the past to increase usability of
grid environments for a particular domain users such as medical imaging
\cite{olabarriaga-deboer-etal:2006,olabarriaga-glatard-etal:2008}. In the case
of SWEs, specific usability issues have been addressed relatively late.
Distributed workflow repositories \cite{stoyanovich-taskar-etal:2010} have been
especially popular recently. These repositories play an important role in
making a large number of scientific workflows available and hence reusable. The
\textit{myexperiment.org} \cite{deroure-goble-etal:2008} is one such popular
repository of Taverna workflows available publicly. Other examples of such
repositories are the workflow libraries maintained as a part of the Kepler
project (library.kepler-project.org/kepler/) and another one with the Vistrails
project \cite{callahan-freire-etal:2006}. Kepler also integrates with DataStaR
\footnote{https://wiki.duraspace.org/display/FEDORACREATE/DataStaR+Use+Case} to
support data sharing among researchers during the research process, and to
promote publishing or archiving data and metadata to discipline-specific data
centers. The workflows hosted in these repositories are tagged and categorized
based upon their features and functionalities. The ``myexperiment" repository
is especially notable for its adoption of social web paradigm where users not
only find and reuse the workflows but can actively collaborate on scientific
experiments over the web.

% ---------------------------------------------------------------------------

\section{Conclusions} \label{sec:sum}
% \begin{sidewaysfigure}
% \begin{center}
% \includegraphics[width=\linewidth]{2/figures/pdf/Taxonomy}
% \caption{A Taxonomy of the Features of Scientific Workflow Environments}
% \label{fig:taxonomy}
% \end{center}
% \end{sidewaysfigure}

%\begin{sidewaysfigure}
\begin{figure}
\begin{center}
\begin{tikzpicture}[]
 \path[mindmap,concept color=darkgray,text=white,font=\sf]
 node[concept] {Scientific Workflows}
 [clockwise from=0]
 child[concept color=darkgray,font=\sf] {
 node[concept] {scalability}
 [clockwise from=90]
 child [concept color=gray,font=\sf] {
 node[concept] {\small{parallelism}}
 [clockwise from=120]
 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{data}}
 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{pipeline}}}
	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{workflow}}}
% 	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\small{activity}}}
 }
 child[concept color=gray,font=\sf] { node[concept] {\scriptsize{optimization}} }
 }
 child[concept color=darkgray,font=\sf] { node[concept] {\small{data management}}}
 child[concept color=darkgray,font=\sf] {
 node[concept] {\small{expressiveness}}
 [clockwise from=-45]
 child[concept color=gray,font=\sf] { node[concept] {iterators} }
 child[concept color=gray,font=\sf] { node[concept] {\small{array programming}} }
 child[concept color=gray,font=\sf] { node[concept] {\small{control structures}} }
 }
 child[concept color=darkgray,font=\sf] {
 node[concept] {usability}
 [clockwise from=-120]
 child[concept color=gray,font=\sf] { node[concept] {\scriptsize{composition}} }
 child[concept color=gray,font=\sf] { node[concept] {extendible} }
 child[concept color=gray,font=\sf] { node[concept] {reusable} }
 }
 child[concept color=darkgray,font=\sf] { node[concept] {interface to DCIs} };
\end{tikzpicture}
\caption{A Taxonomy of the Features of Scientific Workflow Environments}
\label{fig:taxonomy}
\end{center}
\end{figure}

Figure \ref{fig:taxonomy} shows a taxonomy of SWE features studied in this paper. We studied twenty SWEs from different scientific fields developed worldwide. Some trends as observed in the approaches of the SWEs development are noted below:
\begin{itemize}
 \item A recent surge in distributed computing approach of Cloud Computing has given rise to adoption of new enactment technologies. 
 \item SWEs tend to offer complementary properties which has led to several collaborations among the developing groups. Notable among these are the recent ones between Taverna and Galaxy \cite{mohamed-alaa-etal:2010} and Kepler and Pegasus \cite{mandal-deelman-etal:2007}.
 \item Collection oriented SWEs wherein arrays of data are treated as first-class entities has been seen a significant trend among the contemporary SWEs.
 \item A trend of increasing number of SWEs providing a script-based workflow composition interface solely or in combination with a visual interface in the form of GUI has been observed.
 \item Workflow systems originally developed for business domains have been introduced to scientific domains, such as BPEL and Trident.
\end{itemize}

In the present work we explore different dimensions of SWE paradigm. We study a
number of SWE implementations. While we see a number of implementations
dominant in areas suitable to their targeted applications, there is no single
SWE that could sufficiently fulfill the objectives of this work. There is a
lack of an SWE addressing the workflow expressibility and parallelism
requirements while at the same time targeting enactment within the Grid
Environment. Key features such as effective application data combinations or
array-based execution coupled with Grid centric execution are either missing or
partially available in the studied workflows.

However, so far, from our study, we find Taverna as being the closest SWE that
can be a suitable candidate for our enactments. With this motivation, we get
into studying the details of the workings of the Taverna Workflow Environment.

While many contemporary SWEs are studied in this survey, we are aware that this is not an exhaustive list. Some other SWEs have been missed or omitted, notable among them being: ICENI \cite{furmento-mayer-etal:2002}, Calcium \cite{caromel-henrio-etal:2008} and YML \cite{delannoy-emad-etal:2006}.

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{sanebib}
%\nocite{*}
\end{document}

As a new technology, Wireless Sensor Networks (WSNs) has a wide
range of applications [Culler 2001,Bahl 2002,Akyildiz 2001], including
environment monitoring, smart buildings, medical care, industrial and
military applications. Among them, a recent trend is to develop
commercial sensor networks that require pervasive sensing of both
environment and human beings, for example, assisted living
[Akyildiz 2002,Harvard 2001,CROSSBOW] and smart homes
[Harvard 2001,Adya 2001,CROSSBOW].
% quote
\begin{quote}
``For these applications, sensor devices are incorporated into human
cloths [Natarajan 2001,Zhou 2006,Bahl 2002,Adya 2001] for monitoring
health related information like EKG readings, fall detection, and voice recognition".
\end{quote}
While collecting all these multimedia information
[Akyildiz 2002] requires a high network throughput, off-the-shelf
sensor devices only provide very limited bandwidth in a single
channel: 19.2Kbps in MICA2 [Bahl 2002] and 250Kbps in MICAz.

In this article, we propose MMSN, abbreviation for Multifrequency
Media access control for wireless Sensor Networks. The main
contributions of this work can be summarized as follows.
% itemize
\begin{itemize}
\item To the best of our knowledge, the MMSN protocol is the first
multifrequency MAC protocol especially designed for WSNs, in which
each device is equipped with a single radio transceiver and
the MAC layer packet size is very small.
\item Instead of using pairwise RTS/CTS frequency negotiation
[Adya 2001,Culler 2001; Tzamaloukas 2001; Zhou 2006],
we propose lightweight frequency assignments, which are good choices
for many deployed comparatively static WSNs.
\item We develop new toggle transmission and snooping techniques to
enable a single radio transceiver in a sensor device to achieve
scalable performance, avoiding the nonscalable ``one
control channel + multiple data channels'' design [Natarajan 2001].
\end{itemize}

% Head 1
\section{MMSN Protocol}

% Head 2
\subsection{Frequency Assignment}

We propose a suboptimal distribution to be used by each node, which is
easy to compute and does not depend on the number of competing
nodes. A natural candidate is an increasing geometric sequence, in
which
% Numbered Equation
\begin{equation}
\label{eqn:01}
P(t)=\frac{b^{\frac{t+1}{T+1}}-b^{\frac{t}{T+1}}}{b-1},
\end{equation}
where $t=0,{\ldots}\,,T$, and $b$ is a number greater than $1$.

In our algorithm, we use the suboptimal approach for simplicity and
generality. We need to make the distribution of the selected back-off
time slice at each node conform to what is shown in Equation
(\ref{eqn:01}). It is implemented as follows: First, a random
variable $\alpha$ with a uniform distribution within the interval
$(0, 1)$ is generated on each node, then time slice $i$ is selected
according to the following equation:
% Unnumbered Equation
\[
i=\lfloor(T+1)\log_b[\alpha(b-1)+1]\rfloor.
\]
It can be easily proven that the distribution of $i$ conforms to Equation
(\ref{eqn:01}).

So protocols [Bahl 2002,Culler 2001,Zhou 2006,Adya 2001,Culler 2001;
Tzamaloukas-01; Akyildiz-01] that use RTS/CTS
controls\footnote{RTS/CTS controls are required to be implemented by
802.11-compliant devices. They can be used as an optional mechanism
to avoid Hidden Terminal Problems in the 802.11 standard and
protocols based on those similar to [Akyildiz 2001] and
[Adya 2001].} for frequency negotiation and reservation are not
suitable for WSN applications, even though they exhibit good
performance in general wireless ad hoc
networks.

% Head 3
\subsubsection{Exclusive Frequency Assignment}

In exclusive frequency assignment, nodes first exchange their IDs
among two communication hops so that each node knows its two-hop
neighbors' IDs. In the second broadcast, each node beacons all
neighbors' IDs it has collected during the first broadcast period.

% Head 4
\paragraph{Eavesdropping}

Even though the even selection scheme leads to even sharing of
available frequencies among any two-hop neighborhood, it involves a
number of two-hop broadcasts. To reduce the communication cost, we
propose a lightweight eavesdropping scheme.

\subsection{Basic Notations}

As Algorithm~\ref{alg:one} states, for each frequency
number, each node calculates a random number (${\textit{Rnd}}_{\alpha}$) for
itself and a random number (${\textit{Rnd}}_{\beta}$) for each of its two-hop
neighbors with the same pseudorandom number generator.
% Algorithm
\begin{algorithm}[t]
\SetAlgoNoLine
\KwIn{Node $\alpha$'s ID ($ID_{\alpha}$), and node $\alpha$'s
neighbors' IDs within two communication hops.}
\KwOut{The frequency number ($FreNum_{\alpha}$) node $\alpha$ gets assigned.}
$index$ = 0; $FreNum_{\alpha}$ = -1\;
\Repeat{$FreNum_{\alpha} > -1$}{
        $Rnd_{\alpha}$ = Random($ID_{\alpha}$, $index$)\;
        $Found$ = $TRUE$\;
        \For{each node $\beta$ in $\alpha$'s two communication hops
    }{
      $Rnd_{\beta}$ = Random($ID_{\beta}$, $index$)\;
      \If{($Rnd_{\alpha} < Rnd_{\beta}$) \text{or} ($Rnd_{\alpha}$ ==
          $Rnd_{\beta}$ \text{and} $ID_{\alpha} < ID_{\beta}$)\;
      }{
        $Found$ = $FALSE$; break\;
      }
        }
     \eIf{$Found$}{
           $FreNum_{\alpha}$ = $index$\;
         }{
           $index$ ++\;
     }
      }
\caption{Frequency Number Computation}
\label{alg:one}
\end{algorithm}

Bus masters are divided into two disjoint sets, $\mathcal{M}_{RT}$
and $\mathcal{M}_{NRT}$.
% description
\begin{description}
\item[RT Masters]
$\mathcal{M}_{RT}=\{ \vec{m}_{1},\dots,\vec{m}_{n}\}$ denotes the
$n$ RT masters issuing real-time constrained requests. To model the
current request issued by an $\vec{m}_{i}$ in $\mathcal{M}_{RT}$,
three parameters---the recurrence time $(r_i)$, the service cycle
$(c_i)$, and the relative deadline $(d_i)$---are used, with their
relationships.
\item[NRT Masters]
$\mathcal{M}_{NRT}=\{ \vec{m}_{n+1},\dots,\vec{m}_{n+m}\}$ is a set
of $m$ masters issuing nonreal-time constrained requests. In our
model, each $\vec{m}_{j}$ in $\mathcal{M}_{NRT}$ needs only one
parameter, the service cycle, to model the current request it
issues.
\end{description}

Here, a question may arise, since each node has a global ID. Why
don't we just map nodes' IDs within two hops into a group of
frequency numbers and assign those numbers to all nodes within two
hops?

\section{Simulator}
\label{sec:sim}

If the model checker requests successors of a state which are not
created yet, the state space uses the simulator to create the
successors on-the-fly. To create successor states the simulator
conducts the following steps.
% enumerate
\begin{enumerate}
\item Load state into microcontroller model.
\item Determine assignments needed for resolving nondeterminism.
\item For each assignment.
      \begin{enumerate}
      \item either call interrupt handler or simulate effect of next instruction, or
      \item evaluate truth values of atomic propositions.
      \end{enumerate}
\item Return resulting states.
\end{enumerate}
Figure~\ref{fig:one} shows a typical microcontroller C program that
controls an automotive power window lift. The program is one of the
programs used in the case study described in Section~\ref{sec:sim}.
At first sight, the programs looks like an ANSI~C program. It
contains function calls, assignments, if clauses, and while loops.
% Figure
\begin{figure}
\centerline{\includegraphics{acmsmall-mouse}}
\caption{Code before preprocessing.}
\label{fig:one}
\end{figure}

\subsection{Problem Formulation}

The objective of variable coalescence-based offset assignment is to find
both the coalescence scheme and the MWPC on the coalesced graph. We start
with a few definitions and lemmas for variable coalescence.

% Enunciations
\begin{definition}[Coalesced Node (C-Node)]A C-node is a set of
live ranges (webs) in the AG or IG that are coalesced. Nodes within the same
C-node cannot interfere with each other on the IG. Before any coalescing is
done, each live range is a C-node by itself.
\end{definition}

\begin{definition}[C-AG (Coalesced Access Graph)]The C-AG is the access
graph after node coalescence, which is composed of all C-nodes and C-edges.
\end{definition}

\begin{lemma}
The C-MWPC problem is NP-complete.
\end{lemma}
\begin{proof} C-MWPC can be easily reduced to the MWPC problem assuming a
coalescence graph without any edge or a fully connected interference graph.
Therefore, each C-node is an uncoalesced live range after value separation
and C-PC is equivalent to PC. A fully connected interference graph is made
possible when all live ranges interfere with each other. Thus, the C-MWPC
problem is NP-complete.
\end{proof}

\begin{lemma}[Lemma Subhead]The solution to the C-MWPC problem is no
worse than the solution to the MWPC.
\end{lemma}
\begin{proof}
Simply, any solution to the MWPC is also a solution to the
C-MWPC. But some solutions to C-MWPC may not apply to the MWPC (if any
coalescing were made).
\end{proof}

\section{Performance Evaluation}

During all the experiments, the Geographic Forwarding (GF)
[Akyildiz 2001] routing protocol is used. GF exploits geographic
information of nodes and conducts local data-forwarding to achieve
end-to-end routing. Our simulation is
configured according to the settings in
Table~\ref{tab:one}. Each run lasts for 2 minutes and
repeated 100 times. For each data value we present in the results,
we also give its 90\% confidence interval.
% Table
\begin{table}%
\tbl{Simulation Configuration\label{tab:one}}{%
\begin{tabular}{|l|l|}
\hline
TERRAIN{$^a$}   & (200m$\times$200m) Square\\\hline
Node Number     & 289\\\hline
Node Placement  & Uniform\\\hline
Application     & Many-to-Many/Gossip CBR Streams\\\hline
Payload Size    & 32 bytes\\\hline
Routing Layer   & GF\\\hline
MAC Layer       & CSMA/MMSN\\\hline
Radio Layer     & RADIO-ACCNOISE\\\hline
Radio Bandwidth & 250Kbps\\\hline
Radio Range     & 20m--45m\\\hline
\end{tabular}}
\begin{tabnote}%
\Note{Source:}{This is a table
sourcenote. This is a table sourcenote. This is a table
sourcenote.}
\vskip2pt
\Note{Note:}{This is a table footnote.}
\tabnoteentry{$^a$}{This is a table footnote. This is a
table footnote. This is a table footnote.}
\end{tabnote}%
\end{table}%

\section{Conclusions}

In this article, we develop the first multifrequency MAC protocol for
WSN applications in which each device adopts a
single radio transceiver. The different MAC design requirements for
WSNs and general wireless ad-hoc networks are
compared, and a complete WSN multifrequency MAC design (MMSN) is
put forth. During the MMSN design, we analyze and evaluate different
choices for frequency assignments and also discuss the nonuniform
back-off algorithms for the slotted media access design.

% Start of "Sample References" section

\section{Typical references in new ACM Reference Format}
A paginated journal article \cite{Abril07}, an enumerated
journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96},
a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec. document)
\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00}
followed by the same example, however we only output the series if the volume number is given
\cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol. no.),
a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book
in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
an article in a proceedings (of a conference, symposium, workshop for example)
(paginated proceedings article) \cite{Andler79}, a proceedings article
with all possible elements \cite{Smith10}, an example of an enumerated
proceedings article \cite{VanGundy07},
an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85},
a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01}, \cite{Ablamowicz07},
\cite{Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03}
and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001},
work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author
\cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might contain
'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}.
Boris / Barbara Beeton: multi-volume works as books
\cite{MR781536} and \cite{MR781537}.

% Appendix
\appendix
\section*{APPENDIX}
\setcounter{section}{1}
In this appendix, we measure
the channel switching time of Micaz [CROSSBOW] sensor devices.
In our experiments, one mote alternatingly switches between Channels
11 and 12. Every time after the node switches to a channel, it sends
out a packet immediately and then changes to a new channel as soon
as the transmission is finished. We measure the
number of packets the test mote can send in 10 seconds, denoted as
$N_{1}$. In contrast, we also measure the same value of the test
mote without switching channels, denoted as $N_{2}$. We calculate
the channel-switching time $s$ as
\begin{eqnarray}%
s=\frac{10}{N_{1}}-\frac{10}{N_{2}}. \nonumber
\end{eqnarray}%
By repeating the experiments 100 times, we get the average
channel-switching time of Micaz motes: 24.3$\mu$s.

\appendixhead{ZHOU}

% Acknowledgments
\begin{acks}
The authors would like to thank Dr. Maura Turolla of Telecom
Italia for providing specifications about the application scenario.
\end{acks}

% Bibliography
\bibliographystyle{ACM-Reference-Format-Journals}
\bibliography{acmsmall-sample-bibfile}
                             % Sample .bib file with references that match those in
                             % the 'Specifications Document (V1.5)' as well containing
                             % 'legacy' bibs and bibs with 'alternate codings'.
                             % Gerry Murray - March 2012

% History dates
\received{February 2007}{March 2009}{June 2009}

% Electronic Appendix
\elecappendix

\medskip

\section{This is an example of Appendix section head}

Channel-switching time is measured as the time length it takes for
motes to successfully switch from one channel to another. This
parameter impacts the maximum network throughput, because motes
cannot receive or send any packet during this period of time, and it
also affects the efficiency of toggle snooping in MMSN, where motes
need to sense through channels rapidly.

By repeating experiments 100 times, we get the average
channel-switching time of Micaz motes: 24.3 $\mu$s. We then conduct
the same experiments with different Micaz motes, as well as
experiments with the transmitter switching from Channel 11 to other
channels. In both scenarios, the channel-switching time does not have
obvious changes. (In our experiments, all values are in the range of
23.6 $\mu$s to 24.9 $\mu$s.)

\section{Appendix section head}

The primary consumer of energy in WSNs is idle listening. The key to
reduce idle listening is executing low duty-cycle on nodes. Two
primary approaches are considered in controlling duty-cycles in the
MAC layer.

% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2014-04-04 paper.tex
removed unnecessary stuff

@@
 \end{document}
-
-As a new technology, Wireless Sensor Networks (WSNs) has a wide
-range of applications [Culler 2001,Bahl 2002,Akyildiz 2001], including
-environment monitoring, smart buildings, medical care, industrial and
-military applications. Among them, a recent trend is to develop
-commercial sensor networks that require pervasive sensing of both
-environment and human beings, for example, assisted living
-[Akyildiz 2002,Harvard 2001,CROSSBOW] and smart homes
-[Harvard 2001,Adya 2001,CROSSBOW].
-% quote
-\begin{quote}
-``For these applications, sensor devices are incorporated into human
-cloths [Natarajan 2001,Zhou 2006,Bahl 2002,Adya 2001] for monitoring
-health related information like EKG readings, fall detection, and voice recognition".
-\end{quote}
-While collecting all these multimedia information
-[Akyildiz 2002] requires a high network throughput, off-the-shelf
-sensor devices only provide very limited bandwidth in a single
-channel: 19.2Kbps in MICA2 [Bahl 2002] and 250Kbps in MICAz.
-
-In this article, we propose MMSN, abbreviation for Multifrequency
-Media access control for wireless Sensor Networks. The main
-contributions of this work can be summarized as follows.
-% itemize
-\begin{itemize}
-\item To the best of our knowledge, the MMSN protocol is the first
-multifrequency MAC protocol especially designed for WSNs, in which
-each device is equipped with a single radio transceiver and
-the MAC layer packet size is very small.
-\item Instead of using pairwise RTS/CTS frequency negotiation
-[Adya 2001,Culler 2001; Tzamaloukas 2001; Zhou 2006],
-we propose lightweight frequency assignments, which are good choices
-for many deployed comparatively static WSNs.
-\item We develop new toggle transmission and snooping techniques to
-enable a single radio transceiver in a sensor device to achieve
-scalable performance, avoiding the nonscalable ``one
-control channel + multiple data channels'' design [Natarajan 2001].
-\end{itemize}
-
-% Head 1
-\section{MMSN Protocol}
-
-% Head 2
-\subsection{Frequency Assignment}
-
-We propose a suboptimal distribution to be used by each node, which is
-easy to compute and does not depend on the number of competing
-nodes. A natural candidate is an increasing geometric sequence, in
-which
-% Numbered Equation
-\begin{equation}
-\label{eqn:01}
-P(t)=\frac{b^{\frac{t+1}{T+1}}-b^{\frac{t}{T+1}}}{b-1},
-\end{equation}
-where $t=0,{\ldots}\,,T$, and $b$ is a number greater than $1$.
-
-In our algorithm, we use the suboptimal approach for simplicity and
-generality. We need to make the distribution of the selected back-off
-time slice at each node conform to what is shown in Equation
-(\ref{eqn:01}). It is implemented as follows: First, a random
-variable $\alpha$ with a uniform distribution within the interval
-$(0, 1)$ is generated on each node, then time slice $i$ is selected
-according to the following equation:
-% Unnumbered Equation
-\[
-i=\lfloor(T+1)\log_b[\alpha(b-1)+1]\rfloor.
-\]
-It can be easily proven that the distribution of $i$ conforms to Equation
-(\ref{eqn:01}).
-
-So protocols [Bahl 2002,Culler 2001,Zhou 2006,Adya 2001,Culler 2001;
-Tzamaloukas-01; Akyildiz-01] that use RTS/CTS
-controls\footnote{RTS/CTS controls are required to be implemented by
-802.11-compliant devices. They can be used as an optional mechanism
-to avoid Hidden Terminal Problems in the 802.11 standard and
-protocols based on those similar to [Akyildiz 2001] and
-[Adya 2001].} for frequency negotiation and reservation are not
-suitable for WSN applications, even though they exhibit good
-performance in general wireless ad hoc
-networks.
-
-% Head 3
-\subsubsection{Exclusive Frequency Assignment}
-
-In exclusive frequency assignment, nodes first exchange their IDs
-among two communication hops so that each node knows its two-hop
-neighbors' IDs. In the second broadcast, each node beacons all
-neighbors' IDs it has collected during the first broadcast period.
-
-% Head 4
-\paragraph{Eavesdropping}
-
-Even though the even selection scheme leads to even sharing of
-available frequencies among any two-hop neighborhood, it involves a
-number of two-hop broadcasts. To reduce the communication cost, we
-propose a lightweight eavesdropping scheme.
-
-\subsection{Basic Notations}
-
-As Algorithm~\ref{alg:one} states, for each frequency
-number, each node calculates a random number (${\textit{Rnd}}_{\alpha}$) for
-itself and a random number (${\textit{Rnd}}_{\beta}$) for each of its two-hop
-neighbors with the same pseudorandom number generator.
-% Algorithm
-\begin{algorithm}[t]
-\SetAlgoNoLine
-\KwIn{Node $\alpha$'s ID ($ID_{\alpha}$), and node $\alpha$'s
-neighbors' IDs within two communication hops.}
-\KwOut{The frequency number ($FreNum_{\alpha}$) node $\alpha$ gets assigned.}
-$index$ = 0; $FreNum_{\alpha}$ = -1\;
-\Repeat{$FreNum_{\alpha} > -1$}{
-        $Rnd_{\alpha}$ = Random($ID_{\alpha}$, $index$)\;
-        $Found$ = $TRUE$\;
-        \For{each node $\beta$ in $\alpha$'s two communication hops
-    }{
-      $Rnd_{\beta}$ = Random($ID_{\beta}$, $index$)\;
-      \If{($Rnd_{\alpha} < Rnd_{\beta}$) \text{or} ($Rnd_{\alpha}$ ==
-          $Rnd_{\beta}$ \text{and} $ID_{\alpha} < ID_{\beta}$)\;
-      }{
-        $Found$ = $FALSE$; break\;
-      }
-        }
-     \eIf{$Found$}{
-           $FreNum_{\alpha}$ = $index$\;
-         }{
-           $index$ ++\;
-     }
-      }
-\caption{Frequency Number Computation}
-\label{alg:one}
-\end{algorithm}
-
-Bus masters are divided into two disjoint sets, $\mathcal{M}_{RT}$
-and $\mathcal{M}_{NRT}$.
-% description
-\begin{description}
-\item[RT Masters]
-$\mathcal{M}_{RT}=\{ \vec{m}_{1},\dots,\vec{m}_{n}\}$ denotes the
-$n$ RT masters issuing real-time constrained requests. To model the
-current request issued by an $\vec{m}_{i}$ in $\mathcal{M}_{RT}$,
-three parameters---the recurrence time $(r_i)$, the service cycle
-$(c_i)$, and the relative deadline $(d_i)$---are used, with their
-relationships.
-\item[NRT Masters]
-$\mathcal{M}_{NRT}=\{ \vec{m}_{n+1},\dots,\vec{m}_{n+m}\}$ is a set
-of $m$ masters issuing nonreal-time constrained requests. In our
-model, each $\vec{m}_{j}$ in $\mathcal{M}_{NRT}$ needs only one
-parameter, the service cycle, to model the current request it
-issues.
-\end{description}
-
-Here, a question may arise, since each node has a global ID. Why
-don't we just map nodes' IDs within two hops into a group of
-frequency numbers and assign those numbers to all nodes within two
-hops?
-
-\section{Simulator}
-\label{sec:sim}
-
-If the model checker requests successors of a state which are not
-created yet, the state space uses the simulator to create the
-successors on-the-fly. To create successor states the simulator
-conducts the following steps.
-% enumerate
-\begin{enumerate}
-\item Load state into microcontroller model.
-\item Determine assignments needed for resolving nondeterminism.
-\item For each assignment.
-      \begin{enumerate}
-      \item either call interrupt handler or simulate effect of next instruction, or
-      \item evaluate truth values of atomic propositions.
-      \end{enumerate}
-\item Return resulting states.
-\end{enumerate}
-Figure~\ref{fig:one} shows a typical microcontroller C program that
-controls an automotive power window lift. The program is one of the
-programs used in the case study described in Section~\ref{sec:sim}.
-At first sight, the programs looks like an ANSI~C program. It
-contains function calls, assignments, if clauses, and while loops.
-% Figure
-\begin{figure}
-\centerline{\includegraphics{acmsmall-mouse}}
-\caption{Code before preprocessing.}
-\label{fig:one}
-\end{figure}
-
-\subsection{Problem Formulation}
-
-The objective of variable coalescence-based offset assignment is to find
-both the coalescence scheme and the MWPC on the coalesced graph. We start
-with a few definitions and lemmas for variable coalescence.
-
-% Enunciations
-\begin{definition}[Coalesced Node (C-Node)]A C-node is a set of
-live ranges (webs) in the AG or IG that are coalesced. Nodes within the same
-C-node cannot interfere with each other on the IG. Before any coalescing is
-done, each live range is a C-node by itself.
-\end{definition}
-
-\begin{definition}[C-AG (Coalesced Access Graph)]The C-AG is the access
-graph after node coalescence, which is composed of all C-nodes and C-edges.
-\end{definition}
-
-\begin{lemma}
-The C-MWPC problem is NP-complete.
-\end{lemma}
-\begin{proof} C-MWPC can be easily reduced to the MWPC problem assuming a
-coalescence graph without any edge or a fully connected interference graph.
-Therefore, each C-node is an uncoalesced live range after value separation
-and C-PC is equivalent to PC. A fully connected interference graph is made
-possible when all live ranges interfere with each other. Thus, the C-MWPC
-problem is NP-complete.
-\end{proof}
-
-\begin{lemma}[Lemma Subhead]The solution to the C-MWPC problem is no
-worse than the solution to the MWPC.
-\end{lemma}
-\begin{proof}
-Simply, any solution to the MWPC is also a solution to the
-C-MWPC. But some solutions to C-MWPC may not apply to the MWPC (if any
-coalescing were made).
-\end{proof}
-
-\section{Performance Evaluation}
-
-During all the experiments, the Geographic Forwarding (GF)
-[Akyildiz 2001] routing protocol is used. GF exploits geographic
-information of nodes and conducts local data-forwarding to achieve
-end-to-end routing. Our simulation is
-configured according to the settings in
-Table~\ref{tab:one}. Each run lasts for 2 minutes and
-repeated 100 times. For each data value we present in the results,
-we also give its 90\% confidence interval.
-% Table
-\begin{table}%
-\tbl{Simulation Configuration\label{tab:one}}{%
-\begin{tabular}{|l|l|}
-\hline
-TERRAIN{$^a$}   & (200m$\times$200m) Square\\\hline
-Node Number     & 289\\\hline
-Node Placement  & Uniform\\\hline
-Application     & Many-to-Many/Gossip CBR Streams\\\hline
-Payload Size    & 32 bytes\\\hline
-Routing Layer   & GF\\\hline
-MAC Layer       & CSMA/MMSN\\\hline
-Radio Layer     & RADIO-ACCNOISE\\\hline
-Radio Bandwidth & 250Kbps\\\hline
-Radio Range     & 20m--45m\\\hline
-\end{tabular}}
-\begin{tabnote}%
-\Note{Source:}{This is a table
-sourcenote. This is a table sourcenote. This is a table
-sourcenote.}
-\vskip2pt
-\Note{Note:}{This is a table footnote.}
-\tabnoteentry{$^a$}{This is a table footnote. This is a
-table footnote. This is a table footnote.}
-\end{tabnote}%
-\end{table}%
-
-\section{Conclusions}
-
-In this article, we develop the first multifrequency MAC protocol for
-WSN applications in which each device adopts a
-single radio transceiver. The different MAC design requirements for
-WSNs and general wireless ad-hoc networks are
-compared, and a complete WSN multifrequency MAC design (MMSN) is
-put forth. During the MMSN design, we analyze and evaluate different
-choices for frequency assignments and also discuss the nonuniform
-back-off algorithms for the slotted media access design.
-
-% Start of "Sample References" section
-
-\section{Typical references in new ACM Reference Format}
-A paginated journal article \cite{Abril07}, an enumerated
-journal article \cite{Cohen07}, a reference to an entire issue \cite{JCohen96},
-a monograph (whole book) \cite{Kosiur01}, a monograph/whole book in a series (see 2a in spec. document)
-\cite{Harel79}, a divisible-book such as an anthology or compilation \cite{Editor00}
-followed by the same example, however we only output the series if the volume number is given
-\cite{Editor00a} (so Editor00a's series should NOT be present since it has no vol. no.),
-a chapter in a divisible book \cite{Spector90}, a chapter in a divisible book
-in a series \cite{Douglass98}, a multi-volume work as book \cite{Knuth97},
-an article in a proceedings (of a conference, symposium, workshop for example)
-(paginated proceedings article) \cite{Andler79}, a proceedings article
-with all possible elements \cite{Smith10}, an example of an enumerated
-proceedings article \cite{VanGundy07},
-an informally published work \cite{Harel78}, a doctoral dissertation \cite{Clarkson85},
-a master's thesis: \cite{anisi03}, an online document / world wide web resource \cite{Thornburg01}, \cite{Ablamowicz07},
-\cite{Poker06}, a video game (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03}
-and \cite{Lee05} and (Case 3) a patent \cite{JoeScientist001},
-work accepted for publication \cite{rous08}, 'YYYYb'-test for prolific author
-\cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might contain
-'duplicate' DOI and URLs (some SIAM articles) \cite{Kirschmer:2010:AEI:1958016.1958018}.
-Boris / Barbara Beeton: multi-volume works as books
-\cite{MR781536} and \cite{MR781537}.
-
-% Appendix
-\appendix
-\section*{APPENDIX}
-\setcounter{section}{1}
-In this appendix, we measure
-the channel switching time of Micaz [CROSSBOW] sensor devices.
-In our experiments, one mote alternatingly switches between Channels
-11 and 12. Every time after the node switches to a channel, it sends
-out a packet immediately and then changes to a new channel as soon
-as the transmission is finished. We measure the
-number of packets the test mote can send in 10 seconds, denoted as
-$N_{1}$. In contrast, we also measure the same value of the test
-mote without switching channels, denoted as $N_{2}$. We calculate
-the channel-switching time $s$ as
-\begin{eqnarray}%
-s=\frac{10}{N_{1}}-\frac{10}{N_{2}}. \nonumber
-\end{eqnarray}%
-By repeating the experiments 100 times, we get the average
-channel-switching time of Micaz motes: 24.3$\mu$s.
-
-\appendixhead{ZHOU}
-
-% Acknowledgments
-\begin{acks}
-The authors would like to thank Dr. Maura Turolla of Telecom
-Italia for providing specifications about the application scenario.
-\end{acks}
-
-% Bibliography
-\bibliographystyle{ACM-Reference-Format-Journals}
-\bibliography{acmsmall-sample-bibfile}
-                             % Sample .bib file with references that match those in
-                             % the 'Specifications Document (V1.5)' as well containing
-                             % 'legacy' bibs and bibs with 'alternate codings'.
-                             % Gerry Murray - March 2012
-
-% History dates
-\received{February 2007}{March 2009}{June 2009}
-
-% Electronic Appendix
-\elecappendix
-
-\medskip
-
-\section{This is an example of Appendix section head}
-
-Channel-switching time is measured as the time length it takes for
-motes to successfully switch from one channel to another. This
-parameter impacts the maximum network throughput, because motes
-cannot receive or send any packet during this period of time, and it
-also affects the efficiency of toggle snooping in MMSN, where motes
-need to sense through channels rapidly.
-
-By repeating experiments 100 times, we get the average
-channel-switching time of Micaz motes: 24.3 $\mu$s. We then conduct
-the same experiments with different Micaz motes, as well as
-experiments with the transmitter switching from Channel 11 to other
-channels. In both scenarios, the channel-switching time does not have
-obvious changes. (In our experiments, all values are in the range of
-23.6 $\mu$s to 24.9 $\mu$s.)
-
-\section{Appendix section head}
-
-The primary consumer of energy in WSNs is idle listening. The key to
-reduce idle listening is executing low duty-cycle on nodes. Two
-primary approaches are considered in controlling duty-cycles in the
-MAC layer.
-
-% End of v2-acmsmall-sample.tex (March 2012) - Gerry Murray, ACM

### 2014-04-05 paper.tex
removed unnecessary stuff
some new bib entries

@@
 %
-\item \textbf{Scalability and Optimization.} SWEs are more likely to involve extensive usage of distributed resources than conventional programs.
-Owing to the large-scale experiments, for which scientific users turn to automation, their demands for computational power often exceeds that of a single personal computer.
-Such applications are often composed of multiple, distinct computational stages making them well-suited to the workflow paradigm.
-These properties also make them ideal candidates for a large-scale distributed computing environment.
-The case for suitability of SWEs on large-scale distributed infrastructures has been well-argued upon in~\cite{jha-katz-etal:2009}.
-In such a scenario, taking explicit steps to ensure scalable and optimized performance is vital.
-%
-\item \textbf{Data Management.} The scale of data transfer is much higher in SWEs from one execution site to others as well as among storage resources while it is lower (ranging from variables to small data structures) in programs.
-%
-\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).} Building SWE's interfaces to distributed systems is a complex task.
-The distributed nature of resource utilization and asynchronous execution of workflow requires a set of interdependent parameters to be configured correctly.
-Further complexity is added because of complex dataflow requirements, interactions with diverse tools and linking with legacy application codes.
-%
-\item \textbf{Expressiveness.} Programming languages are inherently control flow oriented where data flow is a secondary aspect.
-In fact, the programmer must explicitly make sure that correct the inputs are available to instructions and are in the right format.
-A program proceeds as the execution of its statements from beginning to end and completes its execution in a predefined sequence.
-However, in SWEs, the flow is often determined by the data availability for execution.
-This model of computation is known as dataflow computing.
-A workflow thus does not follow a fixed order of execution and different execution sequences are possible which are dominated by dataflow.
-The unit of work in SWEs is at a higher level of granularity.
-A unit of execution is coarse-grained, and often termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE community.
-This notion of high-level granularity drives the depth to which a workflow can be expressed in a given language form.
-In the rest of this paper, this unit of work will be referred to as \textit{processor}.
-On the other hand, a unit of execution in most programming languages is a much simpler and finer executable statement that accomplishes a small computational task.
-This presents a whole new set of requirements for expressing the processors and how they should interact and operate upon data.
-%
-\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain experts to utilize the computational power.
-This makes it essential for them to have a friendly, and intuitive user-interface for novice and non-expert users.
-For this reason, SWE user interfaces are dominated by the visual paradigm of programming.
-On the other hand, traditional programs are most often written by expert programmers.
-Thus, they have a text-based composition environment albeit with visually assisting programming environments such as IDEs.
-However, they are seldom to the extent that a non-expert user can graphically compose programs.
+
+\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
+extensive usage of distributed resources than conventional programs. Owing
+to the large-scale experiments, for which scientific users turn to
+automation, their demands for computational power often exceeds that of a
+single personal computer. Such applications are often composed of multiple,
+distinct computational stages making them well-suited to the workflow
+paradigm. These properties also make them ideal candidates for a
+large-scale distributed computing environment. The case for suitability of
+SWEs on large-scale distributed infrastructures has been well-argued upon
+in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
+ensure scalable and optimized performance is vital.
+%
+
+\item \textbf{Data Management.} The scale of data transfer is much higher in
+SWEs from one execution site to others as well as among storage resources
+while it is lower (ranging from variables to small data structures) in
+programs.
+%
+
+\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
+Building SWE's interfaces to distributed systems is a complex task. The
+distributed nature of resource utilization and asynchronous execution of
+workflow requires a set of interdependent parameters to be configured
+correctly. Further complexity is added because of complex dataflow
+requirements, interactions with diverse tools and linking with legacy
+application codes.
+
+%
+\item \textbf{Expressiveness.} Programming languages are inherently control
+flow oriented where data flow is a secondary aspect. In fact, the
+programmer must explicitly make sure that correct the inputs are available
+to instructions and are in the right format. A program proceeds as the
+execution of its statements from beginning to end and completes its
+execution in a predefined sequence. However, in SWEs, the flow is often
+determined by the data availability for execution. This model of
+computation is known as dataflow computing. A workflow thus does not follow
+a fixed order of execution and different execution sequences are possible
+which are dominated by dataflow. The unit of work in SWEs is at a higher
+level of granularity. A unit of execution is coarse-grained, and often
+termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
+community. This notion of high-level granularity drives the depth to which
+a workflow can be expressed in a given language form. In the rest of this
+paper, this unit of work will be referred to as \textit{processor}. On the
+other hand, a unit of execution in most programming languages is a much
+simpler and finer executable statement that accomplishes a small
+computational task. This presents a whole new set of requirements for
+expressing the processors and how they should interact and operate upon
+data.
+
+%
+\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
+experts to utilize the computational power. This makes it essential for
+them to have a friendly, and intuitive user-interface for novice and
+non-expert users. For this reason, SWE user interfaces are dominated by the
+visual paradigm of programming. On the other hand, traditional programs are
+most often written by expert programmers. Thus, they have a text-based
+composition environment albeit with visually assisting programming
+environments such as IDEs. However, they are seldom to the extent that a
+non-expert user can graphically compose programs.
 
@@
 \section{Architecture of a Generic SWE}
-Figure \ref{fig:wfms} shows a high-level architecture of a SWE.
-It shows key components and their interaction in a SWE.
-These components are the focus of interest for the present work.
-A \textit{user interface} interacts with the \textit{workflow engine} using commands and messages.
-The \textit{workflow language} becomes an intermediate representation for communication between user interface and the \textit{workflow engine}.
-The workflow engine acts as a compiler/interpreter and scheduler for the workflow tasks and interfaces with the underlying \textit{distributed resources}.
-It also monitors the workflow progress and collects the intermediate and final results.
+Figure \ref{fig:wfms} shows a high-level architecture of a SWE. It shows key
+components and their interaction in a SWE. These components are the focus of
+interest for the present work. A \textit{user interface} interacts with the
+\textit{workflow engine} using commands and messages. The \textit{workflow
+language} becomes an intermediate representation for communication between user
+interface and the \textit{workflow engine}. The workflow engine acts as a
+compiler/interpreter and scheduler for the workflow tasks and interfaces with
+the underlying \textit{distributed resources}. It also monitors the workflow
+progress and collects the intermediate and final results.
+
 %
@@
 %
-SWEs are used as tools to benefit from the underlying large-scale computing infrastructure such as clusters, supercomputers, clouds, grids.
-Such environments are in use today for production level work in different scientific domains worldwide.
-They do the basic task of linking inter-related processors of a multi-staged scientific computation.
-However, they differ in the features they provide based upon the scientific domain they offer service for, and the computational model they follow.
-Often, one of the main reasons for such difference is the specialized components and requirements for a particular domain.
-Owing to this fact, a given workflow system dominates a particular area of science (eg.
-astronomy, biology, etc.).
-Nonetheless, there exist a plethora of SWEs suitable for general purpose workflow style computations.
+SWEs are used as tools to benefit from the underlying large-scale computing
+infrastructure such as clusters, supercomputers, clouds, grids. Such
+environments are in use today for production level work in different scientific
+domains worldwide. They do the basic task of linking inter-related processors
+of a multi-staged scientific computation. However, they differ in the features
+they provide based upon the scientific domain they offer service for, and the
+computational model they follow. Often, one of the main reasons for such
+difference is the specialized components and requirements for a particular
+domain. Owing to this fact, a given workflow system dominates a particular area
+of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora
+of SWEs suitable for general purpose workflow style computations.
 % Ketanmark
@@
 
-MOTEUR has sophisticated data management mechanisms within the workflow definition.
-It provides a diverse set of iterations strategies well adapted for different combinations of input datasets at the processor.
-An iteration strategy allows different combinations of the data arriving at a processor along its multiple inputs.
-Running legacy application code on grids has been eased by MOTEUR by providing a generic grid-service wrapper for these applications \cite{glatard-montagnat-etal:2008a}.
-Grid-services thus created can be readily deployed on the grids by the MOTEUR workflow enactor.
-MOTEUR workflows are written in the scufl language and are compatible with those written for the Taverna SWE.
-It also supports an extension of scufl language called the Gwendia language, developed at the same laboratory.
-%
-\paragraph{Makeflow} Business Process Execution Language (BPEL) \cite{oasis:2007} is a workflow language pioneered by IBM Corporation and used predominantly to define business workflows.
-It is now a broad standard for business workflow applications.
-BPEL is an open language and commercial enactors are available that provide enactment functionality for the BPEL workflows.
-There has been attempts to use BPEL or its extended version in the scientific applications \cite{slominski:2007,bpel}.
-However, the language does not suit well to the data-intensive, grid workflows and has a limited adoption in the scientific community \cite{tan-missier-etal:2010,barga-gannon:2007,scherp-hoing-etal:2009,emmerich-butchart-etal:2005}.
-Apart from that, a lack of an open-source workflow engine with graphical interface also contributes to its limited acceptance in academic research.
-However, BPEL has been famously unsuitable for data intensive scientific applications  BPEL lacks different user-defined strategies to effect multiple invocations of each processor according to the application semantics.
-BPEL also lacks data streams and implicit parallelism forcing a user to explicitly define parallelism using one of its parallel processing constructs such as \texttt{foreach}.
-%
-\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based SWE developed at the Cardiff University.
-It consists of a workflow composition GUI, a framework to integrate workflow with the different execution environments and middlewares and a set of predefined services written as java components.
-A Triana workflow is expressed as an XML grammar similar to a WSDL specification for webservices \cite{taylor-wand-etal:2005}.
-A highlight of Triana is that it provides ready-to-use interface for a wide variety of middleware and distributed infrastructures including P2P, grids and web-services frameworks.
-A Triana workflow provides dataflow as well as control flow support.
-Triana allows conditional branching and looping in its workflows.
-%
-\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at the University of California.
-Kepler workflows follow an actor-director metaphor where an actor represents a unit of execution and the director represents the paradigm for executing the workflows consisting of the actors.
-Depending upon the types of directors, a Kepler workflow may exhibit a different type of execution behavior.
-A workflow designer can chose from among directors based upon the purpose of a workflow.
-Few of these include Process Network (PN) for a parallel execution at the workflow level, Synchronous Data Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and Design (COMAD) \cite{keplercomad} for a collection oriented workflow.
-Kepler uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its workflow representation.
-MoML is an extended XML syntax providing a rich vocabulary to compose Kepler workflows.
-A workflow processor is called an `actor' in Kepler.
-%
-\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the University of North Carolina, Wilmington, facilitates a graphical composition and execution of DAG-based workflows.
-Like Kepler, GridNexus is also based on Ptolemy II framework.
-GridNexus workflows are expressed in an XML-based scripting language called JXPL \cite{hunt-ferner-etal:2005}.
-JXPL is a customizable and extensible scripting language allowing users to write their own workflow processor types.
-GridNexus workflows are grid oriented and can be deployed on grid resources accessed through a persistent WS-RF (Web Services Resource Framework) interface.
-%
-\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed computing framework that can map abstract workflows as DAGs to the concrete underlying resources using intelligent and dynamic scheduling policies.
-A Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in XML).
-DAX describes the logical workflow components, their inputs and outputs and their links and dependencies in a Pegasus workflow.
-These kind of workflows has been shown to enact optimally using Pegasus' smart memory and disk management techniques.
-Pegasus manager buffers and maintains a fix-sized window of executable tasks before accepting new tasks.
-%
-
-\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific workflow environment developed at the University of Chicago.
-A Swift workflow specified in the SwiftScript scripting language is translated to an intermediate XML-based representation and enacted by the Karajan workflow enactment engine.
-The SwiftScript language is proposed as a workflow language founded on a scripting approach programmers are familiar with.
-To make the language adapted to grids, the inventors of SwiftScript generalized the use of futures \cite{Halstead85multilisp:a} for every variables defined in the script.
-Futures are non-blocking assignment variables, using a proxy to ensure immediate execution of assignment and performing lazy blocking on variable value accesses only.
-The systematic use of futures for every variables in the language makes it completely data-driven: the execution progresses asynchronously as long as a data access is not blocking.
-The availability of data enables blocked thread to restart execution.
-The beauty of futures is to make this process completely hidden from the programmer who can use a scripting language he is familiar with without ever manipulating explicit parallel programming constructs.
-Swift facilitates a highly asynchronous workflow execution.
-Swift supports advanced workflow constructs such as parallel loops and conditionals.
-Furthermore Swift supports processing of explicitly defined arrays of data.
-Swift workflows take advantage of POSIX-style shared file system to reuse the data produced by other processors.
+MOTEUR has sophisticated data management mechanisms within the workflow
+definition. It provides a diverse set of iterations strategies well adapted for
+different combinations of input datasets at the processor. An iteration
+strategy allows different combinations of the data arriving at a processor
+along its multiple inputs. Running legacy application code on grids has been
+eased by MOTEUR by providing a generic grid-service wrapper for these
+applications \cite{glatard-montagnat-etal:2008a}. Grid-services thus created
+can be readily deployed on the grids by the MOTEUR workflow enactor. MOTEUR
+workflows are written in the scufl language and are compatible with those
+written for the Taverna SWE. It also supports an extension of scufl language
+called the Gwendia language, developed at the same laboratory. 
+%
+\paragraph{Makeflow} A make-like workflow platform~\cite{makeflow}.
+
+%
+\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based
+SWE developed at the Cardiff University. It consists of a workflow composition
+GUI, a framework to integrate workflow with the different execution
+environments and middlewares and a set of predefined services written as java
+components. A Triana workflow is expressed as an XML grammar similar to a WSDL
+specification for webservices \cite{taylor-wand-etal:2005}. A highlight of
+Triana is that it provides ready-to-use interface for a wide variety of
+middleware and distributed infrastructures including P2P, grids and
+web-services frameworks. A Triana workflow provides dataflow as well as control
+flow support. Triana allows conditional branching and looping in its workflows.
+
+%
+\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE
+based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at
+the University of California. Kepler workflows follow an actor-director
+metaphor where an actor represents a unit of execution and the director
+represents the paradigm for executing the workflows consisting of the actors.
+Depending upon the types of directors, a Kepler workflow may exhibit a
+different type of execution behavior. A workflow designer can chose from among
+directors based upon the purpose of a workflow. Few of these include Process
+Network (PN) for a parallel execution at the workflow level, Synchronous Data
+Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
+Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
+uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its
+workflow representation. MoML is an extended XML syntax providing a rich
+vocabulary to compose Kepler workflows. A workflow processor is called an
+`actor' in Kepler.
+
+%
+\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the
+University of North Carolina, Wilmington, facilitates a graphical composition
+and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
+Ptolemy II framework. GridNexus workflows are expressed in an XML-based
+scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
+customizable and extensible scripting language allowing users to write their
+own workflow processor types. GridNexus workflows are grid oriented and can be
+deployed on grid resources accessed through a persistent WS-RF (Web Services
+Resource Framework) interface.
+
+%
+\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed
+computing framework that can map abstract workflows as DAGs to the concrete
+underlying resources using intelligent and dynamic scheduling policies. A
+Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
+XML). DAX describes the logical workflow components, their inputs and outputs
+and their links and dependencies in a Pegasus workflow. These kind of workflows
+has been shown to enact optimally using Pegasus' smart memory and disk
+management techniques. Pegasus manager buffers and maintains a fix-sized window
+of executable tasks before accepting new tasks.
+
+%
+
+\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific
+workflow environment developed at the University of Chicago. A Swift workflow
+specified in the SwiftScript scripting language is translated to an
+intermediate XML-based representation and enacted by the Karajan workflow
+enactment engine. The SwiftScript language is proposed as a workflow language
+founded on a scripting approach programmers are familiar with. To make the
+language adapted to grids, the inventors of SwiftScript generalized the use of
+futures \cite{Halstead85multilisp:a} for every variables defined in the script.
+Futures are non-blocking assignment variables, using a proxy to ensure
+immediate execution of assignment and performing lazy blocking on variable
+value accesses only. The systematic use of futures for every variables in the
+language makes it completely data-driven: the execution progresses
+asynchronously as long as a data access is not blocking. The availability of
+data enables blocked thread to restart execution. The beauty of futures is to
+make this process completely hidden from the programmer who can use a scripting
+language he is familiar with without ever manipulating explicit parallel
+programming constructs. Swift facilitates a highly asynchronous workflow
+execution. Swift supports advanced workflow constructs such as parallel loops
+and conditionals. Furthermore Swift supports processing of explicitly defined
+arrays of data. Swift workflows take advantage of POSIX-style shared file
+system to reuse the data produced by other processors.
+
 % Put this somewhere else, either in the analysis or may be in the conclusions and summary.
@@
 Conditional and Loop execution is supported using the if and for/while statements in both the languages.
-%
-\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE) \cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
-ASKALON is described as a Scientific PSE capable of composing, hosting, executing, sharing and reusing scientific workflows written in the Abstract Grid Workflow Language (AGWL).
-AGWL is an XML-based workflow language that is capable of expressing scientific application components chains at a high-level of abstraction.
-A concrete workflow consists of the abstract workflow combined with the resources and target execution environment binding the workflow.
-The graphical interface of AGWL is based upon the UML (Unified Modeling Language) diagrams.
-%
-\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service (GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First institute.
-It implements a dynamic workflow composition and enactment model based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}.
-GWES enables key workflow development and deployment lifecycle phases including visual composition, data-staging, Grid middleware interfacing and execution monitoring.
-Currently, GWES is intensively being used in the field of Medical Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany.
+
+%
+\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
+\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
+ASKALON is described as a Scientific PSE capable of composing, hosting,
+executing, sharing and reusing scientific workflows written in the Abstract
+Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
+capable of expressing scientific application components chains at a high-level
+of abstraction. A concrete workflow consists of the abstract workflow combined
+with the resources and target execution environment binding the workflow. The
+graphical interface of AGWL is based upon the UML (Unified Modeling Language)
+diagrams.
+
+%
+\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service
+(GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First
+institute. It implements a dynamic workflow composition and enactment model
+based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}.
+GWES enables key workflow development and deployment lifecycle phases including
+visual composition, data-staging, Grid middleware interfacing and execution
+monitoring. Currently, GWES is intensively being used in the field of Medical
+Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany.
 It has been successfully interfaced with the German D-Grid infrastructure.
@@
 
-\paragraph{The Galaxy Portal/JSON} The P-GRADE Portal \cite{kacsuk-sipos:2005} developed at the MTA SZTAKI Computation and Automation Institute, Budapest, provides an access to grid computational and storage resources from a web browser through a specialized workflow oriented grid portal.
-A workflow development environment interfaces the users through this web based portal.
-The portal interfaces the multiple grid computing middleware and storage resources as a backend component of the environment.
-A highlight of P-GRADE portal is that it provides access to multiple grids and scientific communities in a single environment leading to a highly collaborative inter-community, inter-grid research environment.
-An older version of P-GRADE used the condor-Dagman as workflow engine while the recent version uses its own in-house developed workflow engine called Xen.
+\paragraph{The P-GRADE Portal/--} The P-GRADE Portal \cite{kacsuk-sipos:2005}
+developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
+provides an access to grid computational and storage resources from a web
+browser through a specialized workflow oriented grid portal. A workflow
+development environment interfaces the users through this web based portal. The
+portal interfaces the multiple grid computing middleware and storage resources
+as a backend component of the environment. A highlight of P-GRADE portal is
+that it provides access to multiple grids and scientific communities in a
+single environment leading to a highly collaborative inter-community,
+inter-grid research environment. An older version of P-GRADE used the
+condor-Dagman as workflow engine while the recent version uses its own in-house
+developed workflow engine called Xen.
+
 %
 \paragraph{DaGue/--} Bioflow \cite{bioflow} is a query based workflow language.
-It is motivated by the fact that bioinformatics applications are query based applications that benefit from the complex database queries or saved database `views'.
-A typical Bioflow workflow pipelines a sequence of filter queries in order to get the desired results.
-Bioflow is mainly utilized in the area of bioinformatics and querying genome databases.
-Additionally, a Bioflow workflow provides mechanisms for data manipulation and updation in the existing databases.
-It also includes advanced features such as parallel processing of a set of mutually independent database queries.
-The execution model is based on the popular database query language SQL (Structured Query Language).
-Processes can be defined in Bioflow which are processor counterparts in workflow paradigm and similar to stored-procedures in SQL jargon.
-
-Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical representation of bioflow workflows.
-A Vizbuilder workflow links to an underlying bioflow engine and provides an interface to submit the bioflow workflows for execution.
-%
-\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative workflow language that is inspired by the functional programming paradigm.
-A Martlet workflow conveniently enables distributed parameter sweep applications wherein the language intuitively defines the computation and a matching with the distributed resources can be carried out at workflow execution time.
-List processing applications where a sequence of application components work over an array of data in a Multi-data multi-instruction style setup is well-suited to the Martlet's model of computation.
-%
-\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining system for data intensive applications developed at the University of Konstanz.
-It is a java-based platform with a Graphical Interface that is based on eclipse (\textit{http://www.eclipse.org}) plug-in framework.
-A Knime workflow consists of processors belonging to categories such as readers, manipulators, learners, predictors, etc.
-Each processor has a configuration dialog associated with it that allows user to set its parameters.
-Processors are connected through data and/or control connections.
-A highlight of knime is support for shared- and distributed-memory workflow enactment and support for parallel workflow loops and subworkflows.
-The loops in Knime manifest themselves in the form of subworkflows (called metanodes in Knime) by a repeated enactment of the same subworkflow.
-%
-\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based script oriented workflow language used to encode parallel pipelines.
-StarFlow handles data as well as control flow using a combination of static, dynamic and user annotations analysis.
-A parallelization engine uses data and control dependencies in a workflow in order to build a parallel profile of the workflow.
-StarFlow has been shown to be successfully interface with the SunGrid \cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
-The key feature of StarFlow is its capability to execute workflows to maximum possible parallelism.
-While StarFlow provides a script-based compact and semiautomatic workflow composition environment, currently it lacks a sophisticated GUI for workflow composition.
-
-\paragraph{Galaxy/--} Galaxy \cite{blankenberg-vonkuster-etal:2010} is a bioinformatics domain SWE based framework developed at the Penn State University, Pennsilvania.
-It provides a webportal based environment to build complex pipelines from an available repository of tools (\textit{http://g2.bx.psu.edu}).
-Users can extend the tool repository provided by Galaxy using their own Galaxy server or a web-server exposing the webservice descriptions through public WSDLs.
-One of the major use of Galaxy is to achieve a mix and match of data analysis tools in order to analyze a vast repository of genome based database.
+It is motivated by the fact that bioinformatics applications are query based
+applications that benefit from the complex database queries or saved database
+`views'. A typical Bioflow workflow pipelines a sequence of filter queries in
+order to get the desired results. Bioflow is mainly utilized in the area of
+bioinformatics and querying genome databases. Additionally, a Bioflow workflow
+provides mechanisms for data manipulation and updation in the existing
+databases. It also includes advanced features such as parallel processing of a
+set of mutually independent database queries. The execution model is based on
+the popular database query language SQL (Structured Query Language). Processes
+can be defined in Bioflow which are processor counterparts in workflow paradigm
+and similar to stored-procedures in SQL jargon.
+
+Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical
+representation of bioflow workflows. A Vizbuilder workflow links to an
+underlying bioflow engine and provides an interface to submit the bioflow
+workflows for execution.
+
+%
+\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative
+workflow language that is inspired by the functional programming paradigm. A
+Martlet workflow conveniently enables distributed parameter sweep applications
+wherein the language intuitively defines the computation and a matching with
+the distributed resources can be carried out at workflow execution time. List
+processing applications where a sequence of application components work over an
+array of data in a Multi-data multi-instruction style setup is well-suited to
+the Martlet's model of computation.
+
+%
+\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining
+system for data intensive applications developed at the University of Konstanz.
+It is a java-based platform with a Graphical Interface that is based on eclipse
+(\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists
+of processors belonging to categories such as readers, manipulators, learners,
+predictors, etc. Each processor has a configuration dialog associated with it
+that allows user to set its parameters. Processors are connected through data
+and/or control connections. A highlight of knime is support for shared- and
+distributed-memory workflow enactment and support for parallel workflow loops
+and subworkflows. The loops in Knime manifest themselves in the form of
+subworkflows (called metanodes in Knime) by a repeated enactment of the same
+subworkflow.
+%
+\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based
+script oriented workflow language used to encode parallel pipelines. StarFlow
+handles data as well as control flow using a combination of static, dynamic and
+user annotations analysis. A parallelization engine uses data and control
+dependencies in a workflow in order to build a parallel profile of the
+workflow. StarFlow has been shown to be successfully interface with the SunGrid
+\cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
+The key feature of StarFlow is its capability to execute workflows to maximum
+possible parallelism. While StarFlow provides a script-based compact and
+semiautomatic workflow composition environment, currently it lacks a
+sophisticated GUI for workflow composition.
+
+\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
+bioinformatics domain SWE based framework developed at the Penn State
+University, Pennsilvania. It provides a webportal based environment to build
+complex pipelines from an available repository of tools
+(\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided
+by Galaxy using their own Galaxy server or a web-server exposing the webservice
+descriptions through public WSDLs. One of the major use of Galaxy is to achieve
+a mix and match of data analysis tools in order to analyze a vast repository of
+genome based database.
 
@@
 
-Shown in Table \ref{tbl:scalability} is a summary of the Scalability and optimization mechanisms of the studied SWEs.
-The scalability features considered are through three levels of parallelism in the form of data, workflow and pipeline.
-The optimization features considered are the presence of mechanisms by which static and runtime optimizations are achieved.
+Shown in Table \ref{tbl:scalability} is a summary of the Scalability and
+optimization mechanisms of the studied SWEs. The scalability features
+considered are through three levels of parallelism in the form of data,
+workflow and pipeline. The optimization features considered are the presence of
+mechanisms by which static and runtime optimizations are achieved.
+
 %
@@
 
-Table \ref{tbl:datadesc} presents a summary of the data management mechanisms of the studied SWEs.
-Language and resource level data management mechanisms are highlighted.
-At the language level, data description as coupled with that of the workflow, data types and structures are considered for the SWE studied.
-At the resource level, the major resource types supported and their access mechanisms are highlighted.
-We distinguish the remote form of Grid resource types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those resource access.
-Other resource types such as data repositories or file catalogues are the resource types specific to a given SWE.
-For instance, Askalon uses a specifically configured data repository to access data and application binaries.
+Table \ref{tbl:datadesc} presents a summary of the data management mechanisms
+of the studied SWEs. Language and resource level data management mechanisms are
+highlighted. At the language level, data description as coupled with that of
+the workflow, data types and structures are considered for the SWE studied. At
+the resource level, the major resource types supported and their access
+mechanisms are highlighted. We distinguish the remote form of Grid resource
+types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those
+resource access. Other resource types such as data repositories or file
+catalogues are the resource types specific to a given SWE. For instance,
+Askalon uses a specifically configured data repository to access data and
+application binaries.
+
 %Replication is one of the solution.
@@
 \section{Workflow Language Expressiveness} \label{sec:lang}
-A workflow language is the medium of expression of the task a workflow is supposed to accomplish.
-A workflow language acts as an interface between the workflow enactor and an end-user.
-Apart from correctly expressing the flow of execution and data, a workflow language must be able to link with the application code in a transparent way.
+A workflow language is the medium of expression of the task a workflow is
+supposed to accomplish. A workflow language acts as an interface between the
+workflow enactor and an end-user. Apart from correctly expressing the flow of
+execution and data, a workflow language must be able to link with the
+application code in a transparent way.
 
@@
 
-Table \ref{tbl:expressive} shows a summary of language expressiveness capabilities of the SWEs studied.
-We consider the control structures: conditionals and loops, support for arrays and iterators \cite{montagnat-glatard-etal:2006} as the expressive features of SWEs.
+Table \ref{tbl:expressive} shows a summary of language expressiveness
+capabilities of the SWEs studied. We consider the control structures:
+conditionals and loops, support for arrays and iterators
+\cite{montagnat-glatard-etal:2006} as the expressive features of SWEs.
 
 \section{Usability} \label{sec:usable}
-A workflow environment influences the end-user comfort and convenience towards the goal of efficiently organizing and deploying a scientific experiment involving heavy usage of computational resources.
-However, there is no single factor that determines the usability of a SWE.
-A combination of several qualitative and quantitative factors play a role in determining how usable a given SWE is.
+A workflow environment influences the end-user comfort and convenience towards
+the goal of efficiently organizing and deploying a scientific experiment
+involving heavy usage of computational resources. However, there is no single
+factor that determines the usability of a SWE. A combination of several
+qualitative and quantitative factors play a role in determining how usable a
+given SWE is.
+
 \begin{table}[ht!]
@@
 experiments over the web.
+
+% -------------------- Fit here as not main content ------------------------
+
+Business Process Execution Language (BPEL) \cite{oasis:2007} is a workflow
+language pioneered by IBM Corporation and used predominantly to define business
+workflows. It is now a broad standard for business workflow applications. BPEL
+is an open language and commercial enactors are available that provide
+enactment functionality for the BPEL workflows. There has been attempts to use
+BPEL or its extended version in the scientific applications
+\cite{slominski:2007,bpel}. However, the language does not suit well to the
+data-intensive, grid workflows and has a limited adoption in the scientific
+community
+\cite{tan-missier-etal:2010,barga-gannon:2007,scherp-hoing-etal:2009,emmerich-butchart-etal:2005}.
+Apart from that, a lack of an open-source workflow engine with graphical
+interface also contributes to its limited acceptance in academic research.
+However, BPEL has been famously unsuitable for data intensive scientific
+applications  BPEL lacks different user-defined strategies to effect multiple
+invocations of each processor according to the application semantics. BPEL also
+lacks data streams and implicit parallelism forcing a user to explicitly define
+parallelism using one of its parallel processing constructs such as
+\texttt{foreach}.
 

### 2014-04-06 paper.tex
drop GWES

@@
 % Makeflow
+% Kepler/Ptolemy
 % Dryad
 % Science pipes
-% GWES
 % Spark
@@
 \section{Objectives and Organization}
-The present work analyzes contemporary SWEs used in the context of e-science to
-find out the capabilities and key features in terms of the characteristics
-described in the introduction part of this paper. The rest of this paper is
-organized as follows: Section \ref{sec:overview} presents an overview of the
-key SWEs studied in this work. Section \ref{sec:scal} discusses the
-optimization features and resulting scalable performance as observed in the
-SWEs studied. Section \ref{sec:data} studies the data management and
-description aspects of the SWEs. Section \ref{sec:inter} presents a study of
-SWEs approaches to interface the large-scale computing infrastructures. Section
-\ref{sec:lang} details the language features and expressiveness of the SWEs
-studied. Section \ref{sec:usable} discusses the usability aspects of the SWEs.
-Section \ref{sec:sum} presents a summary and a classification of all the
-aspects of SWEs studied in this work. Throughout the paper, we also present
-tables summarizing the aspects of SWEs considered in respective sections. It
-was not possible to obtain an exhaustive information about all the SWEs
-considered in this work since we rely on the peer reviewed publications for
-information on a given SWE. Some features, such as extendability of SWE
-strongly depend on the extension points designed in the code; consequently, we
-consider it present in only those SWEs that have a documented description or
-demonstration of the given feature.
+The present work analyzes contemporary SWEs in order to find the capabilities
+and key features in terms of the characteristics described in the introduction
+part of this paper. The rest of this paper is organized as follows: Section
+\ref{sec:overview} presents an overview of the key SWEs studied in this work.
+Section \ref{sec:scal} discusses the optimization features and resulting
+scalable performance as observed in the SWEs studied. Section \ref{sec:data}
+studies the data management and description aspects of the SWEs. Section
+\ref{sec:inter} presents a study of SWEs approaches to interface the
+large-scale computing infrastructures. Section \ref{sec:lang} details the
+language features and expressiveness of the SWEs studied. Section
+\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
+\ref{sec:sum} presents a summary and a classification of all the aspects of
+SWEs studied in this work. Throughout the paper, we also present tables
+summarizing the aspects of SWEs considered in respective sections. It was not
+possible to obtain an exhaustive information about all the SWEs considered in
+this work since we rely on the peer reviewed publications for information on a
+given SWE. Some features, such as extendability of SWE strongly depend on the
+extension points designed in the code; consequently, we consider it present in
+only those SWEs that have a documented description or demonstration of the
+given feature.
 
@@
 
-% Put this somewhere else, either in the analysis or may be in the conclusions and summary.
-%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language developed as a part of the current work is similar in spirit to the SwiftScript.
-Similar to SwiftScript, gscript adopts futures to express asynchronous execution in a script-based language.
-A SwiftScript workflow contains statically typed datasets that can be specified using ``annotations'' in the script.
-Unlike SwiftScript, gscript does provide user-defined strategies in order to effect data combinations in the workflow.
-SwiftScript provides for fault-tolerance and exception handling at run-time which is lacking currently in gscript.
-The underlying datasets organization in swift is represented using an XDTM XML-representation.
-This means that effectively, a workflow is SwiftScript plus the XDTM data representation.
-A workflow is simpler and compact in gscript as its data description is embedded within a workflow representation.
-SwiftScript supports creation of nested-workflows in the form of compound procedures.
-Conditional and Loop execution is supported using the if and for/while statements in both the languages.
+% Put this somewhere else, either in the analysis or may be in the conclusions
+% and summary.
+%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language
+%developed as a part of the current work is similar in spirit to the
+%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
+%asynchronous execution in a script-based language. A SwiftScript workflow
+%contains statically typed datasets that can be specified using ``annotations''
+%in the script. Unlike SwiftScript, gscript does provide user-defined
+%strategies in order to effect data combinations in the workflow. SwiftScript
+%provides for fault-tolerance and exception handling at run-time which is
+%lacking currently in gscript. The underlying datasets organization in swift is
+%represented using an XDTM XML-representation. This means that effectively, a
+%workflow is SwiftScript plus the XDTM data representation. A workflow is
+%simpler and compact in gscript as its data description is embedded within a
+%workflow representation. SwiftScript supports creation of nested-workflows in
+%the form of compound procedures. Conditional and Loop execution is supported
+%using the if and for/while statements in both the languages.
 
@@
 diagrams.
-
-%
-\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service
-(GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First
-institute. It implements a dynamic workflow composition and enactment model
-based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}.
-GWES enables key workflow development and deployment lifecycle phases including
-visual composition, data-staging, Grid middleware interfacing and execution
-monitoring. Currently, GWES is intensively being used in the field of Medical
-Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany.
-It has been successfully interfaced with the German D-Grid infrastructure.
-%
 
@@
 Galaxy/- &-- &-- &\checkmark &-- &-- &-- \\\hline
-GWES/GWorkflowDL &-- &-- &\checkmark &-- &\checkmark &-- \\\hline
 MOTEUR2/Gwendia~ & \checkmark &\checkmark & \checkmark &-- &\checkmark &-- \\\hline
@@
 StarFlow/-&Strong&Python-based&Python-based&--&--\\\hline
-GWES/GWorkflowDL&Strong&java-based&java-based&--&--\\\hline
 % 
@@
 StarFlow/- &Interfacing External Schedulars & Generic Clouds \\\hline
-GWES/GWorkflowDL&Globus WS-Gram& German D-Grid \\\hline
 Pegasus/DAX & Using Condor & OSG (www.opensciencegrid.org) \\\hline
@@
 StarFlow/- & Conditionals and Loops & -- & -- \\\hline
-GWES/GWorkflowDL & Conditionals& -- & -- \\\hline
 Pegasus/DAX & -- & \checkmark & -- \\\hline
@@
 StarFlow/- & Text & \checkmark & -- \\\hline
-GWES/GWorkflowDL & GUI & \checkmark & \checkmark \\\hline
 Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
@@
 \texttt{foreach}.
+%
+\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service
+(GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First
+institute. It implements a dynamic workflow composition and enactment model
+based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}.
+GWES enables key workflow development and deployment lifecycle phases including
+visual composition, data-staging, Grid middleware interfacing and execution
+monitoring. Currently, GWES is intensively being used in the field of Medical
+Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany.
+It has been successfully interfaced with the German D-Grid infrastructure.
+%
 

### 2014-04-07 paper.tex
new bib entries citing recent publications

@@
 % DaGue
+% Wings
 
@@
 %
+% Goals and benefits of a workflow management system
 The goals of a workflow management system for scientific computing are as
@@
 \begin{enumerate}
+\item \textbf{Bulk Work.}
+Workflows are often used to perform computational work in bulk. This could be
+bulk data processing or running multiple distinct processing on a single set of
+data. In addition to carrying out such work, workflow environments provide
+oppurtunities for giving structure and order to a set of tasks and often with
+sufficient expressivity, choices and selections to perform tasks selectively.
+
 \item \textbf{User Interaction.} Adaptation to new methods of computing is not
@@
 conveniently.  These tasks include composing and marshalling of execution,
-browsing and selecting data, visualizing results. Furthermore, dissemination of
-methods and results to communities and collaborators are some of the activities
-that SWEs make convenient for end-users.
+browsing and selecting data, visualizing results. Furthermore,
+dissemination of methods and results to communities and collaborators are
+some of the activities that SWEs make convenient for end-users.
 
@@
 \end{enumerate}
+
+% How SWEs differ from traditional programming languages and runtimes
 
@@
 \item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
+%TODO: Add content about new and emerging Distributed Computing Infrastructures: Clouds, GPUs and exascale supercomputers
 Building SWE's interfaces to distributed systems is a complex task. The
@@
 application codes.
-
 %
@@
 of SWEs suitable for general purpose workflow style computations.
-% Ketanmark
 
@@
 \section{Objectives and Organization}
+% What are the goals of this work?
+% How is this paper supposed to help:
+% -- Existing scientific users who have not adopted workflows
+% -- Existing users who have adopted workflows
+% -- New users
+% -- Experts in the field
 The present work analyzes contemporary SWEs in order to find the capabilities
@@
 
+Recent advances in Kepler include: MapReduce style computation, bioinformatics
+computation, cloud computing~\cite{kepler2013}.
+
 %
@@
 %
-
 \paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific
@@
 system to reuse the data produced by other processors.
+
+%Recent advances in Swift: 
+1. Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via its
+coasters mechanism~\cite{ucc2011}.
+
+2. Swift/T~\cite{turbine2013}, runs scientific workflows over
+HPC resources. Have been shown to work on petascale supercomputers such as the
+IBM BlueGene and Cray series of machines.
 

### 2014-04-07 paper.tex
comments to sections

@@
 % Wings
+% PGRADE Portal
 
@@
 \item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
-%TODO: Add content about new and emerging Distributed Computing Infrastructures: Clouds, GPUs and exascale supercomputers
+%TODO: Add content about new and emerging Distributed Computing
+    %Infrastructures: Clouds, GPUs and exascale supercomputers
 Building SWE's interfaces to distributed systems is a complex task. The
@@
 % -- Experts in the field
+
+% Say that different workflow systems have different design goals. It is hard
+% to chose one from the existing or build your own. It is almost equally
+% complex to build own just as it is complex to create own programming
+% platform.
+We hope to provide by the way of this paper a guide for new adopters on how to chose a workflow language appropriate for their needs.
+We ask and answer the following questions of the workflow systems that we study in this paper:
+%-- What were the original design goals?
+%-- What are the main application areas that the system is currently being used?
+%-- What are the characteristics and capabilities of the language? expressivity features?
+%-- What are the characteristics and capabilities of the run-time? parallelism? concurrency? task and data grouping?
+%-- What kinds of computing and computational infrastructure does the workflow system cater to? Grids, Clouds, Clusters, Supercomputers, fast IB networks? Accelerators?
+%-- How have they evolved? What are the possible future directions?
+
 The present work analyzes contemporary SWEs in order to find the capabilities
@@
 %
-\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the
-University of North Carolina, Wilmington, facilitates a graphical composition
-and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
-Ptolemy II framework. GridNexus workflows are expressed in an XML-based
-scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
-customizable and extensible scripting language allowing users to write their
-own workflow processor types. GridNexus workflows are grid oriented and can be
-deployed on grid resources accessed through a persistent WS-RF (Web Services
-Resource Framework) interface.
+\paragraph{Dryad/--}
 
@@
 %The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language
-%developed as a part of the current work is similar in spirit to the
+%is similar in spirit to the
 %SwiftScript. Similar to SwiftScript, gscript adopts futures to express
-%asynchronous execution in a script-based language. A SwiftScript workflow
-%contains statically typed datasets that can be specified using ``annotations''
-%in the script. Unlike SwiftScript, gscript does provide user-defined
-%strategies in order to effect data combinations in the workflow. SwiftScript
-%provides for fault-tolerance and exception handling at run-time which is
-%lacking currently in gscript. The underlying datasets organization in swift is
-%represented using an XDTM XML-representation. This means that effectively, a
-%workflow is SwiftScript plus the XDTM data representation. A workflow is
+%asynchronous execution in a script-based language. A Swift script 
+%contains statically typed datasets that can be declared
+%in the script. Unlike Swift script, gscript does provide user-defined
+%strategies in order to effect data combinations in the workflow. The underlying datasets organization in Swift is
+%represented using XML-representation. This means that effectively, a
+%workflow is a DAG. A workflow is
 %simpler and compact in gscript as its data description is embedded within a
-%workflow representation. SwiftScript supports creation of nested-workflows in
-%the form of compound procedures. Conditional and Loop execution is supported
-%using the if and for/while statements in both the languages.
-
-%
+%workflow representation. Conditional and Loop execution is supported
+%using the if and foreach statements in both languages.
+%
+
 \paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
@@
 %
+
+The GridNexus \cite{GridNexus} SWE, developed at the
+University of North Carolina, Wilmington, facilitates a graphical composition
+and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
+Ptolemy II framework. GridNexus workflows are expressed in an XML-based
+scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
+customizable and extensible scripting language allowing users to write their
+own workflow processor types. GridNexus workflows are grid oriented and can be
+deployed on grid resources accessed through a persistent WS-RF (Web Services
+Resource Framework) interface.
 

### 2014-04-08 paper.tex
cleaning up by removing old workflow descriptions to the bottom. Hopefully will be able to salvage something out of them

@@
 
+3. GeMTC project investigating running many-task applications over GPUs via Swift/T.
+
 % Put this somewhere else, either in the analysis or may be in the conclusions
@@
 %
-\paragraph{DaGue/--} Bioflow \cite{bioflow} is a query based workflow language.
-It is motivated by the fact that bioinformatics applications are query based
-applications that benefit from the complex database queries or saved database
-`views'. A typical Bioflow workflow pipelines a sequence of filter queries in
-order to get the desired results. Bioflow is mainly utilized in the area of
-bioinformatics and querying genome databases. Additionally, a Bioflow workflow
-provides mechanisms for data manipulation and updation in the existing
-databases. It also includes advanced features such as parallel processing of a
-set of mutually independent database queries. The execution model is based on
-the popular database query language SQL (Structured Query Language). Processes
-can be defined in Bioflow which are processor counterparts in workflow paradigm
-and similar to stored-procedures in SQL jargon.
-
-Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical
-representation of bioflow workflows. A Vizbuilder workflow links to an
-underlying bioflow engine and provides an interface to submit the bioflow
-workflows for execution.
-
-%
-\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative
-workflow language that is inspired by the functional programming paradigm. A
-Martlet workflow conveniently enables distributed parameter sweep applications
-wherein the language intuitively defines the computation and a matching with
-the distributed resources can be carried out at workflow execution time. List
-processing applications where a sequence of application components work over an
-array of data in a Multi-data multi-instruction style setup is well-suited to
-the Martlet's model of computation.
-
-%
-\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining
-system for data intensive applications developed at the University of Konstanz.
-It is a java-based platform with a Graphical Interface that is based on eclipse
-(\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists
-of processors belonging to categories such as readers, manipulators, learners,
-predictors, etc. Each processor has a configuration dialog associated with it
-that allows user to set its parameters. Processors are connected through data
-and/or control connections. A highlight of knime is support for shared- and
-distributed-memory workflow enactment and support for parallel workflow loops
-and subworkflows. The loops in Knime manifest themselves in the form of
-subworkflows (called metanodes in Knime) by a repeated enactment of the same
-subworkflow.
-%
-\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based
-script oriented workflow language used to encode parallel pipelines. StarFlow
-handles data as well as control flow using a combination of static, dynamic and
-user annotations analysis. A parallelization engine uses data and control
-dependencies in a workflow in order to build a parallel profile of the
-workflow. StarFlow has been shown to be successfully interface with the SunGrid
-\cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
-The key feature of StarFlow is its capability to execute workflows to maximum
-possible parallelism. While StarFlow provides a script-based compact and
-semiautomatic workflow composition environment, currently it lacks a
-sophisticated GUI for workflow composition.
+\paragraph{DaGue/--}
+
+%
 
@@
 Vizbuilder/bioflow &-- &\checkmark &-- &--&\checkmark &Query optimization\\\hline
-Martlet/- & \checkmark & \checkmark & -- &-- &\checkmark&-- \\\hline
-Knime/- & \checkmark &\checkmark &\checkmark & -- & \checkmark & -- \\\hline
-StarFlow/- &\checkmark & \checkmark & \checkmark & \checkmark & \checkmark & -- \\\hline
 Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark &AI-based techniques \\\hline
@@
 
+%
+Bioflow \cite{bioflow} is a query based workflow language.
+It is motivated by the fact that bioinformatics applications are query based
+applications that benefit from the complex database queries or saved database
+`views'. A typical Bioflow workflow pipelines a sequence of filter queries in
+order to get the desired results. Bioflow is mainly utilized in the area of
+bioinformatics and querying genome databases. Additionally, a Bioflow workflow
+provides mechanisms for data manipulation and updation in the existing
+databases. It also includes advanced features such as parallel processing of a
+set of mutually independent database queries. The execution model is based on
+the popular database query language SQL (Structured Query Language). Processes
+can be defined in Bioflow which are processor counterparts in workflow paradigm
+and similar to stored-procedures in SQL jargon.
+
+Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical
+representation of bioflow workflows. A Vizbuilder workflow links to an
+underlying bioflow engine and provides an interface to submit the bioflow
+workflows for execution.
+
+\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative
+workflow language that is inspired by the functional programming paradigm. A
+Martlet workflow conveniently enables distributed parameter sweep applications
+wherein the language intuitively defines the computation and a matching with
+the distributed resources can be carried out at workflow execution time. List
+processing applications where a sequence of application components work over an
+array of data in a Multi-data multi-instruction style setup is well-suited to
+the Martlet's model of computation.
+
+%
+\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining
+system for data intensive applications developed at the University of Konstanz.
+It is a java-based platform with a Graphical Interface that is based on eclipse
+(\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists
+of processors belonging to categories such as readers, manipulators, learners,
+predictors, etc. Each processor has a configuration dialog associated with it
+that allows user to set its parameters. Processors are connected through data
+and/or control connections. A highlight of knime is support for shared- and
+distributed-memory workflow enactment and support for parallel workflow loops
+and subworkflows. The loops in Knime manifest themselves in the form of
+subworkflows (called metanodes in Knime) by a repeated enactment of the same
+subworkflow.
+%
+\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based
+script oriented workflow language used to encode parallel pipelines. StarFlow
+handles data as well as control flow using a combination of static, dynamic and
+user annotations analysis. A parallelization engine uses data and control
+dependencies in a workflow in order to build a parallel profile of the
+workflow. StarFlow has been shown to be successfully interface with the SunGrid
+\cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
+The key feature of StarFlow is its capability to execute workflows to maximum
+possible parallelism. While StarFlow provides a script-based compact and
+semiautomatic workflow composition environment, currently it lacks a
+sophisticated GUI for workflow composition. 
 % ---------------------------------------------------------------------------

### 2014-04-08 paper.tex
added citation for the Wings system
dropping some systems from tables

@@
 
+\paragraph{Wings/--} Wings is a Workflow system which does software choreography based on AI techniques~\cite{wings2013}
 %
@@
 Most workflow enactors exploit the parallelism at three-levels. First, at the
-workflow-level wherein the independent branches of a workflow are enacted
+task-level wherein the independent branches of a workflow graph are enacted
 concurrently. Second, at the data-level, where a same processor is instantiated
@@
 the low-level parallelism that is encoded within the application components
-independent of the workflow itself. At this level, the workflow enactor may
+independent of the workflow. At this level, the workflow enactor may
 need to make a matching based upon the application type. For instance, a
-workflow enactor may schedule the MPI-based processors on the resources that
+workflow enactor may schedule the GPU-based processors on the resources that
 indicate as possessing such capabilities.
@@
 P-GRADE/- &\checkmark &\checkmark &\checkmark &--&--&--\\\hline
-Vizbuilder/bioflow &-- &\checkmark &-- &--&\checkmark &Query optimization\\\hline
 Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark &AI-based techniques \\\hline
@@
 considered are through three levels of parallelism in the form of data,
-workflow and pipeline. The optimization features considered are the presence of
+task and pipeline. The optimization features considered are the presence of
 mechanisms by which static and runtime optimizations are achieved.
@@
 &&&&File Catalogs&Protocols\\\hline
-%
-Vizbuilder/bioflow&Strong&SQL-types&--&Database&SQL,PL/SQL\\\hline
-%
-\multirow{2}{*}{Martlet/-}&Loose&URI&Lists&Distributed& By\\
-&&&&Filesystems&Reference\\\hline
-% 
-\multirow{2}{*}{Knime/-}&Loose&String, Integer,&Filestreams,&FileSystems,&Datamining\\
-&&Double&Arrays&Databases&Tools\\\hline
-% 
-StarFlow/-&Strong&Python-based&Python-based&--&--\\\hline
 %
@@
 P-GRADE/- & Inherent & NGS \\\hline
-Vizbuilder/bioflow & DCI Independent & None Inherently \\\hline
-Martlet/- & interfaces multiple middlewares & Generic Grids \\\hline
-Knime/- & Extensions & SGE Clusters \\\hline
-StarFlow/- &Interfacing External Schedulars & Generic Clouds \\\hline
 Pegasus/DAX & Using Condor & OSG (www.opensciencegrid.org) \\\hline
@@
 while at the same time dealing with parallelism and array processing. They also
-fall short of providing implicit parallelism and must be defined explicitly by
+fall short of providing implicit parallelism which must be defined explicitly by
 the user designing workflows. Description of remote data is another important
@@
 P-GRADE/- & Conditionals & -- & \checkmark \\\hline
-Vizbuilder/bioflow & Conditionals and Loops & -- & -- \\\hline
-Martlet/- & Conditionals and Loops & -- & -- \\\hline
-Knime/- & Loops & -- & \checkmark \\\hline
-StarFlow/- & Conditionals and Loops & -- & -- \\\hline
 Pegasus/DAX & -- & \checkmark & -- \\\hline
@@
 P-GRADE/- & Web Portal & \checkmark & -- \\\hline
-Vizbuilder/bioflow & GUI/Text & \checkmark & -- \\\hline
-Martlet/- & Text & -- & -- \\\hline
-Knime/- & GUI & \checkmark & -- \\\hline
-StarFlow/- & Text & \checkmark & -- \\\hline
 Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
@@
  child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{pipeline}}}
-	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{workflow}}}
+	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{task}}}
 % 	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\small{activity}}}
@@
 
-Figure \ref{fig:taxonomy} shows a taxonomy of SWE features studied in this paper.
-We studied twenty SWEs from different scientific fields developed worldwide.
-Some trends as observed in the approaches of the SWEs development are noted below:
+Figure \ref{fig:taxonomy} shows a taxonomy of SWE features studied in this
+paper. We studied twenty SWEs from different scientific fields developed
+worldwide. Some trends as observed in the approaches of the SWEs development
+are noted below:
+
 \begin{itemize}
- \item A recent surge in distributed computing approach of Cloud Computing has given rise to adoption of new enactment technologies. 
- \item SWEs tend to offer complementary properties which has led to several collaborations among the developing groups.
-Notable among these are the recent ones between Taverna and Galaxy \cite{mohamed-alaa-etal:2010} and Kepler and Pegasus \cite{mandal-deelman-etal:2007}.
- \item Collection oriented SWEs wherein arrays of data are treated as first-class entities has been seen a significant trend among the contemporary SWEs.
- \item A trend of increasing number of SWEs providing a script-based workflow composition interface solely or in combination with a visual interface in the form of GUI has been observed.
- \item Workflow systems originally developed for business domains have been introduced to scientific domains, such as BPEL and Trident.
+ \item A recent surge in distributed computing approach of Cloud Computing has
+     given rise to adoption of new enactment technologies.
+
+ \item SWEs tend to offer complementary properties which has led to several
+     collaborations among the developing groups. Notable among these are the
+     recent ones between Taverna and Galaxy \cite{mohamed-alaa-etal:2010} and
+     Kepler and Pegasus \cite{mandal-deelman-etal:2007}.
+
+ \item Collection oriented SWEs wherein arrays of data are treated as
+     first-class entities has been seen a significant trend among the
+     contemporary SWEs.
+
+ \item A trend of increasing number of SWEs providing a script-based workflow
+     composition interface solely or in combination with a visual interface in
+     the form of GUI has been observed.
+
+ \item Workflow systems originally developed for business domains have been
+     introduced to scientific domains, such as BPEL and Trident.
+
 \end{itemize}

### 2014-04-15 paper.tex
double space
comments
new fig and desc

@@
 
+\usepackage{setspace}
+%\doublespacing
+\linespread{2}
+
 % Metadata Information
@@
 \acmMonth{3}
+
+\newcommand{\jm}[1]{
+    \noindent{\textcolor{red}{\bf\em #1 -- Johan}}
+}
+\newcommand{\km}[1]{
+    \noindent{\textcolor{blue}{\bf\em #1 -- Ketan}}
+}
 
@@
 
-In this work we conduct a literature study of the key features of the
-contemporary scientific workflow environments. The criteria for study are
-defined. An overview of studied scientific workflow environments is given
-followed by the identification and description of the features offered by each
-of the environment. A tabulated summary for each feature gives a high-level
-view of the offered features. Finally, a taxonomy is proposed placing these
-features into respective places in a hierarchy.
+\km{TODO: Expand in the end}
+
+In this work we conduct a literature survey of the key features of the
+contemporary scientific workflow environments. We define the criteria for the
+survey based chiefly on user needs arising from diverse applications and
+infrastructure. An overview of studied scientific workflow environments is
+presented followed by the identification and description of the features offered by
+each of the environment. A tabulated summary for each feature class gives a
+high-level view of the offered features. Finally, a taxonomy is proposed
+placing these features into respective places in a hierarchy.
+
+\jm{Test comment}
 
@@
 
-\section{Architecture of a Generic SWE}
-Figure \ref{fig:wfms} shows a high-level architecture of a SWE. It shows key
+\section{Conceptual Overview}
+In this section we present a conceptual overview of the subjects of this study--Workflows and Scientific Workflow Environments.
+
+\subsection{Workflow}
+Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
+Graph (DAG). For the purpose of this text, we define a workflow as a
+computation with at least two discrete tasks either forming distinct
+computational stages or running for two or more distinct datasets as
+parameters. Consequently, a one-off task or running a task for a single copy of
+data is \emph{not} a workflow. For the rest of this paper, we will use the above
+definition of a workflow.
+%
+\begin{figure}[htb]
+\begin{center}
+\includegraphics[width=4cm]{figures/workflow}
+\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
+\label{fig:wf}
+\end{center}
+\end{figure}
+%
+\subsection{Architecture of a Generic SWE}
+Figure~\ref{fig:wfms} shows a high-level architecture of a SWE. It shows key
 components and their interaction in a SWE. These components are the focus of

### 2014-04-16 paper.tex
edits

@@
 level. However, traditional scripts do not deal with parallelism and dataflow
-aspects of workflow paradigm that makes them a limited contenders. 
+aspects of workflow paradigm which makes them a strong contenders for the
+purpose of high-level application programmers.
 %
@@
 Workflows are often used to perform computational work in bulk. This could be
-bulk data processing or running multiple distinct processing on a single set of
-data. In addition to carrying out such work, workflow environments provide
-oppurtunities for giving structure and order to a set of tasks and often with
-sufficient expressivity, choices and selections to perform tasks selectively.
+bulk data processing or running multiple distinct processing tasks on a single
+set of data. In addition to carrying out such work, workflow environments
+provide oppurtunities for giving structure and order to a set of tasks and
+often with sufficient expressivity, choices and selections to carry out a bulk
+of work selectively. Some of the examples of such work are \emph{parameter
+sweeps} where tasks are run over a series of parameter values, \emph{ensemble
+studies} where tasks are carried out over discrete groups of parameters over a
+space.
 
@@
 entities.  SWEs are an ideal platform enabling users to perform many tasks
-conveniently.  These tasks include composing and marshalling of execution,
-browsing and selecting data, visualizing results. Furthermore,
-dissemination of methods and results to communities and collaborators are
-some of the activities that SWEs make convenient for end-users.
+conveniently.  These tasks include visually formulating a task, composing and
+marshalling of workflow, browsing and selecting data, visualizing and sharing
+results.
 
@@
 adaptation of new techniques. Clouds are the latest examples of new techniques
-of computational platform access while GPGPUs examples in upgradation of
+of computational platform access while GPGPUs are examples of upgradation of
 existing processing environments. Additionally, with the advent of fast and
 dedicated networks such as ESNet and federated visualization resources, remote
-visualization is gaining prominence among scientific users. SWEs play important
-role as platforms to give transparent and hassle free access to such novel and
-multiple environments to users by hiding the technicalities of access methods.
+visualization is gaining prominence among scientific users. SWEs play
+important role as platforms to give transparent and hassle free access to such
+novel and multiple environments to users by hiding the technicalities of
+access methods and usage and configuration related nuances.
 
 \item \textbf{Repeatable Execution.} Scientific users often require a
-repeatable execution of same workflow for reproduction of results and
-demonstration of methods. SWEs can record and save workflows, associated
-parameters, source data and historical results to aid in such activities.
-Additionally, SWEs facilitate an incremental development of such methods
-leading to augmentation of past efforts to current and future developments.
+ repeatable execution of same application(with same parameters and input data)
+ for reproduction of results and demonstration of methods. SWEs can record and
+ save workflows, associated parameters, input data and historical results to
+ aid in such activities.  Additionally, SWEs facilitate an incremental
+ development of these methods leading to augmentation of past efforts to current
+ and future developments.
 
@@
 plays key roles. SWEs can be integrated or extended with metadata and
-provenance techniques rendering powerful computing traceability features.
-
-\item \textbf{Sharing and Collaboration.} Workflow managers are often targetted
-towards a community of users. For instance, the Galaxy workflow system is
-targetted specifically for the bioinformatics users. Galaxy attracts many users
-from the same community thus promoting sharing and collaboration via a common
-computational platform. Another such example is Taverna which is geared towards
-sharing and collaborative workflow oriented scientific computation enabling
-users to discover new services introduced to a community.
+provenance techniques rendering powerful computing and traceability features.
+
+\item \textbf{Sharing and Collaboration.} Workflow systems are often targetted
+towards a particular community of users. Furthermore, dissemination of methods and results
+to communities and collaborators are some of the activities that SWEs make
+convenient for end-users. For instance, the Galaxy workflow system is targetted
+specifically for the bioinformatics users encouraging them to collaborate over
+problems and share results. Another such example is Taverna which is geared
+towards sharing and collaborative workflow oriented scientific computation
+enabling users to discover new services introduced to a community. Sharing and
+collaboration of applications and related data have become ubiquitious with the
+rise of social web. Initiatives such as
+\emph{myExperiment}~\cite{deroure-goble-etal:2008} and recently
+\emph{gitHub}~\cite{github-web} have enabled easy sharing and collaboration of
+reproducible workflows on the social web.
 
@@
 
-% How SWEs differ from traditional programming languages and runtimes
-
-As a consequence to these design goals, the SWEs differ in many aspects with
-respect to traditional programming environments. The following list identifies
-the major differences between traditional and the SWE programming
-paradigm:
+The above described goals also form a basis for the need for a SWE. The need
+for the use of SWE can arise as a result of one or more of such requirements.
+Traditional programming platforms are not suitable to address these
+requirements as they operate at a different level of abstraction. SWEs differ
+in many respects among themselves as well as with respect to traditional
+programming environments. The following list identifies the major differences
+between traditional and the SWE programming paradigm:
 
@@
 extensive usage of distributed resources than conventional programs. Owing
-to the large-scale experiments, for which scientific users turn to
-automation, their demands for computational power often exceeds that of a
-single personal computer. Such applications are often composed of multiple,
-distinct computational stages making them well-suited to the workflow
-paradigm. These properties also make them ideal candidates for a
-large-scale distributed computing environment. The case for suitability of
-SWEs on large-scale distributed infrastructures has been well-argued upon
+to the large-scale experiments, for which scientific users turn to automation,
+their demands for computational power often exceeds that of a single personal
+computer. These properties make them ideal candidates for a large-scale
+distributed computing environment. The case for suitability of SWEs on
+large-scale distributed infrastructures has been well-argued upon
 in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
-ensure scalable and optimized performance is vital.
+ensure scalable and optimized performance is vital. Additionally, most SWEs are
+capable of executing workflows at small scale with minimal tuning. This
+multi-scalability enables quick and easy conceptualization and prototyping of
+new workflows.
 %
@@
 \item \textbf{Data Management.} The scale of data transfer is much higher in
-SWEs from one execution site to others as well as among storage resources
-while it is lower (ranging from variables to small data structures) in
-programs.
+SWEs from one execution site to others as well as among storage resources while
+it is lower (ranging from variables to small data structures) in programs.
+Owing to such scales, applications often need access to special data movement
+and access protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
+shipped with capabilities to handle such protocols.
 %
@@
     %Infrastructures: Clouds, GPUs and exascale supercomputers
-Building SWE's interfaces to distributed systems is a complex task. The
+Building SWE's interface to distributed systems is a complex task. The
 distributed nature of resource utilization and asynchronous execution of
 workflow requires a set of interdependent parameters to be configured
-correctly. Further complexity is added because of complex dataflow
-requirements, interactions with diverse tools and linking with legacy
-application codes.
-%
+correctly. Modern computing infrastructures such as Clouds, GPUs and
+supercomputers with hybrid nodes pose new challenges of interface development.
+As SWEs involve running many tasks over an infrastructure it becomes crucial to
+interface with these systems such that the tasks are scheduled in an optimal
+manner. Many SWEs implement
+\emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
+reserve resources via pseudo-tasks that act as placeholders for real tasks.
+Additionally, decision to manage working data storage on disc, memory and cache
+are some of the distinctions between a traditional programming and workflow
+platforms.
+
 \item \textbf{Expressiveness.} Programming languages are inherently control
@@
 execution in a predefined sequence. However, in SWEs, the flow is often
-determined by the data availability for execution. This model of
-computation is known as dataflow computing. A workflow thus does not follow
-a fixed order of execution and different execution sequences are possible
-which are dominated by dataflow. The unit of work in SWEs is at a higher
-level of granularity. A unit of execution is coarse-grained, and often
-termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
-community. This notion of high-level granularity drives the depth to which
-a workflow can be expressed in a given language form. In the rest of this
-paper, this unit of work will be referred to as \textit{processor}. On the
-other hand, a unit of execution in most programming languages is a much
-simpler and finer executable statement that accomplishes a small
-computational task. This presents a whole new set of requirements for
-expressing the processors and how they should interact and operate upon
-data.
+determined by the data availability for execution. This model of computation is
+known as dataflow computing. A workflow thus does not follow a fixed order of
+execution. Different execution sequences are possible which are dominated by
+dataflow. The unit of work in SWEs is at a higher level of granularity. A unit
+of execution is coarse-grained, and often termed as ``processor'', ``task'',
+``service'' or ``actor'' in the SWE community. This notion of high-level
+granularity drives the depth to which a workflow can be expressed in a given
+language form. In the rest of the paper, we refer to a unit of execution of a
+workflow system as \emph{processor}. On the other hand, a unit of execution in
+most programming languages is a much simpler and finer executable statement
+that accomplishes a small computational task. This presents a whole new set of
+requirements for expressing the processors and how they should interact with
+computational resources and operate upon data.
 
@@
 most often written by expert programmers. Thus, they have a text-based
-composition environment albeit with visually assisting programming
-environments such as IDEs. However, they are seldom to the extent that a
-non-expert user can graphically compose programs.
+composition environment albeit with visually assisting programming environments
+such as IDEs. However, they are seldom to the extent that a non-expert user can
+graphically compose programs. While a graphical interface enhances usability, a
+text-based workflow environment can arguably give better control to an
+experienced user. Consequently, many workflow systems offer a dual interface of
+graphical and text medium to compose workflows.
 
@@
 aspects we cover in the current work. We do not include other interesting SWE
-features such as provenance and interoperability in this analysis as they are
-out of scope of the objectives of the current paper.
+features such as application predictability and dynamic restructure of DAGs in
+this analysis as they are out of scope of the objectives of the current paper.
 
@@
 \subsection{Architecture of a Generic SWE}
-Figure~\ref{fig:wfms} shows a high-level architecture of a SWE. It shows key
+Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
 components and their interaction in a SWE. These components are the focus of
 interest for the present work. A \textit{user interface} interacts with the
-\textit{workflow engine} using commands and messages. The \textit{workflow
-language} becomes an intermediate representation for communication between user
-interface and the \textit{workflow engine}. The workflow engine acts as a
-compiler/interpreter and scheduler for the workflow tasks and interfaces with
-the underlying \textit{distributed resources}. It also monitors the workflow
-progress and collects the intermediate and final results.
+\textit{workflow system} using commands and messages on the inside while with
+user on the outside. The \textit{workflow language} becomes an intermediate
+representation for communication between user interface and the
+\textit{workflow system}. The workflow system acts as a compiler/interpreter
+and a high-level scheduler for the processors and interfaces with the underlying
+\textit{distributed resources}. It also monitors the workflow progress and
+collects the intermediate and final results.
 
@@
 paper, especially the workflow languages expressiveness and SWEs interactions
-with the distributed computing infrastructures. A more recent work by Goderis
-and co-authors \cite{goderis-brooks-etal:2009} studies a topic more closely
+with the distributed computing infrastructures. A more recent work by Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely
 related to the workflow language expressiveness: it compares the models of
@@
 workflow features and capabilities including SWE composition, representation
-and resource mapping capabilities. The recent survey from Stratan et. al.
-\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
+and resource mapping capabilities. The recent survey from Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
 account the latencies involved during the enactment mechanisms. A small survey
-and analysis of file-access characteristics of Data-intensive workflows has
-been done by Shibata et. al.~\cite{Shibata}. Early surveys on workflow
+and analysis of file-access characteristics for data-intensive workflows has
+been done by Shibata~\cite{Shibata}. Early surveys on workflow
 languages and suitability of general purpose languages to workflows is
@@
 addressed over the
-years~\cite{glatard-montagnat-etal:2008a,delaitre-kiss-etal:2005,rojasbalderrama-montagnat-etal:2010}.
+years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.
 
@@
 \cite{dagman} to abstractions for parallel computations such as Petri nets
-\cite{alt-hoheisel:2005}, meta-models \cite{nemo-glatard-etal:2007},
+\cite{alt-hoheisel:2005},
 data-driven languages such as Scufl \cite{turi-missier-etal:2007} and scripting
-languages such as SwiftScript \cite{swift}. Glatard's PhD thesis
+languages such as Swift script \cite{swift}. Glatard's PhD thesis
 \cite{glatard:2007} presents a taxonomy of SWE languages classifying them into
@@
 properties such as potential deadlocks, data-driven languages ease the
-description of applications logic for nonexpert users and scripting is
+description of application logic for non-expert users and scripting is
 extensively used by programmers for prototyping, etc.
@@
 and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
-possible, is produced as soon as possible. In spite of the presence of many
-different approaches to workflow languages, new application requirements
-continue to pose workflow expressiveness challenges. The community is yet to
+possible, is produced as soon as possible. Despite the presence of many
+different approaches to workflow systems, new requirements
+continue to pose workflow challenges. The community is yet to
 realize a truly generic workflow language that can express a wide variety of

### 2014-04-16 paper.tex
edits

@@
 \section{Introduction}
-
+\label{sec:intro}
 \vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
@@
 perceived and visualized. The MA
-DAG\footnote{http://graal.ens-lyon.fr/diet/workflow.html}
-\cite{caron-desprez:2005}, engine is one example of a DAG generator engine. MA
+DAG~\cite{caron-desprez:2005},\km{could not find any recent paper on MA DAG, interesting concept} engine is one example of a DAG generator engine. MA
 DAG has evolved from a regular DAG generator to a more abstract DAG generator
@@
 different approaches to workflow systems, new requirements
-continue to pose workflow challenges. The community is yet to
-realize a truly generic workflow language that can express a wide variety of
-flow patterns encountered in scientific domains.
+continue to pose workflow challenges. Some of the current challenges facing SWEs are:
+\begin{enumerate}
+    \item \textbf{Multisite execution.} Interfacing to existing computation and
+        storate infrastructure is not trivial, not to mention the new and
+        unforeseen access and compute modes. SWEs are yet to formulate a
+        generalized interface towards the major platforms such that they will
+        run on them out of the box.
+
+    \item \textbf{Manysite execution.} Even if a SWE has interfaces built for
+        more than one infrastructure, it is hard to optimally schedule large
+        workflows to many remote sites simultaneously. The problem of limited
+        control over remote sites and distributing interdependent data and
+        tasks makes it non-trivial to efficiently load-balance work among them.
+
+    \item \textbf{Expressivity.} Workflow languages have evolved over the years
+        with reasonable expressivity being supported. However, they are often
+        compared with those of the traditional programming languages. As of
+        current state of the art, the workflow languages expressivity still
+        have to catch up to the richness of programming languages across
+        different generations such as C and python.  The community is yet to
+        realize a truly generic workflow language that can express a wide
+        variety of flow patterns encountered in scientific domains.
+
+    \item \textbf{Interoperability.} Interoperable workflows will benefit to a
+        broad audience of users. Many efforts at this are made in the recent
+        past, such as by the consurtium projects and by individual players.
+        However, a truly interoperable set of standards for workflows are yet
+        to emerge.
+
+\end{enumerate}
 
@@
 
-The present work analyzes contemporary SWEs in order to find the capabilities
-and key features in terms of the characteristics described in the introduction
-part of this paper. The rest of this paper is organized as follows: Section
+In the present work we survey and study the various contemporary SWEs in order
+to find the capabilities and key features in terms of the characteristics
+described in section~\ref{sec:intro}. We believe this study will serve to
+address the following objectives:
+
+\begin{enumerate}
+  \item \textbf{Educate.} With more and more science being done with the help of
+      computational methods, SWEs can add significant value to the process.
+      They can accelerate the process, enable community driven science, drive
+      the development at a manageable level of abstraction for users.
+      However, scientific users are often not experienced programmers or
+      compute savvy. This work will serve as a useful starting guide to
+      educate science users of the available tools and techniques with
+      features that might be suitable for their methods.
+
+ \item \textbf{catalog.} We study a number of important features as offered by
+     contemporary SWEs. This will help catalog and collect the SWEs by their
+     offerings and identify mechanisms and needs for new and existing features.
+     This will also help identify current trends in scientific computing
+     community.
+
+ \item \textbf{application.} We identify the applications and application
+     classes well adapted to the workflow systems and are benefitting by them.
+     This will help readers relate to the application they are associated to.
+     We hope that readers can make a choice of which SWE to use if there is an
+     adaptation scope.
+
+ \item \textbf{evolution.} We comment on the evolution of the existing workflow
+     systems and the possible current and future development directions which
+     will help readers be aware of the general development direction of the
+     SWEs.
+
+\end{enumerate}
+
+The rest of this paper is organized as follows: Section
 \ref{sec:overview} presents an overview of the key SWEs studied in this work.
@@
 possible to obtain an exhaustive information about all the SWEs considered in
-this work since we rely on the peer reviewed publications for information on a
-given SWE. Some features, such as extendability of SWE strongly depend on the
-extension points designed in the code; consequently, we consider it present in
-only those SWEs that have a documented description or demonstration of the
-given feature.
+this work. We rely on the following sources for our information: 
+\begin{itemize}
+    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
+    \item \textbf{Technical Reports.} Technical reports and theses.
+    \item \textbf{Web Presence.} Documentation and other information from the web pages of the host SWE.
+    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
+\end{itemize}
+
+Some features, such as extendability of SWE strongly depend on the extension
+points designed in the code; consequently, we consider it present in only those
+SWEs that have an explicit documented description or demonstration of the given
+feature.
 
 \section{Overview of the Workflow Environments} \label{sec:overview}
-In this section we present an overview of the Workflow Environments studied.
-The paragraphs headers are organized by the name of the workflow environment
-followed by the language they chiefly support separated by a forward slash (/).
-The cases where the workflow language is not specifically defined or conversely
-where there are multiple engines supporting a generic workflow language are
+In this section we present an overview of the SWEs studied.  The paragraph
+headers are organized by the name of the SWE followed by the name of internal
+representation language use and support separated by a forward slash (/).  The
+cases where an intermediate language is not specifically defined or conversely
+where there are multiple systems supporting a generic representation are
 denoted by an asterisk (*) or a hyphen (-) at the respective places.
@@
 Taverna includes a GUI-based rich-client workbench for workflow design and a
-workflow enactment engine. Taverna can enact workflows expressed in the scufl
+workflow enactment runtime. Taverna can enact workflows expressed in the scufl
 language. Scufl is a near Turing complete dataflow language with well-defined
@@
 previous processor ends its execution.  The processors are of different kinds
-depending on the application code to be invoked: typically, processors are Web
-Services, with input and output \textit{ports} that correspond to the
-operation's invocations and response sequences defined in the service`s WSDL
-interface.  Java classes and local scripts can be used as workflow processors,
-as long as they expose their signature as input and output ports as required by
-the model.  The engine orchestrates the execution of the processors in a way
-that is consistent with the dependencies, and manages the flow of data through
-the processors' ports. The overall workflow execution is data-driven, with a
-push model: a processor's execution is started as soon as all of its inputs are
+depending on the application code to be invoked: typically being services, with
+input and output \textit{ports} that correspond to the operation's invocations
+and response sequences defined in the service`s interface expressed as WSDL.
+Java classes and local scripts can be used as workflow processors, as long as
+they expose their signature as input and output ports as required by the model.
+The engine orchestrates the execution of the processors in a way that is
+consistent with the dependencies, and manages the flow of data through the
+processors' ports. The overall workflow execution is data-driven, with a \emph{push}
+model: a processor's execution is started as soon as all of its inputs are
 available.
@@
 
-Taverna is not natively interfaced with a Grid infrastructure but the same can
-be achieved with the help of extending the engine through one of its SPIs.
+Taverna is not natively interfaced with external computational infrastructure but the same can
+be achieved with the help of extending the engine through one of its SPIs~\cite{maheshwari-missier-etal:2009}. 
+\km{ToDo: Taverna current activity, applications support, GPU, Cloud support}
 
@@
 developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
-MOTEUR facilitates executing workflows consisting of remote web- and
-grid-services. It exploits service parallelism at workflow level and data
-parallelism for multiple datasets.
-
-MOTEUR has sophisticated data management mechanisms within the workflow
-definition. It provides a diverse set of iterations strategies well adapted for
-different combinations of input datasets at the processor. An iteration
-strategy allows different combinations of the data arriving at a processor
-along its multiple inputs. Running legacy application code on grids has been
-eased by MOTEUR by providing a generic grid-service wrapper for these
-applications \cite{glatard-montagnat-etal:2008a}. Grid-services thus created
-can be readily deployed on the grids by the MOTEUR workflow enactor. MOTEUR
-workflows are written in the scufl language and are compatible with those
-written for the Taverna SWE. It also supports an extension of scufl language
-called the Gwendia language, developed at the same laboratory. 
+MOTEUR facilitates executing workflows consisting of remote services. It
+exploits service parallelism at workflow level and data parallelism for
+multiple datasets.
+
+MOTEUR has sophisticated language-level expressions for data management
+mechanisms within a workflow definition. It provides a diverse set of
+parallel iterators that are well adapted for different combinations of input datasets
+at the processor. An iteration strategy allows different combinations of the
+data arriving at a processor along its multiple inputs. Running legacy
+application code on distributed systems has been eased by MOTEUR by providing a
+generic grid-service wrapper for these applications
+\cite{glatard-montagnat-etal:2008a}. Services thus created can be readily
+deployed by the MOTEUR runtime. MOTEUR workflows are written in the scufl
+language and are compatible with those written for the Taverna SWE. It also
+supports an extension of scufl language called the Gwendia~\cite{gwendia}
+language, developed at the same laboratory. 
 %
@@
 components. A Triana workflow is expressed as an XML grammar similar to a WSDL
-specification for webservices \cite{taylor-wand-etal:2005}. A highlight of
+description for webservices \cite{taylor-wand-etal:2005}. A highlight of
 Triana is that it provides ready-to-use interface for a wide variety of

### 2014-04-21 paper.tex
more edits

@@
         to emerge.
-
 \end{enumerate}
@@
 
- \item \textbf{catalog.} We study a number of important features as offered by
+ \item \textbf{Catalog.} We study a number of important features as offered by
      contemporary SWEs. This will help catalog and collect the SWEs by their
@@
 
- \item \textbf{application.} We identify the applications and application
+ \item \textbf{Application.} We identify the applications and application
      classes well adapted to the workflow systems and are benefitting by them.
@@
 
- \item \textbf{evolution.} We comment on the evolution of the existing workflow
+ \item \textbf{Evolution.} We comment on the evolution of the existing workflow
      systems and the possible current and future development directions which
@@
 the University of California. Kepler workflows follow an actor-director
-metaphor where an actor represents a unit of execution and the director
+metaphor where an actor represents a processor and the director
 represents the paradigm for executing the workflows consisting of the actors.
@@
 computation, cloud computing~\cite{kepler2013}.
-
-%
-\paragraph{Dryad/--}
+\km{Todo: describe Kepler MapReduce in more detail, add Kepler's cloud abilities description}
 
@@
 computing framework that can map abstract workflows as DAGs to the concrete
-underlying resources using intelligent and dynamic scheduling policies. A
+underlying resources using intelligent\km{elaborate} and dynamic\km{elaborate} scheduling policies. A
 Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
@@
 management techniques. Pegasus manager buffers and maintains a fix-sized window
-of executable tasks before accepting new tasks.
-
-\paragraph{Wings/--} Wings is a Workflow system which does software choreography based on AI techniques~\cite{wings2013}
-%
-\paragraph{Swift/Swift Script} Swift \cite{swift} is a script-based scientific
-workflow environment developed at the University of Chicago. A Swift workflow
-specified in the SwiftScript scripting language is translated to an
-intermediate XML-based representation and enacted by the Karajan workflow
-enactment engine. The SwiftScript language is proposed as a workflow language
-founded on a scripting approach programmers are familiar with. To make the
-language adapted to grids, the inventors of SwiftScript generalized the use of
-futures \cite{Halstead85multilisp:a} for every variables defined in the script.
-Futures are non-blocking assignment variables, using a proxy to ensure
+of executable tasks before accepting new tasks. Recent developments in Pegasus
+enables it to run workflows on cloud
+environments~\cite{peg-cloud}.\km{elaborate, does not seem to do GPU
+computing}.
+
+\paragraph{Wings/--} Wings is a Workflow system which does software choreography based on AI techniques~\cite{wings2013}. \km{elaborate}.
+
+%
+\paragraph{Swift/Script, Tcl} Swift \cite{swift} is a script-based scientific
+workflow environment developed at the University of Chicago and Argonne
+National Laboratory. A Swift workflow specified in the Swift scripting language
+and is translated to an intermediate XML-based representation and enacted by
+the Karajan workflow enactment engine. The SwiftScript language is proposed as
+a workflow language founded on a scripting approach that programmers are
+familiar with. To make the language data parallel, Swift script generalizes the
+use of futures \cite{Halstead85multilisp:a} for every variables defined in the
+script.  Futures are non-blocking assignment variables, using a proxy to ensure
 immediate execution of assignment and performing lazy blocking on variable
@@
 asynchronously as long as a data access is not blocking. The availability of
-data enables blocked thread to restart execution. The beauty of futures is to
-make this process completely hidden from the programmer who can use a scripting
-language he is familiar with without ever manipulating explicit parallel
-programming constructs. Swift facilitates a highly asynchronous workflow
-execution. Swift supports advanced workflow constructs such as parallel loops
-and conditionals. Furthermore Swift supports processing of explicitly defined
-arrays of data. Swift workflows take advantage of POSIX-style shared file
-system to reuse the data produced by other processors.
-
+data enables blocked thread to restart execution. Swift facilitates a highly
+asynchronous workflow execution. Swift supports advanced workflow constructs
+such as parallel loops and conditionals. Furthermore Swift supports processing
+of explicitly defined arrays of data. Swift workflows take advantage of
+POSIX-style shared file system to reuse the data produced by other processors.
+Recent advances in Swift:
 %Recent advances in Swift: 
-1. Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via its
-coasters mechanism~\cite{ucc2011}.
-
-2. Swift/T~\cite{turbine2013}, runs scientific workflows over
-HPC resources. Have been shown to work on petascale supercomputers such as the
-IBM BlueGene and Cray series of machines.
-
-3. GeMTC project investigating running many-task applications over GPUs via Swift/T.
+\begin{enumerate}
+    \item Swift has been shown to run on multiple
+        clouds~\cite{ccgrid2013} via its coasters mechanism~\cite{ucc2011}.
+
+    \item Swift/T~\cite{turbine2013}, runs scientific workflows over HPC
+        resources. Have been shown to work on petascale supercomputers such as
+        the IBM BlueGene and Cray series of machines.
+
+    \item GeMTC project investigating running many-task applications over GPUs via Swift/T.
+\end{enumerate}
 
@@
 graphical interface of AGWL is based upon the UML (Unified Modeling Language)
-diagrams.
+diagrams.\km{Todo: current developments and advances.}
 
@@
 condor-Dagman as workflow engine while the recent version uses its own in-house
-developed workflow engine called Xen.
-
-%
-\paragraph{DaGue/--}
+developed workflow engine called Xen. \km{Investigate cloud and GPU capabilities}
 
@@
 (\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided
-by Galaxy using their own Galaxy server or a web-server exposing the webservice
-descriptions through public WSDLs. One of the major use of Galaxy is to achieve
-a mix and match of data analysis tools in order to analyze a vast repository of
-genome based database.
+by Galaxy using their own Galaxy server. One of the major use of Galaxy is to
+achieve a mix and match of data analysis tools in order to analyze a vast
+repository of genome based database.
 
@@
 performance of the workflow engine increases linearly with the increasing
-number of similar tasks.  Here, similar is an important term that means the
-nature of tasks should be the same in terms of processing and data handling.
-Depending upon the lag or increase in the linearity we say a workflow enactor
-scales up or down.  In the present section, we focus on the scalability aspects
-of the core workflow engine. We do not consider the specific scalability issues
-related to a distributed infrastructure such as network latencies and protocol
-offsets. However, there have been efforts to thwart such issues at the workflow
-level by combining workflow enactment techniques with those of distributed
-computing.  One such example is the map/reduce framework \cite{mapreduce}.
-Recently, the map/reduce framework, that aims at performing distributed
-processing on large datasets, has gained a significant popularity in workflow
-community. It is based on the master/worker metaphor of distributed processing.
-It employs the commonly used \textit{map} and \textit{reduce} functions from
-the functional programming languages. The map function chops a large chunk of
-data into smaller pieces and distributes it to \textit{workers} which in turn
-can also chop it to further smaller datasets forming a tree-like hierarchy. The
-worker process the data and return the results back to their parents. The
-reduce step collects the results from the workers and combines them to obtain
-the final result. One of the main reason for the popularity of the map/reduce
-framework for workflows is the recent attention to collections based datasets
-and processing. There exists implementations of the map/reduce framework
-written in different languages with enhancements and support services such as a
-sophisticated query service. Hadoop (\textit{http://hadoop.apache.org/}) is one
-such implementation which has already been interfaced by the Kepler
-SWE~\cite{wang-crawl-etal:2009}.
-
-Large-scale workflows with thousands of tasks often pose memory and disk
+number of similar tasks. Depending upon the lag or increase in the linearity we
+say a workflow enactor scales up or down.  In this section, we focus on
+the scalability aspects of the workflow runtime. We do not consider the
+specific scalability issues related to a distributed infrastructure such as
+network latencies and protocol related overheads. However, there have been efforts to
+thwart such issues at the workflow level by combining workflow enactment
+techniques with those of distributed computing.  One such example is the
+map/reduce framework \cite{mapreduce}.  Recently, the map/reduce framework,
+that aims at performing distributed processing on large datasets, has gained a
+significant popularity in workflow community. It is based on the master/worker
+metaphor of distributed processing.  It employs the commonly used \textit{map}
+and \textit{reduce} functions from the functional programming languages. The
+map function chops a large chunk of data into smaller pieces and distributes it
+to \textit{workers} which in turn can also chop it to further smaller datasets
+forming a tree-like hierarchy. The worker process the data and return the
+results back to their parents. The reduce step collects the results from the
+workers and combines them to obtain the final result. One of the main reason
+for the popularity of the map/reduce framework for workflows is the recent
+attention to collections based datasets and processing. There exists
+implementations of the map/reduce framework written in different languages with
+enhancements and support services such as a sophisticated query service. Hadoop
+(\textit{http://hadoop.apache.org/}) is one such implementation which has
+already been interfaced by the Kepler SWE~\cite{wang-crawl-etal:2009}.
+
+Large-scale applications with thousands of tasks often pose memory and disk
 management difficulties leading to scalability issues.  However, the
-performance of a workflow engine is dependent on several factors including the
+performance of a workflow system is dependent on several factors including the
 type of tasks and approach for submitting tasks for execution.  There is not
@@
 et. al. \cite{callaghan-deelman-etal:2010}.  One way to compare the scalability
-of different workflow environment is to experiment with a similar set of
+of different workflow environments is to experiment with a similar set of
 workflows on a given system. In addition, different SWEs interface to different
 DCIs. This is a time consuming process.  In the current study we do not
-directly compare the workflow engine scalability but we study the similarities
+directly compare the workflow system scalability but we study the similarities
 and differences in the techniques employed by workflow engines in order to
@@
 
-In the context of workflow enactors, scalability can be further classified in
-terms of the computational and data transfer scalability. Computational
-scalability could be expressed as the rate of increase in number of tasks
-executed with respect to the increase in the elapsed time.
+In the context of workflow runtime, scalability can be further classified in
+terms of data motion scalability. It can be expressed as the rate of increase
+in number of data movement quantity with respect to the increase in the elapsed
+time.
 
@@
 
-Most workflow enactors exploit the parallelism at three-levels. First, at the
+Most workflow systems exploit the parallelism at three-levels. First, at the
 task-level wherein the independent branches of a workflow graph are enacted
@@
 In this section we analyze the scalability measures and techniques employed by
-contemporary SWEs.
-
-Illustrated below are some of the techniques used to achieve scalability in
-various SWEs.
+contemporary SWEs. Illustrated below are some of the techniques used to achieve
+scalability in various SWEs.
 
@@
 scheduler level. In Swift-Karajan system this parameter is regulated
-automatically as part of a ``throttling parallelfor" construct. A similar
+automatically as part of a ``throttling parallel foreach" construct. A similar
 configurable parameter is provided by the Knime thread manager in order to
@@
 
-In Swift systems the structure of a workflow is constructed and expanded
-dynamically as opposed to that of Pegasus' static planning. A static workflow
-graph allows optimizations based on the global state of the workflow, whereas a
-dynamic workflow structure allows adaptations to changing environments.
-Depending upon the execution environment and access conditions to the available
-resources, each approach can result in the overall optimization of a given
-application.
+In some systems the structure of a workflow is constructed and expanded
+dynamically as opposed to static planning. A static workflow graph allows
+optimizations based on the global state of the workflow, whereas a dynamic
+workflow structure allows adaptations to changing environments.  Depending upon
+the execution environment and access conditions to the available resources,
+each approach can result in the overall optimization of a given application.
 

### 2014-04-25 paper.tex
minor

@@
 computing framework that can map abstract workflows as DAGs to the concrete
-underlying resources using intelligent\km{elaborate} and dynamic\km{elaborate} scheduling policies. A
+underlying resources using intelligent and dynamic\km{elaborate intelligent and dynamic} scheduling policies. A
 Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
@@
 enables it to run workflows on cloud
-environments~\cite{peg-cloud}.\km{elaborate, does not seem to do GPU
-computing}.
-
-\paragraph{Wings/--} Wings is a Workflow system which does software choreography based on AI techniques~\cite{wings2013}. \km{elaborate}.
+environments~\cite{peg-cloud}.\km{Pegasus does not seem to have GPU computing features}.
+
+\paragraph{Wings/--} Wings is a Workflow system which does software
+choreography based on AI techniques~\cite{wings2013}. \km{ToDo: expand}.
 

### 2014-05-02 paper.tex
new ref for T
update citations

@@
 perceived and visualized. The MA
-DAG~\cite{caron-desprez:2005},\km{could not find any recent paper on MA DAG, interesting concept} engine is one example of a DAG generator engine. MA
+DAG~\cite{caron-desprez:2005},\km{could not find any recent paper on MA DAG} engine is one example of a DAG generator engine. MA
 DAG has evolved from a regular DAG generator to a more abstract DAG generator
@@
 this work. We rely on the following sources for our information: 
-\begin{itemize}
+\begin{enumerate}
     \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
     \item \textbf{Technical Reports.} Technical reports and theses.
-    \item \textbf{Web Presence.} Documentation and other information from the web pages of the host SWE.
+    \item \textbf{Web Presence.} Documentation and other information from the web pages hosting the SWE.
     \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
-\end{itemize}
+\end{enumerate}
 
@@
 \paragraph{Taverna/scufl} The Taverna Workflow Management System
-\cite{oinn-addis-etal:2004} is an open source scientific dataflow manager
+\cite{taverna2013} is an open source scientific dataflow manager
 developed by the mygrid~consortium in the UK.
@@
 %
-\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE
-based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at
+\paragraph{Kepler/MoML} Kepler~\cite{kepler2013} is a SWE
+based upon the Ptollemy II framework~\cite{eker-janneck-etal:2003} developed at
 the University of California. Kepler workflows follow an actor-director
@@
 \paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
-\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
+\cite{fahringer2014} is developed at the University of Innsbruck.
 ASKALON is described as a Scientific PSE capable of composing, hosting,
@@
 
-\paragraph{The P-GRADE Portal/--} The P-GRADE Portal \cite{kacsuk-sipos:2005}
+\paragraph{The P-GRADE Portal/--} The P-GRADE Portal~\cite{pgrade2011}
 developed at the MTA SZTAKI Computation and Automation Institute, Budapest,

### 2014-05-02 paper.tex
update citations

@@
 Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
-uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its
+uses the Modeling Markup Language (MoML) for its
 workflow representation. MoML is an extended XML syntax providing a rich
@@
 
-Business Process Execution Language (BPEL) \cite{oasis:2007} is a workflow
+Business Process Execution Language (BPEL) is a workflow
 language pioneered by IBM Corporation and used predominantly to define business
@@
 BPEL or its extended version in the scientific applications
-\cite{slominski:2007,bpel}. However, the language does not suit well to the
+\cite{bpel}. However, the language does not suit well to the
 data-intensive, grid workflows and has a limited adoption in the scientific

### 2014-05-06 paper.tex
update citations

@@
 \item \textbf{Bulk Work.}
-Workflows are often used to perform computational work in bulk. This could be
-bulk data processing or running multiple distinct processing tasks on a single
-set of data. In addition to carrying out such work, workflow environments
-provide oppurtunities for giving structure and order to a set of tasks and
-often with sufficient expressivity, choices and selections to carry out a bulk
-of work selectively. Some of the examples of such work are \emph{parameter
-sweeps} where tasks are run over a series of parameter values, \emph{ensemble
-studies} where tasks are carried out over discrete groups of parameters over a
-space.
+Workflows are important because they enable to perform computational work in
+bulk. This could be bulk data processing or running multiple distinct
+processing tasks on a single set of data. In addition to carrying out such
+work, workflow environments provide oppurtunities for giving structure and
+order to a set of tasks and often with sufficient expressivity, choices and
+selections to carry out a bulk of work selectively. Some of the examples of
+such work are \emph{parameter sweeps} where tasks are run over a series of
+parameter values, \emph{ensemble studies} where tasks are carried out over
+discrete groups of parameters over a space.
 
@@
 entities.  SWEs are an ideal platform enabling users to perform many tasks
-conveniently.  These tasks include visually formulating a task, composing and
-marshalling of workflow, browsing and selecting data, visualizing and sharing
-results.
+conveniently.  These tasks include visually formulating a task, composing
+and marshalling of workflow, browsing and selecting data, visualizing and
+sharing results.
 
@@
 Directed Acyclic Graph form \cite{dataflow,lee-messerschmitt:1987}. Early
-research on dataflow languages as visual language \cite{sutherland:1966}
+research on dataflow languages as visual language~\cite{sutherland:1966}
 reinforce the notion of the suitability of dataflow paradigm for workflows.

### 2014-05-09 paper.tex
bib fixes

@@
 
-% == ENACTOR SCALABILITY AND OPTIMIZATION ==
-\section{Enactor Scalability and Optimization} \label{sec:scal}
+% == SYSTEM SCALABILITY AND OPTIMIZATION ==
+\section{Scalability and Optimization} \label{sec:scal}
 %We have results comparing \MOTEUR and Taverna performance!
@@
 large number of jobs simultaneously. It is able to dispatch large number of
-concurrent instances of processors. An ability to implement superpipelined
-\cite{hwang:1992} execution of workflow tasks leads to a maximum expression of
-parallelism.
+concurrent instances of processors. An ability to implement
+superpipelined~\cite{hwang:1992} execution of workflow tasks leads to a maximum
+expression of parallelism~\cite{pipeline}.
 
@@
 The Pegasus workflow manager introduced subworkflows because of scalability
-issues at the local host for a large number of nodes in a DAG. A hierarchical
+issues at the localhost for a large number of nodes in a DAG. A hierarchical
 representation helps it scale at the local planning level as well as at the
@@
 criteria. Pegasus performs graph transformation with the knowledge of the whole
-workflow graph.
+workflow graph. A similar strategy is grouping tasks by application-level
+wrappers written in higher-level programming languages such as bash or Python.
 
@@
 \begin{table}[ht!]
+%TODO: Caption is not printing well, fix
+%\caption{A summary of SWEs Scalability and optimization Mechanisms}
 \begin{center}
 % use packages: array
-\begin{tabular}{|l|l|l|l|l|l|l|}
-\hline
-\textbf{SWE} / & \multicolumn{3}{c|}{\textbf{Parallelism}} & \multicolumn{3}{c|}{\textbf{Optimization}}\\
-\cline{2-7}
-\textbf{Language} & \textbf{data} & \textbf{workflow} & \textbf{pipeline} & \textbf{static} & \textbf{runtime} & \textbf{other}\\ \hline
-Taverna/Scufl2 & \checkmark & \checkmark & \checkmark & -- & \checkmark & -- \\\hline
-+/BPEL &foreach & \checkmark & -- &-- &\checkmark &-- \\\hline
-Triana/- & \checkmark & \checkmark & \checkmark & -- & \checkmark &--\\\hline
-MOTEUR/Scufl & \checkmark & \checkmark & \checkmark &-- &\checkmark &-- \\\hline
-Kepler/MoML & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark &--\\\hline
-GridNexus/GXPL & -- & \checkmark & -- &-- &\checkmark &-- \\\hline
-Swift/SwiftScript & \checkmark & \checkmark & \checkmark & \checkmark &\checkmark &-- \\\hline
-Askalon/AGWL &--& \checkmark & -- & \checkmark & \checkmark & -- \\\hline
-P-GRADE/- &\checkmark &\checkmark &\checkmark &--&--&--\\\hline
-Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark &AI-based techniques \\\hline
-Galaxy/- &-- &-- &\checkmark &-- &-- &-- \\\hline
-MOTEUR2/Gwendia~ & \checkmark &\checkmark & \checkmark &-- &\checkmark &-- \\\hline
+\begin{tabular}{|l|l|l|l|l|l|}
+\hline
+\textbf{SWE} / & \multicolumn{3}{c|}{\textbf{Parallelism}} & \multicolumn{2}{c|}{\textbf{Optimization}}\\
+\cline{2-6}
+\textbf{Language} & \textbf{data} & \textbf{task} & \textbf{pipeline} & \textbf{static} & \textbf{runtime}\\
+\hline
+Taverna/Scufl2 & \checkmark & \checkmark & \checkmark & -- & \checkmark \\
+\hline
+Triana/- & \checkmark & \checkmark & \checkmark & -- & \checkmark \\
+\hline
+MOTEUR/Scufl & \checkmark & \checkmark & \checkmark &-- &\checkmark \\
+\hline
+Kepler/MoML & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+\hline
+GridNexus/GXPL & -- & \checkmark & -- &-- &\checkmark \\
+\hline
+Swift/SwiftScript & \checkmark & \checkmark & \checkmark & \checkmark &\checkmark \\
+\hline
+Askalon/AGWL &--& \checkmark & -- & \checkmark & \checkmark \\
+\hline
+P-GRADE/- &\checkmark &\checkmark &\checkmark &--&--\\
+\hline
+Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark \\
+\hline
+Galaxy/- &-- &-- &\checkmark &-- &-- \\
+\hline
+MOTEUR2/Gwendia~ & \checkmark &\checkmark & \checkmark &-- &\checkmark \\
+\hline
 \end{tabular}
-%\caption{A summary of SWEs Scalability and optimization Mechanisms}
 \label{tbl:scalability}

### 2014-05-19 paper.tex
bib fixes
Taverna text

@@
 \end{enumerate}
-
 \section{Objectives and Organization}
@@
      contemporary SWEs. This will help catalog and collect the SWEs by their
-     offerings and identify mechanisms and needs for new and existing features.
-     This will also help identify current trends in scientific computing
-     community.
+     offerings and identify mechanisms and needs for new and existing features
+     in one place.  This will also help identify current trends and approaches
+     to computation in scientific communities.
 
@@
 
-A Taverna workflow consists of a collection of processors connected by data
-links, which establish a dependency between the output(s) of a processor and
-the input(s) of another. Apart from data links, the processors may also be
-connected through coordination links that make a processor invoke when a
-previous processor ends its execution.  The processors are of different kinds
+A Taverna workflow consists of a collection of processors connected by data and
+coordination links, which establish a dependency between the output(s) of a
+processor and the input(s) of another. The processors are of different kinds
 depending on the application code to be invoked: typically being services, with
 input and output \textit{ports} that correspond to the operation's invocations
-and response sequences defined in the service`s interface expressed as WSDL.
+and response sequences defined in the service`s interface expressed as REST or WSDL.
 Java classes and local scripts can be used as workflow processors, as long as
@@
 consistent with the dependencies, and manages the flow of data through the
-processors' ports. The overall workflow execution is data-driven, with a \emph{push}
-model: a processor's execution is started as soon as all of its inputs are
-available.
+processors' ports. The overall workflow execution is data-driven, with a
+\emph{push} model: a processor's execution is started as soon as all of its
+inputs are available.
 
@@
 engine's functionality is facilitated through a Service Provider Interface
-(SPI). A third party developer intending to extend a functionality must write
-its logic as an SPI and publish it as an extension SPI. This may manifest
-itself in different ways depending upon the intended requirements of the
-developer, examples being as a task in the service panel or internally within
-the enactor as additional functionality.
-
-The Taverna engine implements a sophisticated mechanism for enactment of
-parallel workflows. These mechanisms manifest in the form of pipelines and an
-advanced superpipelined enactment of workflows. In the case of pipelined
-execution, a simple pipeline of tasks are overlapped into a back to back
-execution for a single data set. Such a mechanism induces multiple downstream
-workflow processors running in parallel for a dataset. In the case of
-superpipelines, multiple such pipelines instances run for multi-datasets
+(SPI). The Taverna engine implements a sophisticated mechanism for enactment of
+parallel workflows. These mechanisms manifest in the form of complex pipelines
+and an advanced superpipelined enactment of workflows. Such a mechanism induces
+multiple downstream workflow processors running in parallel for a dataset. In
+the case of superpipelines, multiple pipeline instances run for multi-datasets
 providing a high degree of parallelism limited only by the availability of
-underlying hardware resources.
-
-Taverna is not natively interfaced with external computational infrastructure but the same can
-be achieved with the help of extending the engine through one of its SPIs~\cite{maheshwari-missier-etal:2009}. 
-\km{ToDo: Taverna current activity, applications support, GPU, Cloud support}
+underlying resources. Taverna is not natively interfaced with external
+computational infrastructure but the same can be achieved with the help of
+extending the engine through one of its
+SPIs~\cite{maheshwari-missier-etal:2009}. Recent implementations of Taverna
+offers a client-server based environment in which a Taverna server can remotely
+serve the workflow composition and execution services. This mode is useful in
+the cloud computing environments where the server decides where to run the
+Taverna services in the cloud.
 
@@
 familiar with. To make the language data parallel, Swift script generalizes the
-use of futures \cite{Halstead85multilisp:a} for every variables defined in the
+use of futures~\cite{futures} for every variables defined in the
 script.  Futures are non-blocking assignment variables, using a proxy to ensure

### 2014-05-19 paper.tex
Makeflow text

@@
 extending the engine through one of its
-SPIs~\cite{maheshwari-missier-etal:2009}. Recent implementations of Taverna
-offers a client-server based environment in which a Taverna server can remotely
-serve the workflow composition and execution services. This mode is useful in
-the cloud computing environments where the server decides where to run the
-Taverna services in the cloud.
+SPIs~\cite{maheshwari-missier-etal:2009}. In a crucial development, recent
+implementations of Taverna offers a client-server based environment in which a
+Taverna \emph{server} can remotely serve the workflow composition and execution
+services. This mode is useful in the cloud computing environments where the
+server decides where to run the Taverna services in the cloud.
 
@@
 %
-\paragraph{Makeflow} A make-like workflow platform~\cite{makeflow}.
+\paragraph{Makeflow} Makeflow~\cite{makeflow} is a make-like workflow platform.
+Makeflow addresses the often-encountered problem of software and library
+dependencies on workflow tasks execution for heterogeneous environments.
+Makeflow offers techniques to ``link" and package the application execution
+elements such as executables, libraries and configuration parameters.
 

### 2014-05-19 paper.tex
small changes

@@
 % Title portion
-\title{A Survey of Scientific Workflow Management Systems}
+\title{State of the Art in Scientific Workflow Systems}
 \author{KETAN MAHESHWARI
@@
 Makeflow offers techniques to ``link" and package the application execution
-elements such as executables, libraries and configuration parameters.
+elements such as executables, libraries and configuration parameters. This
+improves the durability of a workflow and decreases the efforts required to
+maintain a working version of an application encoded as a workflow.
+
+%Internal representation of a Makeflow workflow is in...
+%Its expressivity features include...
+%Makeflow's runtime features include...
 
@@
 vocabulary to compose Kepler workflows. A workflow processor is called an
-`actor' in Kepler.
-
-Recent advances in Kepler include: MapReduce style computation, bioinformatics
+`actor' in Kepler. Recent advances in Kepler include MapReduce style computation, bioinformatics
 computation, cloud computing~\cite{kepler2013}.
-\km{Todo: describe Kepler MapReduce in more detail, add Kepler's cloud abilities description}
 

### 2014-05-23 paper.tex
calibrate

@@
 \label{sec:intro}
-\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE from here on)
+\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE)
 are, in principle similar to the traditional programming environments
@@
 aspects of workflow paradigm which makes them a strong contenders for the
-purpose of high-level application programmers.
-%
-%Add more content here to beef up the context.
-%
+purpose of high-level application programming. Computation in science is
+dominated by computation for data analysis via one or more mathematical or
+scientific model execution. These models of computation are composed of
+multiple stages where each stage executes a program requiring input parameters
+and data while producing results in the form of data which in turn acts as
+input to the next stage. SWEs can ease these activities by supporting the
+expression, execution and automation.
 % Goals and benefits of a workflow management system
-The goals of a workflow management system for scientific computing are as
-follows:
+The goals and benefits of an SWE are as follows:
 
@@
 bulk. This could be bulk data processing or running multiple distinct
-processing tasks on a single set of data. In addition to carrying out such
+processing tasks on a single set of data. In addition to support such
 work, workflow environments provide oppurtunities for giving structure and
@@
 selections to carry out a bulk of work selectively. Some of the examples of
-such work are \emph{parameter sweeps} where tasks are run over a series of
+this approach are \emph{parameter sweeps} where tasks are run over a series of
 parameter values, \emph{ensemble studies} where tasks are carried out over
@@
 entities. SWEs are an ideal platform enabling users to perform many tasks
-conveniently.  These tasks include visually formulating a task, composing
-and marshalling of workflow, browsing and selecting data, visualizing and
-sharing results.
+conveniently. These tasks include visually formulating an analysis, composing
+and marshalling of processes, browsing and selecting resulting data,
+visualizing and sharing final results.
 
@@
 access modalities change rapidly with upgradation of existing technologies and
-adaptation of new techniques. Clouds are the latest examples of new techniques
-of computational platform access while GPGPUs are examples of upgradation of
+adaptation of new techniques. Clouds are an example of new techniques
+of computational platform access while GPGPUs are example of upgradation of
 existing processing environments. Additionally, with the advent of fast and
-dedicated networks such as ESNet and federated visualization resources, remote
-visualization is gaining prominence among scientific users. SWEs play
-important role as platforms to give transparent and hassle free access to such
-novel and multiple environments to users by hiding the technicalities of
+dedicated networks such as ESNet~\cite{esnet} and federated remote
+visualization resources are gaining prominence among scientific users. SWEs
+play important role as platforms to give transparent and hassle free access to
+such novel and multiple environments to users by hiding the technicalities of
 access methods and usage and configuration related nuances.
@@
 crucial entities in many research fields. They help connecting results from
-a multitude of research aspects. A provenance trace of computation can help
-checkpoint methods and reproduce and modify partial computation. Automated
+a multitude of research points of view. A provenance trace of a computation can
+help checkpoint methods and reproduce and modify partial computation. Automated
 observations of deviations in scientific phenomena from provenance data is one
@@
 
-\item \textbf{Sharing and Collaboration.} Workflow systems are often targetted
-towards a particular community of users. Furthermore, dissemination of methods and results
+\item \textbf{Sharing and Collaboration.} SWEs are often targetted
+towards a particular community of users. Dissemination of methods and results
 to communities and collaborators are some of the activities that SWEs make
-convenient for end-users. For instance, the Galaxy workflow system is targetted
+convenient for end-users. For instance, the
+Galaxy~\cite{blankenberg-vonkuster-etal:2010} workflow system is targetted
 specifically for the bioinformatics users encouraging them to collaborate over
-problems and share results. Another such example is Taverna which is geared
-towards sharing and collaborative workflow oriented scientific computation
-enabling users to discover new services introduced to a community. Sharing and
-collaboration of applications and related data have become ubiquitious with the
-rise of social web. Initiatives such as
+problems and share results. Another such example is Taverna~\cite{taverna2013}
+which is geared towards sharing and collaborative workflow oriented scientific
+computation enabling users to discover new services introduced to a community.
+Sharing and collaboration of applications and related data have become
+ubiquitious with the rise of social web. Initiatives such as
 \emph{myExperiment}~\cite{deroure-goble-etal:2008} and recently
@@
 
-The above described goals also form a basis for the need for a SWE. The need
-for the use of SWE can arise as a result of one or more of such requirements.
-Traditional programming platforms are not suitable to address these
-requirements as they operate at a different level of abstraction. SWEs differ
-in many respects among themselves as well as with respect to traditional
-programming environments. The following list identifies the major differences
+The above described goals and benefits also form a basis for the need for a
+SWE. The need for the use of SWE can arise as a result of one or more of such
+requirements. Traditional programming platforms are not suitable to address
+these requirements as they operate at a different level of abstraction. SWEs
+differ in many respects among themselves as well as with respect to traditional
+programming environments. Consequently, a choice of the suitable SWE must be
+made by a researcher. The following list identifies the major differences
 between traditional and the SWE programming paradigm:
@@
 %
-
 \item \textbf{Scalability and Optimization.} SWEs are more likely to involve
@@
 %
-
 \item \textbf{Data Management.} The scale of data transfer is much higher in
@@
 Owing to such scales, applications often need access to special data movement
-and access protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
+and access and movement protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
 shipped with capabilities to handle such protocols.
@@
 \item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
-%TODO: Add content about new and emerging Distributed Computing
-    %Infrastructures: Clouds, GPUs and exascale supercomputers
-Building SWE's interface to distributed systems is a complex task. The
-distributed nature of resource utilization and asynchronous execution of
-workflow requires a set of interdependent parameters to be configured
-correctly. Modern computing infrastructures such as Clouds, GPUs and
-supercomputers with hybrid nodes pose new challenges of interface development.
-As SWEs involve running many tasks over an infrastructure it becomes crucial to
-interface with these systems such that the tasks are scheduled in an optimal
-manner. Many SWEs implement
+Supporting an interface to a distributed system is a complex task from the
+point of view of an SWE. The distributed nature of resource utilization and
+asynchronous execution of workflow requires a set of interdependent parameters
+to be configured correctly. Modern computing infrastructures such as Clouds,
+GPUs and supercomputers with hybrid nodes (CPU + GPGPU) pose new challenges of
+interface development.  As SWEs involve running many tasks over an
+infrastructure it becomes crucial to interface with these systems such that the
+tasks are scheduled in an optimal manner. Many SWEs implement
 \emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
@@
 are some of the distinctions between a traditional programming and workflow
-platforms.
+platforms in the context of distributed and scalable computation.
 
@@
 flow oriented where data flow is a secondary aspect. In fact, the
-programmer must explicitly make sure that correct the inputs are available
+programmer must explicitly make sure that correct inputs are available
 to instructions and are in the right format. A program proceeds as the
@@
 granularity drives the depth to which a workflow can be expressed in a given
-language form. In the rest of the paper, we refer to a unit of execution of a
+language form. In the rest of the paper, we refer to this unit of execution of a
 workflow system as \emph{processor}. On the other hand, a unit of execution in
 most programming languages is a much simpler and finer executable statement
-that accomplishes a small computational task. This presents a whole new set of
+that accomplishes a small task. This presents a whole new set of
 requirements for expressing the processors and how they should interact with
@@
 non-expert users. For this reason, SWE user interfaces are dominated by the
-visual paradigm of programming. On the other hand, traditional programs are
-most often written by expert programmers. Thus, they have a text-based
+visual paradigm of programming. On the other hand, as the sophistication of
+expression and user expertise increases, the visual environment temds to get
+limited in capabilities it can offer. Thus, most SWEs have a text-based
 composition environment albeit with visually assisting programming environments
-such as IDEs. However, they are seldom to the extent that a non-expert user can
-graphically compose programs. While a graphical interface enhances usability, a
-text-based workflow environment can arguably give better control to an
-experienced user. Consequently, many workflow systems offer a dual interface of
-graphical and text medium to compose workflows.
+such as IDEs. However, these textual environments are seldom to the extent that
+a non-expert user can graphically compose programs. While a graphical interface
+enhances usability, a text-based workflow environment can arguably give better
+control to an experienced user. Consequently, many workflow systems offer a
+dual interface of graphical and text medium to compose workflows.
 
@@
 We concede that the above is not an exhaustive list of differences in features.
-However, it is reasonably inclusive and we believe it covers the breadth of
-aspects we cover in the current work. We do not include other interesting SWE
-features such as application predictability and dynamic restructure of DAGs in
-this analysis as they are out of scope of the objectives of the current paper.
+However, it is reasonably inclusive and we believe it covers the breadth of the
+aspects we discuss in the current work. We do not include other interesting SWE
+features such as application predictability in this analysis as they are out of
+scope of the objectives of the current paper.
 
 \section{Conceptual Overview}
-In this section we present a conceptual overview of the subjects of this study--Workflows and Scientific Workflow Environments.
+In this section we present a conceptual overview of the subjects of this
+study--Workflows and Scientific Workflow Environments.
 
@@
 Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
-Graph (DAG). For the purpose of this text, we define a workflow as a
+Graph (DAG). Execution typically begins with the input data being provided to
+the first task and ends at the results being delivered at the end of the last
+task. Tasks are represented by boxes and data/control flow between the task is
+represented as arrows. Note that in cases where tasks are connected with pure
+control flow, it can be simulated by passing token data between tasks.
+
+For the purpose of this text, we define a workflow as a specification of
 computation with at least two discrete tasks either forming distinct
@@
 parameters. Consequently, a one-off task or running a task for a single copy of
-data is \emph{not} a workflow. For the rest of this paper, we will use the above
-definition of a workflow.
-%
+data is \emph{not} a workflow.
+
+For the rest of this paper, we will use the above definition of a workflow.
+%
+
 \begin{figure}[htb]
@@
 Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
-components and their interaction in a SWE. These components are the focus of
+components and their interaction in an SWE. These components are the focus of
 interest for the present work. A \textit{user interface} interacts with the
-\textit{workflow system} using commands and messages on the inside while with
+\textit{workflow engine} using commands and messages on the inside while with
 user on the outside. The \textit{workflow language} becomes an intermediate
 representation for communication between user interface and the
-\textit{workflow system}. The workflow system acts as a compiler/interpreter
-and a high-level scheduler for the processors and interfaces with the underlying
-\textit{distributed resources}. It also monitors the workflow progress and
-collects the intermediate and final results.
+\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
+and a high-level scheduler for the processors and interfaces with the
+underlying \textit{distributed resources}. It resolves data dependencies and
+optimizes the flow at runtime and monitors the workflow progress and collects
+the intermediate and final results.
 
@@
 SWEs are used as tools to benefit from the underlying large-scale computing
-infrastructure such as clusters, supercomputers, clouds, grids. Such
+infrastructure such as clusters, supercomputers, clouds, and grids. Such
 environments are in use today for production level work in different scientific
@@
 of a multi-staged scientific computation. However, they differ in the features
-they provide based upon the scientific domain they offer service for, and the
-computational model they follow. Often, one of the main reasons for such
-difference is the specialized components and requirements for a particular
-domain. Owing to this fact, a given workflow system dominates a particular area
-of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora
-of SWEs suitable for general purpose workflow style computations.
-
-\section{Related Work}
-Surveys and studies on workflows similar to the present one have been done in
-the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
-SWEs, studying many existing systems for distributed computing. This taxonomy
-is based, to a large extent on the capabilities of the workflow enactors. It
-provides only a limited insight to the SWE aspects considered in the present
-paper, especially the workflow languages expressiveness and SWEs interactions
-with the distributed computing infrastructures. A more recent work by Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely
-related to the workflow language expressiveness: it compares the models of
-computations of various workflow directors implemented in Kepler
-\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
-composition of different models in sub-workflows of a global orchestration.
-This work illustrates how different enactors designed for different application
-needs have different expressiveness power. Deelman et. al.
-\cite{deelman-overview} present a high-level overview of the scientific
-workflow features and capabilities including SWE composition, representation
-and resource mapping capabilities. The recent survey from Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance takes into
-account the latencies involved during the enactment mechanisms. A small survey
-and analysis of file-access characteristics for data-intensive workflows has
-been done by Shibata~\cite{Shibata}. Early surveys on workflow
-languages and suitability of general purpose languages to workflows is
-conducted in \cite{generalpurpose}. These surveys were conducted before a few
-years and in the meantime SWEs have come a long way in terms of capabilities
-and enactment approaches. New techniques such as the map/reduce framework
-\cite{mapreduce} have been introduced as a promising new approach to workflow
-enactment over distributed infrastructures. Several issues put forth by such
-surveys such as customized service wrapping and enactment have been adequately
-addressed over the
-years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.
-
-Another well-studied feature is the workflow languages. Many different
-languages have been considered within the SWE community, from raw Directed
-Acyclic Graphs (DAGs) of computational processes such as CONDOR DAGMan
-\cite{dagman} to abstractions for parallel computations such as Petri nets
-\cite{alt-hoheisel:2005},
-data-driven languages such as Scufl \cite{turi-missier-etal:2007} and scripting
-languages such as Swift script \cite{swift}. Glatard's PhD thesis
-\cite{glatard:2007} presents a taxonomy of SWE languages classifying them into
-functional, graph, service-oriented and executable representations. A detailed
-classification and taxonomy of parallelism patterns in Grid Workflows has been
-presented by Pautasso and Alonso in \cite{pautasso1}. This classification is
-more generic while the one presented in Glatard's thesis has emphasis on
-service based workflows.  Each of these approaches can be defended through some
-aspects well covered in their design: DAGs are convenient for scheduling
-\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
-properties such as potential deadlocks, data-driven languages ease the
-description of application logic for non-expert users and scripting is
-extensively used by programmers for prototyping, etc.
+they provide based upon the scientific domain or user community they offer
+service for, and the computational model they follow. Often, one of the main
+reasons for such difference is the specialized components and requirements for
+a particular domain. Owing to this fact, a given workflow system dominates a
+particular area of science (eg. astronomy, biology, etc.). Nonetheless, there
+exist a plethora of SWEs suitable for general purpose workflow style
+computations.
 
 DAGs are by far, the most popular abstraction of the way workflows are
-perceived and visualized. The MA
-DAG~\cite{caron-desprez:2005},\km{could not find any recent paper on MA DAG} engine is one example of a DAG generator engine. MA
-DAG has evolved from a regular DAG generator to a more abstract DAG generator
-laced with control structures. MA DAG benefits from the workflow scheduling
-mechanisms included in the middleware. Due to the impossibility to produce
-complete DAGs owing to the presence of control structures such as conditionals
-and loops, MA DAG generate partial DAGs greedily: a sub-DAG, as complete as
-possible, is produced as soon as possible. Despite the presence of many
-different approaches to workflow systems, new requirements
-continue to pose workflow challenges. Some of the current challenges facing SWEs are:
+perceived and visualized. The MA DAG~\cite{caron-desprez:2005}, engine is one
+example of a DAG generator engine. MA DAG has evolved from a regular DAG
+generator to a more abstract DAG generator laced with control structures. MA
+DAG benefits from the workflow scheduling mechanisms included in the
+middleware. Due to the impossibility to produce complete DAGs owing to the
+presence of control structures such as conditionals and loops, MA DAG generate
+partial DAGs greedily: a sub-DAG, as complete as possible, is produced as soon
+as possible. Despite the presence of many different approaches to workflow
+systems, new requirements continue to pose workflow challenges. Some of the
+current challenges facing SWEs are:
+
 \begin{enumerate}
@@
 \end{enumerate}
+
+\section{Related Work}
+Surveys and studies on workflows similar to the present one have been done in
+the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
+SWEs in the previous decade, studying many existing systems for distributed
+computing. This taxonomy presented is based, to a large extent on the
+capabilities of the workflow engines. It does provide insights to many of the
+SWE aspects considered in the present paper, especially the workflow languages
+expressiveness and SWEs interactions with the distributed computing
+infrastructures in the early days of the compute Grids. A more recent work by
+Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely related to
+the workflow language expressiveness: it compares the models of computations of
+various workflow directors implemented in Kepler
+\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
+composition of different models in sub-workflows of a global orchestration.
+This work illustrates how different enactors designed for different application
+needs have different expressiveness power. Deelman et. al.
+\cite{deelman-overview} present a high-level overview of the scientific
+workflow features and capabilities including SWE composition, representation
+and resource mapping capabilities. A recent survey from
+Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance
+takes into account the latencies involved during the enactment mechanisms. A
+small survey and analysis of file-access characteristics for data-intensive
+workflows has been done by Shibata~\cite{Shibata}. Early surveys on workflow
+languages and suitability of general purpose languages to workflows is
+conducted in~\cite{generalpurpose}. These surveys were conducted before a few
+years and in the meantime SWEs have come a long way in terms of capabilities
+and enactment approaches. New techniques such as the map/reduce framework
+\cite{mapreduce} have been introduced as a promising new approach to workflow
+enactment over distributed infrastructures. Several issues put forth by such
+surveys such as customized service wrapping and enactment have been adequately
+addressed over the
+years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.
+
+Another well-studied feature is the workflow languages. Many different
+languages have been considered within the SWE community, from raw Directed
+Acyclic Graphs (DAGs) of computational processes such as CONDOR
+DAGMan~\cite{dagman} to abstractions for parallel computations such as Petri
+nets~\cite{alt-hoheisel:2005}, data-driven languages such as
+Scufl~\cite{turi-missier-etal:2007} and scripting languages such as Swift
+script~ \cite{swift}. Glatard's PhD thesis~\cite{glatard:2007} presents a
+taxonomy of SWE languages classifying them into functional, graph,
+service-oriented and executable representations. A detailed classification and
+taxonomy of parallelism patterns in Grid Workflows has been presented by
+Pautasso and Alonso in~\cite{pautasso1}. This classification is more generic
+while the one presented in Glatard's thesis has emphasis on service based
+workflows.  Each of these approaches can be defended through some aspects well
+covered in their design: DAGs are convenient for scheduling
+\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
+properties such as potential deadlocks, data-driven languages ease the
+description of application logic for non-expert users and scripting is
+extensively used by programmers for prototyping, etc.
+
 \section{Objectives and Organization}
@@
 \begin{enumerate}
-  \item \textbf{Educate.} With more and more science being done with the help of
-      computational methods, SWEs can add significant value to the process.
-      They can accelerate the process, enable community driven science, drive
-      the development at a manageable level of abstraction for users.
-      However, scientific users are often not experienced programmers or
-      compute savvy. This work will serve as a useful starting guide to
-      educate science users of the available tools and techniques with
-      features that might be suitable for their methods.
+\item \textbf{Educate.} With more science being done with the help of
+computational methods, SWEs can add significant value to the process.  They can
+accelerate the process, enable community driven science, drive the development
+at a manageable level of abstraction for users.  However, scientific users are
+often not experienced programmers or computation savvy. This work will serve as
+a useful starting guide to educate science users and students of the available
+tools and techniques with features suitable for their methods.
 
 \item \textbf{Catalog.} We study a number of important features as offered by
-     contemporary SWEs. This will help catalog and collect the SWEs by their
+the contemporary SWEs. This will help catalog and collect the SWEs by their
 offerings and identify mechanisms and needs for new and existing features
@@
 We hope that readers can make a choice of which SWE to use if there is an
-     adaptation scope.
+adaptation similar in scope to the ones describe in this paper.
 
@@
 SWEs.
-
 \end{enumerate}
@@
 \ref{sec:sum} presents a summary and a classification of all the aspects of
-SWEs studied in this work. Throughout the paper, we also present tables
+SWEs studied in this work.
+
+Throughout the paper, we also present tables
 summarizing the aspects of SWEs considered in respective sections. It was not
@@
 
-Some features, such as extendability of SWE strongly depend on the extension
+Some features, such as extendability of an SWE strongly depend on the extension
 points designed in the code; consequently, we consider it present in only those
@@
 denoted by an asterisk (*) or a hyphen (-) at the respective places.
+
 \paragraph{Taverna/scufl} The Taverna Workflow Management System
@@
 \emph{push} model: a processor's execution is started as soon as all of its
-inputs are available.
-
-Taverna is a flexible and extendible workflow environment
+inputs are available. Taverna is a flexible and extendible workflow environment
 \cite{missier-soiland-reyes-etal:2010,sroka2009a} in that extending the
@@
 
-\paragraph{MOTEUR/scufl} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
+\paragraph{MOTEUR/scufl, GWENDIA} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
 developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
@@
 %
-\paragraph{Makeflow} Makeflow~\cite{makeflow} is a make-like workflow platform.
+\paragraph{Makeflow/make-like spec} Makeflow~\cite{makeflow} is a make-like workflow platform.
 Makeflow addresses the often-encountered problem of software and library
@@
 %Makeflow's runtime features include...
-
-%
-\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based
+%
+\paragraph{Triana/XML} Triana \cite{taylor-shields-etal:2003} is a java-based
 SWE developed at the Cardiff University. It consists of a workflow composition
@@
 Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
-uses the Modeling Markup Language (MoML) for its
-workflow representation. MoML is an extended XML syntax providing a rich
-vocabulary to compose Kepler workflows. A workflow processor is called an
-`actor' in Kepler. Recent advances in Kepler include MapReduce style computation, bioinformatics
-computation, cloud computing~\cite{kepler2013}.
-
-%
-\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed
-computing framework that can map abstract workflows as DAGs to the concrete
-underlying resources using intelligent and dynamic\km{elaborate intelligent and dynamic} scheduling policies. A
-Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
-XML). DAX describes the logical workflow components, their inputs and outputs
-and their links and dependencies in a Pegasus workflow. These kind of workflows
-has been shown to enact optimally using Pegasus' smart memory and disk
-management techniques. Pegasus manager buffers and maintains a fix-sized window
-of executable tasks before accepting new tasks. Recent developments in Pegasus
-enables it to run workflows on cloud
-environments~\cite{peg-cloud}.\km{Pegasus does not seem to have GPU computing features}.
-
-\paragraph{Wings/--} Wings is a Workflow system which does software
-choreography based on AI techniques~\cite{wings2013}. \km{ToDo: expand}.
+uses the Modeling Markup Language (MoML) for its workflow representation. MoML
+is an extended XML syntax providing a rich vocabulary to compose Kepler
+workflows. A workflow processor is called an `actor' in Kepler. Recent advances
+in Kepler include MapReduce style computation, bioinformatics computation,
+cloud computing~\cite{kepler2013}.
+
+%
+\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE that is well-suited
+to generate and execute large abstract DAGs made up of tasks with dataflow.
+Abstract description is in the form of a DAG via a language called DAX. The
+concrete expression takes into account the properties of application task
+requirements and underlying compute resources. A Pegasus workflow is expressed
+in a simple XML-based language called DAX (DAG in XML). DAX describes the
+logical workflow components, their inputs and outputs and their links and
+dependencies in a Pegasus workflow. These kind of workflows has been shown to
+enact optimally using Pegasus' smart memory and disk management techniques.
+Large scale workflows from astronomy, biology, and high-energy physics have
+been shown to run via Pegasus on distributed computing resources such as
+clusters, Grids and supercomputers. Pegasus manager buffers and maintains a
+fix-sized window of executable tasks before accepting new tasks.  Regarding
+massively parallel task scheduling, Pegasus provides a task
+clustering technique that groups parallel jobs based on fixed parameters such
+as number of groups. Recent developments in Pegasus enables it to run workflows
+on cloud environments~\cite{peg-cloud}.
+
+\paragraph{Wings/--} The Workflow instance generation and specialization
+(Wings) provides automatic workflow validation and generation based on
+high-level workflow templates~\cite{wings2013}. The system receives high-level
+workflow specification in the form of Pegasus/DAX workflows and generates
+concrete executable workflows. It generates workflows by selecting components,
+data sets, and parameters in three stages while eliminating candidates that are
+not viable.
+
+% How does Wings achieve the translation between abstract and concrete workflows?
+% What are the internal representations of workflows?
+% What are the constraints under which Wings works?
+% How is Wings implemented?
 

### 2014-05-27 paper.tex
pegasus desc

@@
 %
-\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE that is well-suited
-to generate and execute large abstract DAGs made up of tasks with dataflow.
-Abstract description is in the form of a DAG via a language called DAX. The
-concrete expression takes into account the properties of application task
-requirements and underlying compute resources. A Pegasus workflow is expressed
-in a simple XML-based language called DAX (DAG in XML). DAX describes the
-logical workflow components, their inputs and outputs and their links and
-dependencies in a Pegasus workflow. These kind of workflows has been shown to
-enact optimally using Pegasus' smart memory and disk management techniques.
+\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE that is
+well-suited to generate and execute large workflows from abstract DAGs
+describing dataflow.  Abstract description of a DAG is done via an XML-based
+language called DAX. The concrete expression takes into account the properties
+of application task requirements and underlying distributed compute resources.
+The main advantage of such approach is that a single abstract workflow
+description can be mapped to multiple execution environments based on
+requirements. DAX describes the logical workflow components, their inputs and
+outputs and their links and dependencies in a Pegasus workflow. These kind of
+workflows have been shown to enact optimally using Pegasus' dynamic task,
+memory and disk management techniques. For instance, for massively parallel
+task scheduling, Pegasus provides a task clustering technique that groups
+parallel jobs based on fixed parameters such as number of groups. Additionally,
+a manager buffers and maintains a fix-sized window of executable tasks before
+accepting new tasks.
+
 Large scale workflows from astronomy, biology, and high-energy physics have
 been shown to run via Pegasus on distributed computing resources such as
-clusters, Grids and supercomputers. Pegasus manager buffers and maintains a
-fix-sized window of executable tasks before accepting new tasks.  Regarding
-massively parallel task scheduling, Pegasus provides a task
-clustering technique that groups parallel jobs based on fixed parameters such
-as number of groups. Recent developments in Pegasus enables it to run workflows
-on cloud environments~\cite{peg-cloud}.
+clusters, Grids and supercomputers.  Recent developments in Pegasus enables it
+to run workflows on cloud environments~\cite{peg-cloud}. Pegasus system is
+known to be extendible as illustrated by a recent integration
+effort~\cite{peg-hubzero} between Pegasus and the Hubzero
+portal~\cite{hubzero}.
 

### 2014-05-27 paper.tex
pegasus desc

@@
 been shown to run via Pegasus on distributed computing resources such as
-clusters, Grids and supercomputers.  Recent developments in Pegasus enables it
+Clusters, Grids and Supercomputers.  Recent developments in Pegasus enables it
 to run workflows on cloud environments~\cite{peg-cloud}. Pegasus system is

### 2014-05-28 paper.tex
wings desc

@@
 %
-\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE that is
+\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE
 well-suited to generate and execute large workflows from abstract DAGs
-describing dataflow.  Abstract description of a DAG is done via an XML-based
-language called DAX. The concrete expression takes into account the properties
-of application task requirements and underlying distributed compute resources.
-The main advantage of such approach is that a single abstract workflow
-description can be mapped to multiple execution environments based on
-requirements. DAX describes the logical workflow components, their inputs and
-outputs and their links and dependencies in a Pegasus workflow. These kind of
-workflows have been shown to enact optimally using Pegasus' dynamic task,
-memory and disk management techniques. For instance, for massively parallel
-task scheduling, Pegasus provides a task clustering technique that groups
-parallel jobs based on fixed parameters such as number of groups. Additionally,
-a manager buffers and maintains a fix-sized window of executable tasks before
-accepting new tasks.
+describing dataflow among interdependent tasks. Abstract description of a DAG
+is done via an XML-based language called DAX. The concrete expression takes
+into account the properties of application task requirements and underlying
+distributed compute resources.  The main advantage of such approach is that a
+single abstract workflow description can be mapped to multiple execution
+environments based on requirements. DAX describes the logical workflow
+components, their inputs and outputs and their links and dependencies in a
+Pegasus workflow. These kind of workflows have been shown to enact optimally
+using Pegasus' dynamic task, memory and disk management techniques. For
+instance, for massively parallel task scheduling, Pegasus provides a task
+clustering technique that groups parallel jobs based on fixed parameters such
+as number of groups. Additionally, a manager buffers and maintains a fix-sized
+window of executable tasks before accepting new tasks.
 
@@
 high-level workflow templates~\cite{wings2013}. The system receives high-level
-workflow specification in the form of Pegasus/DAX workflows and generates
-concrete executable workflows. It generates workflows by selecting components,
-data sets, and parameters in three stages while eliminating candidates that are
-not viable.
+workflow specification in the form of Pegasus/DAX workflows and validates them
+for correct coordination based on data and resource constraints. It generates
+workflows by selecting components, data sets, and parameters in three stages
+while eliminating candidates that are not viable. This approach is useful in
+the context of desiging complex computational experiments by domain scientists
+in that it eases the process of workflow level debugging.
 

### 2014-05-28 paper.tex
wings desc

@@
 the context of desiging complex computational experiments by domain scientists
-in that it eases the process of workflow level debugging.
+in that it eases the process of workflow level debugging. Wings uses W3C's OWL,
+Resource Description Framework (RDF), and Semantic Web Rule Language (SWRL) to
+represent workflows and their associated constraints.
 
 % How does Wings achieve the translation between abstract and concrete workflows?
-% What are the internal representations of workflows?
 % What are the constraints under which Wings works?
@@
 %
-\paragraph{Swift/Script, Tcl} Swift \cite{swift} is a script-based scientific
+\paragraph{Swift/Script, Tcl} Swift~\cite{swift} is a script-based scientific
 workflow environment developed at the University of Chicago and Argonne

### 2014-05-30 paper.tex
small fix

@@
 POSIX-style shared file system to reuse the data produced by other processors.
-Recent advances in Swift:
-%Recent advances in Swift: 
-\begin{enumerate}
-    \item Swift has been shown to run on multiple
-        clouds~\cite{ccgrid2013} via its coasters mechanism~\cite{ucc2011}.
-
-    \item Swift/T~\cite{turbine2013}, runs scientific workflows over HPC
-        resources. Have been shown to work on petascale supercomputers such as
-        the IBM BlueGene and Cray series of machines.
-
-    \item GeMTC project investigating running many-task applications over GPUs via Swift/T.
-\end{enumerate}
+Recently, Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via
+its coasters mechanism~\cite{ucc2011}. Swift/T~\cite{turbine2013}, a new HPC
+implementation of Swift system, runs scientific workflows over large-scale
+supercomputing resources. It has been shown to work on petascale supercomputers
+such as the IBM BlueGene and Cray series of machines. GeMTC project
+investigating running many-task applications over GPUs via Swift/T.
 
@@
 % and summary.
-%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language
+%The \textit{gscript}~\cite{maheshwari-montagnat:2010} workflow language
 %is similar in spirit to the

### 2014-05-30 paper.tex
galaxy desc

@@
 developed workflow engine called Xen. \km{Investigate cloud and GPU capabilities} 
-
 %
@@
 \paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
-bioinformatics domain SWE based framework developed at the Penn State
-University, Pennsilvania. It provides a webportal based environment to build
-complex pipelines from an available repository of tools
-(\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided
-by Galaxy using their own Galaxy server. One of the major use of Galaxy is to
-achieve a mix and match of data analysis tools in order to analyze a vast
-repository of genome based database.
+SWE developed at the Penn State University, Pennsilvania. It provides a
+webportal based environment to build complex pipelines from an available
+repository of tools (\textit{http://g2.bx.psu.edu}).  Galaxy is currently used
+mainly by bioinformatics community. Users can extend the tool repository
+provided by Galaxy using their own Galaxy server. One of the major use of
+Galaxy is to achieve a mix and match of data analysis tools in order to analyze
+a vast repository of genome based database. Galaxy offers a web-based,
+interactive platform for data analysis via reusable and reproducible workflows.
+While excelling in the usability aspects, the Galaxy environment has a limited
+parallelism capabilities and a small number of existing implementations for
+interfacing with diverse, scalable computing infrastructures. The user can draw
+workflows in which multiple copies of a task are created to operate on multiple
+inputs at once, but this is not a scalable approach.
 

### 2014-05-30 paper.tex
provenance desc

@@
 a multitude of research points of view. A provenance trace of a computation can
-help checkpoint methods and reproduce and modify partial computation. Automated
-observations of deviations in scientific phenomena from provenance data is one
-of the key triggers for new scientific discoveries. Fields of studies involving
-genomic and protiens are examples of domains where provenance and metadata
-plays key roles. SWEs can be integrated or extended with metadata and
-provenance techniques rendering powerful computing and traceability features.
+help checkpoint methods and reproduce and modify computation fully or
+partially. Automated observations of deviations in scientific phenomena from
+provenance data is one of the key triggers for new scientific discoveries.
+Fields of studies involving genomic and protiens are examples of domains where
+provenance and metadata plays key roles. SWEs can be integrated or extended
+with metadata and provenance techniques rendering powerful computing and
+traceability features. In fact, most SWEs of today have separate modules or
+extensions for handling provenance and/or metadata.
 

### 2014-05-30 paper.tex
bib add

@@
 traceability features. In fact, most SWEs of today have separate modules or
-extensions for handling provenance and/or metadata.
+extensions for provenance and metadata management.
 
@@
 such as the IBM BlueGene and Cray series of machines. GeMTC project
-investigating running many-task applications over GPUs via Swift/T.
+investigating running many-task applications over GPUs via Swift/T. Efforts to develop metadata and provenance modules have been undertaken recently~\cite{swiftprov}.
 

### 2014-06-01 paper.tex
separate desc

@@
 
-\paragraph{Taverna/scufl} The Taverna Workflow Management System
-\cite{taverna2013} is an open source scientific dataflow manager
-developed by the mygrid~consortium in the UK.
-%
-Taverna includes a GUI-based rich-client workbench for workflow design and a
-workflow enactment runtime. Taverna can enact workflows expressed in the scufl
-language. Scufl is a near Turing complete dataflow language with well-defined
-semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.
-
-A Taverna workflow consists of a collection of processors connected by data and
-coordination links, which establish a dependency between the output(s) of a
-processor and the input(s) of another. The processors are of different kinds
-depending on the application code to be invoked: typically being services, with
-input and output \textit{ports} that correspond to the operation's invocations
-and response sequences defined in the service`s interface expressed as REST or WSDL.
-Java classes and local scripts can be used as workflow processors, as long as
-they expose their signature as input and output ports as required by the model.
-The engine orchestrates the execution of the processors in a way that is
-consistent with the dependencies, and manages the flow of data through the
-processors' ports. The overall workflow execution is data-driven, with a
-\emph{push} model: a processor's execution is started as soon as all of its
-inputs are available. Taverna is a flexible and extendible workflow environment
-\cite{missier-soiland-reyes-etal:2010,sroka2009a} in that extending the
-engine's functionality is facilitated through a Service Provider Interface
-(SPI). The Taverna engine implements a sophisticated mechanism for enactment of
-parallel workflows. These mechanisms manifest in the form of complex pipelines
-and an advanced superpipelined enactment of workflows. Such a mechanism induces
-multiple downstream workflow processors running in parallel for a dataset. In
-the case of superpipelines, multiple pipeline instances run for multi-datasets
-providing a high degree of parallelism limited only by the availability of
-underlying resources. Taverna is not natively interfaced with external
-computational infrastructure but the same can be achieved with the help of
-extending the engine through one of its
-SPIs~\cite{maheshwari-missier-etal:2009}. In a crucial development, recent
-implementations of Taverna offers a client-server based environment in which a
-Taverna \emph{server} can remotely serve the workflow composition and execution
-services. This mode is useful in the cloud computing environments where the
-server decides where to run the Taverna services in the cloud.
-
-\paragraph{MOTEUR/scufl, GWENDIA} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
-developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
-MOTEUR facilitates executing workflows consisting of remote services. It
-exploits service parallelism at workflow level and data parallelism for
-multiple datasets.
-
-MOTEUR has sophisticated language-level expressions for data management
-mechanisms within a workflow definition. It provides a diverse set of
-parallel iterators that are well adapted for different combinations of input datasets
-at the processor. An iteration strategy allows different combinations of the
-data arriving at a processor along its multiple inputs. Running legacy
-application code on distributed systems has been eased by MOTEUR by providing a
-generic grid-service wrapper for these applications
-\cite{glatard-montagnat-etal:2008a}. Services thus created can be readily
-deployed by the MOTEUR runtime. MOTEUR workflows are written in the scufl
-language and are compatible with those written for the Taverna SWE. It also
-supports an extension of scufl language called the Gwendia~\cite{gwendia}
-language, developed at the same laboratory. 
-%
-\paragraph{Makeflow/make-like spec} Makeflow~\cite{makeflow} is a make-like workflow platform.
-Makeflow addresses the often-encountered problem of software and library
-dependencies on workflow tasks execution for heterogeneous environments.
-Makeflow offers techniques to ``link" and package the application execution
-elements such as executables, libraries and configuration parameters. This
-improves the durability of a workflow and decreases the efforts required to
-maintain a working version of an application encoded as a workflow.
-
-%Internal representation of a Makeflow workflow is in...
-%Its expressivity features include...
-%Makeflow's runtime features include...
-%
-\paragraph{Triana/XML} Triana \cite{taylor-shields-etal:2003} is a java-based
-SWE developed at the Cardiff University. It consists of a workflow composition
-GUI, a framework to integrate workflow with the different execution
-environments and middlewares and a set of predefined services written as java
-components. A Triana workflow is expressed as an XML grammar similar to a WSDL
-description for webservices \cite{taylor-wand-etal:2005}. A highlight of
-Triana is that it provides ready-to-use interface for a wide variety of
-middleware and distributed infrastructures including P2P, grids and
-web-services frameworks. A Triana workflow provides dataflow as well as control
-flow support. Triana allows conditional branching and looping in its workflows.
-
-%
-\paragraph{Kepler/MoML} Kepler~\cite{kepler2013} is a SWE
-based upon the Ptollemy II framework~\cite{eker-janneck-etal:2003} developed at
-the University of California. Kepler workflows follow an actor-director
-metaphor where an actor represents a processor and the director
-represents the paradigm for executing the workflows consisting of the actors.
-Depending upon the types of directors, a Kepler workflow may exhibit a
-different type of execution behavior. A workflow designer can chose from among
-directors based upon the purpose of a workflow. Few of these include Process
-Network (PN) for a parallel execution at the workflow level, Synchronous Data
-Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
-Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
-uses the Modeling Markup Language (MoML) for its workflow representation. MoML
-is an extended XML syntax providing a rich vocabulary to compose Kepler
-workflows. A workflow processor is called an `actor' in Kepler. Recent advances
-in Kepler include MapReduce style computation, bioinformatics computation,
-cloud computing~\cite{kepler2013}.
-
-%
-\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE
-well-suited to generate and execute large workflows from abstract DAGs
-describing dataflow among interdependent tasks. Abstract description of a DAG
-is done via an XML-based language called DAX. The concrete expression takes
-into account the properties of application task requirements and underlying
-distributed compute resources.  The main advantage of such approach is that a
-single abstract workflow description can be mapped to multiple execution
-environments based on requirements. DAX describes the logical workflow
-components, their inputs and outputs and their links and dependencies in a
-Pegasus workflow. These kind of workflows have been shown to enact optimally
-using Pegasus' dynamic task, memory and disk management techniques. For
-instance, for massively parallel task scheduling, Pegasus provides a task
-clustering technique that groups parallel jobs based on fixed parameters such
-as number of groups. Additionally, a manager buffers and maintains a fix-sized
-window of executable tasks before accepting new tasks.
-
-Large scale workflows from astronomy, biology, and high-energy physics have
-been shown to run via Pegasus on distributed computing resources such as
-Clusters, Grids and Supercomputers.  Recent developments in Pegasus enables it
-to run workflows on cloud environments~\cite{peg-cloud}. Pegasus system is
-known to be extendible as illustrated by a recent integration
-effort~\cite{peg-hubzero} between Pegasus and the Hubzero
-portal~\cite{hubzero}.
-
-\paragraph{Wings/--} The Workflow instance generation and specialization
-(Wings) provides automatic workflow validation and generation based on
-high-level workflow templates~\cite{wings2013}. The system receives high-level
-workflow specification in the form of Pegasus/DAX workflows and validates them
-for correct coordination based on data and resource constraints. It generates
-workflows by selecting components, data sets, and parameters in three stages
-while eliminating candidates that are not viable. This approach is useful in
-the context of desiging complex computational experiments by domain scientists
-in that it eases the process of workflow level debugging. Wings uses W3C's OWL,
-Resource Description Framework (RDF), and Semantic Web Rule Language (SWRL) to
-represent workflows and their associated constraints.
-
-% How does Wings achieve the translation between abstract and concrete workflows?
-% What are the constraints under which Wings works?
-% How is Wings implemented?
-
-%
-\paragraph{Swift/Script, Tcl} Swift~\cite{swift} is a script-based scientific
-workflow environment developed at the University of Chicago and Argonne
-National Laboratory. A Swift workflow specified in the Swift scripting language
-and is translated to an intermediate XML-based representation and enacted by
-the Karajan workflow enactment engine. The SwiftScript language is proposed as
-a workflow language founded on a scripting approach that programmers are
-familiar with. To make the language data parallel, Swift script generalizes the
-use of futures~\cite{futures} for every variables defined in the
-script.  Futures are non-blocking assignment variables, using a proxy to ensure
-immediate execution of assignment and performing lazy blocking on variable
-value accesses only. The systematic use of futures for every variables in the
-language makes it completely data-driven: the execution progresses
-asynchronously as long as a data access is not blocking. The availability of
-data enables blocked thread to restart execution. Swift facilitates a highly
-asynchronous workflow execution. Swift supports advanced workflow constructs
-such as parallel loops and conditionals. Furthermore Swift supports processing
-of explicitly defined arrays of data. Swift workflows take advantage of
-POSIX-style shared file system to reuse the data produced by other processors.
-Recently, Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via
-its coasters mechanism~\cite{ucc2011}. Swift/T~\cite{turbine2013}, a new HPC
-implementation of Swift system, runs scientific workflows over large-scale
-supercomputing resources. It has been shown to work on petascale supercomputers
-such as the IBM BlueGene and Cray series of machines. GeMTC project
-investigating running many-task applications over GPUs via Swift/T. Efforts to develop metadata and provenance modules have been undertaken recently~\cite{swiftprov}.
-
+\input{text/taverna}
+\input{text/moteur}
+\input{text/makeflow}
+\input{text/triana}
+\input{text/kepler}
+\input{text/peg}
+\input{text/wings}
+\input{text/swift}
+\input{text/askalon}
+\input{text/pgrade}
+\input{text/galaxy}
 % Put this somewhere else, either in the analysis or may be in the conclusions
@@
 %
-
-\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
-\cite{fahringer2014} is developed at the University of Innsbruck.
-ASKALON is described as a Scientific PSE capable of composing, hosting,
-executing, sharing and reusing scientific workflows written in the Abstract
-Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
-capable of expressing scientific application components chains at a high-level
-of abstraction. A concrete workflow consists of the abstract workflow combined
-with the resources and target execution environment binding the workflow. The
-graphical interface of AGWL is based upon the UML (Unified Modeling Language)
-diagrams.\km{Todo: current developments and advances.}
-
-\paragraph{The P-GRADE Portal/--} The P-GRADE Portal~\cite{pgrade2011}
-developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
-provides an access to grid computational and storage resources from a web
-browser through a specialized workflow oriented grid portal. A workflow
-development environment interfaces the users through this web based portal. The
-portal interfaces the multiple grid computing middleware and storage resources
-as a backend component of the environment. A highlight of P-GRADE portal is
-that it provides access to multiple grids and scientific communities in a
-single environment leading to a highly collaborative inter-community,
-inter-grid research environment. An older version of P-GRADE used the
-condor-Dagman as workflow engine while the recent version uses its own in-house
-developed workflow engine called Xen. \km{Investigate cloud and GPU capabilities} 
-%
-
-\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
-SWE developed at the Penn State University, Pennsilvania. It provides a
-webportal based environment to build complex pipelines from an available
-repository of tools (\textit{http://g2.bx.psu.edu}).  Galaxy is currently used
-mainly by bioinformatics community. Users can extend the tool repository
-provided by Galaxy using their own Galaxy server. One of the major use of
-Galaxy is to achieve a mix and match of data analysis tools in order to analyze
-a vast repository of genome based database. Galaxy offers a web-based,
-interactive platform for data analysis via reusable and reproducible workflows.
-While excelling in the usability aspects, the Galaxy environment has a limited
-parallelism capabilities and a small number of existing implementations for
-interfacing with diverse, scalable computing infrastructures. The user can draw
-workflows in which multiple copies of a task are created to operate on multiple
-inputs at once, but this is not a scalable approach.
+%
 

### 2014-06-01 text/askalon.tex
separate desc

\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer2014} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.\km{Todo: current developments and advances.}

### 2014-06-01 text/galaxy.tex
separate desc

\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
SWE developed at the Penn State University, Pennsilvania. It provides a
webportal based environment to build complex pipelines from an available
repository of tools (\textit{http://g2.bx.psu.edu}).  Galaxy is currently used
mainly by bioinformatics community. Users can extend the tool repository
provided by Galaxy using their own Galaxy server. One of the major use of
Galaxy is to achieve a mix and match of data analysis tools in order to analyze
a vast repository of genome based database. Galaxy offers a web-based,
interactive platform for data analysis via reusable and reproducible workflows.
While excelling in the usability aspects, the Galaxy environment has a limited
parallelism capabilities and a small number of existing implementations for
interfacing with diverse, scalable computing infrastructures. The user can draw
workflows in which multiple copies of a task are created to operate on multiple
inputs at once, but this is not a scalable approach.

### 2014-06-01 text/kepler.tex
separate desc

\paragraph{Kepler/MoML} Kepler~\cite{kepler2013} is a SWE
based upon the Ptollemy II framework~\cite{eker-janneck-etal:2003} developed at
the University of California. Kepler workflows follow an actor-director
metaphor where an actor represents a processor and the director
represents the paradigm for executing the workflows consisting of the actors.
Depending upon the types of directors, a Kepler workflow may exhibit a
different type of execution behavior. A workflow designer can chose from among
directors based upon the purpose of a workflow. Few of these include Process
Network (PN) for a parallel execution at the workflow level, Synchronous Data
Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
uses the Modeling Markup Language (MoML) for its workflow representation. MoML
is an extended XML syntax providing a rich vocabulary to compose Kepler
workflows. A workflow processor is called an `actor' in Kepler. Recent advances
in Kepler include MapReduce style computation, bioinformatics computation,
cloud computing~\cite{kepler2013}.

### 2014-06-01 text/makeflow.tex
separate desc

\paragraph{Makeflow/make-like spec} Makeflow~\cite{makeflow} is a make-like workflow platform.
Makeflow addresses the often-encountered problem of software and library
dependencies on workflow tasks execution for heterogeneous environments.
Makeflow offers techniques to ``link" and package the application execution
elements such as executables, libraries and configuration parameters. This
improves the durability of a workflow and decreases the efforts required to
maintain a working version of an application encoded as a workflow.

%Internal representation of a Makeflow workflow is in...
%Its expressivity features include...
%Makeflow's runtime features include...
%

### 2014-06-01 text/moteur.tex
separate desc

\paragraph{MOTEUR/scufl, GWENDIA} MOTEUR \cite{glatard-montagnat-etal:2008} is a SWE
developed at the I3S laboratory, University of Nice at Sophia Antipolis (UNS).
MOTEUR facilitates executing workflows consisting of remote services. It
exploits service parallelism at workflow level and data parallelism for
multiple datasets.

MOTEUR has sophisticated language-level expressions for data management
mechanisms within a workflow definition. It provides a diverse set of
parallel iterators that are well adapted for different combinations of input datasets
at the processor. An iteration strategy allows different combinations of the
data arriving at a processor along its multiple inputs. Running legacy
application code on distributed systems has been eased by MOTEUR by providing a
generic grid-service wrapper for these applications
\cite{glatard-montagnat-etal:2008a}. Services thus created can be readily
deployed by the MOTEUR runtime. MOTEUR workflows are written in the scufl
language and are compatible with those written for the Taverna SWE. It also
supports an extension of scufl language called the Gwendia~\cite{gwendia}
language, developed at the same laboratory. 
%

### 2014-06-01 text/peg.tex
separate desc

%
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is an SWE
well-suited to generate and execute large workflows from abstract DAGs
describing dataflow among interdependent tasks. Abstract description of a DAG
is done via an XML-based language called DAX. The concrete expression takes
into account the properties of application task requirements and underlying
distributed compute resources.  The main advantage of such approach is that a
single abstract workflow description can be mapped to multiple execution
environments based on requirements. DAX describes the logical workflow
components, their inputs and outputs and their links and dependencies in a
Pegasus workflow. These kind of workflows have been shown to enact optimally
using Pegasus' dynamic task, memory and disk management techniques. For
instance, for massively parallel task scheduling, Pegasus provides a task
clustering technique that groups parallel jobs based on fixed parameters such
as number of groups. Additionally, a manager buffers and maintains a fix-sized
window of executable tasks before accepting new tasks.

Large scale workflows from astronomy, biology, and high-energy physics have
been shown to run via Pegasus on distributed computing resources such as
Clusters, Grids and Supercomputers.  Recent developments in Pegasus enables it
to run workflows on cloud environments~\cite{peg-cloud}. Pegasus system is
known to be extendible as illustrated by a recent integration
effort~\cite{peg-hubzero} between Pegasus and the Hubzero
portal~\cite{hubzero}.

### 2014-06-01 text/pgrade.tex
separate desc

\paragraph{The P-GRADE Portal/--} The P-GRADE Portal~\cite{pgrade2011}
developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
provides an access to grid computational and storage resources from a web
browser through a specialized workflow oriented grid portal. A workflow
development environment interfaces the users through this web based portal. The
portal interfaces the multiple grid computing middleware and storage resources
as a backend component of the environment. A highlight of P-GRADE portal is
that it provides access to multiple grids and scientific communities in a
single environment leading to a highly collaborative inter-community,
inter-grid research environment. An older version of P-GRADE used the
condor-Dagman as workflow engine while the recent version uses its own in-house
developed workflow engine called Xen. \km{Investigate cloud and GPU capabilities}

### 2014-06-01 text/swift.tex
separate desc

\paragraph{Swift/Script, Tcl} Swift~\cite{swift} is a script-based scientific
workflow environment developed at the University of Chicago and Argonne
National Laboratory. A Swift workflow specified in the Swift scripting language
and is translated to an intermediate XML-based representation and enacted by
the Karajan workflow enactment engine. The SwiftScript language is proposed as
a workflow language founded on a scripting approach that programmers are
familiar with. To make the language data parallel, Swift script generalizes the
use of futures~\cite{futures} for every variables defined in the
script.  Futures are non-blocking assignment variables, using a proxy to ensure
immediate execution of assignment and performing lazy blocking on variable
value accesses only. The systematic use of futures for every variables in the
language makes it completely data-driven: the execution progresses
asynchronously as long as a data access is not blocking. The availability of
data enables blocked thread to restart execution. Swift facilitates a highly
asynchronous workflow execution. Swift supports advanced workflow constructs
such as parallel loops and conditionals. Furthermore Swift supports processing
of explicitly defined arrays of data. Swift workflows take advantage of
POSIX-style shared file system to reuse the data produced by other processors.
Recently, Swift has been shown to run on multiple clouds~\cite{ccgrid2013} via
its coasters mechanism~\cite{ucc2011}. Swift/T~\cite{turbine2013}, a new HPC
implementation of Swift system, runs scientific workflows over large-scale
supercomputing resources. It has been shown to work on petascale supercomputers
such as the IBM BlueGene and Cray series of machines. GeMTC project
investigating running many-task applications over GPUs via Swift/T. Efforts to develop metadata and provenance modules have been undertaken recently~\cite{swiftprov}.

### 2014-06-01 text/taverna.tex
separate desc

\paragraph{Taverna/scufl} The Taverna Workflow Management System
\cite{taverna2013} is an open source scientific dataflow manager
developed by the mygrid~consortium in the UK.
%
Taverna includes a GUI-based rich-client workbench for workflow design and a
workflow enactment runtime. Taverna can enact workflows expressed in the scufl
language. Scufl is a near Turing complete dataflow language with well-defined
semantics \cite{turi-missier-etal:2007,glatard-montagnat:2008}.

A Taverna workflow consists of a collection of processors connected by data and
coordination links, which establish a dependency between the output(s) of a
processor and the input(s) of another. The processors are of different kinds
depending on the application code to be invoked: typically being services, with
input and output \textit{ports} that correspond to the operation's invocations
and response sequences defined in the service`s interface expressed as REST or WSDL.
Java classes and local scripts can be used as workflow processors, as long as
they expose their signature as input and output ports as required by the model.
The engine orchestrates the execution of the processors in a way that is
consistent with the dependencies, and manages the flow of data through the
processors' ports. The overall workflow execution is data-driven, with a
\emph{push} model: a processor's execution is started as soon as all of its
inputs are available. Taverna is a flexible and extendible workflow environment
\cite{missier-soiland-reyes-etal:2010,sroka2009a} in that extending the
engine's functionality is facilitated through a Service Provider Interface
(SPI). The Taverna engine implements a sophisticated mechanism for enactment of
parallel workflows. These mechanisms manifest in the form of complex pipelines
and an advanced superpipelined enactment of workflows. Such a mechanism induces
multiple downstream workflow processors running in parallel for a dataset. In
the case of superpipelines, multiple pipeline instances run for multi-datasets
providing a high degree of parallelism limited only by the availability of
underlying resources. Taverna is not natively interfaced with external
computational infrastructure but the same can be achieved with the help of
extending the engine through one of its
SPIs~\cite{maheshwari-missier-etal:2009}. In a crucial development, recent
implementations of Taverna offers a client-server based environment in which a
Taverna \emph{server} can remotely serve the workflow composition and execution
services. This mode is useful in the cloud computing environments where the
server decides where to run the Taverna services in the cloud.

### 2014-06-01 text/triana.tex
separate desc

\paragraph{Triana/XML} Triana \cite{taylor-shields-etal:2003} is a java-based
SWE developed at the Cardiff University. It consists of a workflow composition
GUI, a framework to integrate workflow with the different execution
environments and middlewares and a set of predefined services written as java
components. A Triana workflow is expressed as an XML grammar similar to a WSDL
description for webservices \cite{taylor-wand-etal:2005}. A highlight of
Triana is that it provides ready-to-use interface for a wide variety of
middleware and distributed infrastructures including P2P, grids and
web-services frameworks. A Triana workflow provides dataflow as well as control
flow support. Triana allows conditional branching and looping in its workflows.

### 2014-06-01 text/wings.tex
separate desc

\paragraph{Wings/--} The Workflow instance generation and specialization
(Wings) provides automatic workflow validation and generation based on
high-level workflow templates~\cite{wings2013}. The system receives high-level
workflow specification in the form of Pegasus/DAX workflows and validates them
for correct coordination based on data and resource constraints. It generates
workflows by selecting components, data sets, and parameters in three stages
while eliminating candidates that are not viable. This approach is useful in
the context of desiging complex computational experiments by domain scientists
in that it eases the process of workflow level debugging. Wings uses W3C's OWL,
Resource Description Framework (RDF), and Semantic Web Rule Language (SWRL) to
represent workflows and their associated constraints.

% How does Wings achieve the translation between abstract and concrete workflows?
% What are the constraints under which Wings works?
% How is Wings implemented?

%

### 2014-06-02 paper.tex
more refactor

@@
 %
-\subsection{Architecture of a Generic SWE}
+
+\subsection{Architecture}
 Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
@@
 
-\section{Related Work}
-Surveys and studies on workflows similar to the present one have been done in
-the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
-SWEs in the previous decade, studying many existing systems for distributed
-computing. This taxonomy presented is based, to a large extent on the
-capabilities of the workflow engines. It does provide insights to many of the
-SWE aspects considered in the present paper, especially the workflow languages
-expressiveness and SWEs interactions with the distributed computing
-infrastructures in the early days of the compute Grids. A more recent work by
-Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely related to
-the workflow language expressiveness: it compares the models of computations of
-various workflow directors implemented in Kepler
-\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
-composition of different models in sub-workflows of a global orchestration.
-This work illustrates how different enactors designed for different application
-needs have different expressiveness power. Deelman et. al.
-\cite{deelman-overview} present a high-level overview of the scientific
-workflow features and capabilities including SWE composition, representation
-and resource mapping capabilities. A recent survey from
-Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance
-takes into account the latencies involved during the enactment mechanisms. A
-small survey and analysis of file-access characteristics for data-intensive
-workflows has been done by Shibata~\cite{Shibata}. Early surveys on workflow
-languages and suitability of general purpose languages to workflows is
-conducted in~\cite{generalpurpose}. These surveys were conducted before a few
-years and in the meantime SWEs have come a long way in terms of capabilities
-and enactment approaches. New techniques such as the map/reduce framework
-\cite{mapreduce} have been introduced as a promising new approach to workflow
-enactment over distributed infrastructures. Several issues put forth by such
-surveys such as customized service wrapping and enactment have been adequately
-addressed over the
-years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.
-
-Another well-studied feature is the workflow languages. Many different
-languages have been considered within the SWE community, from raw Directed
-Acyclic Graphs (DAGs) of computational processes such as CONDOR
-DAGMan~\cite{dagman} to abstractions for parallel computations such as Petri
-nets~\cite{alt-hoheisel:2005}, data-driven languages such as
-Scufl~\cite{turi-missier-etal:2007} and scripting languages such as Swift
-script~ \cite{swift}. Glatard's PhD thesis~\cite{glatard:2007} presents a
-taxonomy of SWE languages classifying them into functional, graph,
-service-oriented and executable representations. A detailed classification and
-taxonomy of parallelism patterns in Grid Workflows has been presented by
-Pautasso and Alonso in~\cite{pautasso1}. This classification is more generic
-while the one presented in Glatard's thesis has emphasis on service based
-workflows.  Each of these approaches can be defended through some aspects well
-covered in their design: DAGs are convenient for scheduling
-\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
-properties such as potential deadlocks, data-driven languages ease the
-description of application logic for non-expert users and scripting is
-extensively used by programmers for prototyping, etc.
-
-\section{Objectives and Organization}
-% What are the goals of this work?
-% How is this paper supposed to help:
-% -- Existing scientific users who have not adopted workflows
-% -- Existing users who have adopted workflows
-% -- New users
-% -- Experts in the field
-
-% Say that different workflow systems have different design goals. It is hard
-% to chose one from the existing or build your own. It is almost equally
-% complex to build own just as it is complex to create own programming
-% platform.
-We hope to provide by the way of this paper a guide for new adopters on how to chose a workflow language appropriate for their needs.
-We ask and answer the following questions of the workflow systems that we study in this paper:
-%-- What were the original design goals?
-%-- What are the main application areas that the system is currently being used?
-%-- What are the characteristics and capabilities of the language? expressivity features?
-%-- What are the characteristics and capabilities of the run-time? parallelism? concurrency? task and data grouping?
-%-- What kinds of computing and computational infrastructure does the workflow system cater to? Grids, Clouds, Clusters, Supercomputers, fast IB networks? Accelerators?
-%-- How have they evolved? What are the possible future directions?
-
-In the present work we survey and study the various contemporary SWEs in order
-to find the capabilities and key features in terms of the characteristics
-described in section~\ref{sec:intro}. We believe this study will serve to
-address the following objectives:
-
-\begin{enumerate}
-\item \textbf{Educate.} With more science being done with the help of
-computational methods, SWEs can add significant value to the process.  They can
-accelerate the process, enable community driven science, drive the development
-at a manageable level of abstraction for users.  However, scientific users are
-often not experienced programmers or computation savvy. This work will serve as
-a useful starting guide to educate science users and students of the available
-tools and techniques with features suitable for their methods.
-
-\item \textbf{Catalog.} We study a number of important features as offered by
-the contemporary SWEs. This will help catalog and collect the SWEs by their
-offerings and identify mechanisms and needs for new and existing features
-in one place. This will also help identify current trends and approaches
-to computation in scientific communities.
-
-\item \textbf{Application.} We identify the applications and application
-classes well adapted to the workflow systems and are benefitting by them.
-This will help readers relate to the application they are associated to.
-We hope that readers can make a choice of which SWE to use if there is an
-adaptation similar in scope to the ones describe in this paper.
-
-\item \textbf{Evolution.} We comment on the evolution of the existing workflow
-systems and the possible current and future development directions which
-will help readers be aware of the general development direction of the
-SWEs.
-\end{enumerate}
-
-The rest of this paper is organized as follows: Section
-\ref{sec:overview} presents an overview of the key SWEs studied in this work.
-Section \ref{sec:scal} discusses the optimization features and resulting
-scalable performance as observed in the SWEs studied. Section \ref{sec:data}
-studies the data management and description aspects of the SWEs. Section
-\ref{sec:inter} presents a study of SWEs approaches to interface the
-large-scale computing infrastructures. Section \ref{sec:lang} details the
-language features and expressiveness of the SWEs studied. Section
-\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
-\ref{sec:sum} presents a summary and a classification of all the aspects of
-SWEs studied in this work.
-
-Throughout the paper, we also present tables
-summarizing the aspects of SWEs considered in respective sections. It was not
-possible to obtain an exhaustive information about all the SWEs considered in
-this work. We rely on the following sources for our information: 
-\begin{enumerate}
-    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
-    \item \textbf{Technical Reports.} Technical reports and theses.
-    \item \textbf{Web Presence.} Documentation and other information from the web pages hosting the SWE.
-    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
-\end{enumerate}
-
-Some features, such as extendability of an SWE strongly depend on the extension
-points designed in the code; consequently, we consider it present in only those
-SWEs that have an explicit documented description or demonstration of the given
-feature.
+\input{text/related}
+\input{text/objectives}
 
@@
 \input{text/galaxy}
-% Put this somewhere else, either in the analysis or may be in the conclusions
-% and summary.
+% Put this in analysis section. 
 %The \textit{gscript}~\cite{maheshwari-montagnat:2010} workflow language
-%is similar in spirit to the
-%SwiftScript. Similar to SwiftScript, gscript adopts futures to express
+%is similar in spirit to the SwiftScript. Similar to SwiftScript, gscript adopts futures to express
 %asynchronous execution in a script-based language. A Swift script 
@@
 
-The software architecture of an enactor plays a key role in determining the
+The software architecture of the system plays a key role in determining the
 scalability for the workflow applications exhibiting a large number of tasks.
@@
 
-In this section we analyze the scalability measures and techniques employed by
-contemporary SWEs. Illustrated below are some of the techniques used to achieve
-scalability in various SWEs.
+In this section we analyze the scalability and runtime optimization measures
+and techniques employed by SWEs. Illustrated below are some of the techniques
+used to achieve scalability in various SWEs.
 

### 2014-06-02 text/objectives.tex
more refactor

\section{Objectives and Organization}
% What are the goals of this work?
% How is this paper supposed to help:
% -- Existing scientific users who have not adopted workflows
% -- Existing users who have adopted workflows
% -- New users
% -- Experts in the field

% Say that different workflow systems have different design goals. It is hard
% to chose one from the existing or build your own. It is almost equally
% complex to build own just as it is complex to create own programming
% platform. We hope to provide by the way of this paper a guide for new adopters on how to chose a workflow language appropriate for their needs. We ask and answer the following questions of the workflow systems that we study in this paper:
%-- What were the original design goals?
%-- What are the main application areas that the system is currently being used?
%-- What are the characteristics and capabilities of the language? expressivity features?
%-- What are the characteristics and capabilities of the run-time? parallelism? concurrency? task and data grouping?
%-- What kinds of computing and computational infrastructure does the workflow system cater to? Grids, Clouds, Clusters, Supercomputers, fast IB networks? Accelerators?
%-- How have they evolved? What are the possible future directions?

In the present work we survey and study the various contemporary SWEs in order
to find the capabilities and key features in terms of the characteristics
described in section~\ref{sec:intro}. We believe this study will serve to
address the following objectives:

\begin{enumerate}
\item \textbf{Educate.} With more science being done with the help of
computational methods, SWEs can add significant value to the process.  They can
accelerate the process, enable community driven science, drive the development
at a manageable level of abstraction for users.  However, scientific users are
often not experienced programmers or computation savvy. This work will serve as
a useful starting guide to educate science users and students of the available
tools and techniques with features suitable for their methods.

\item \textbf{Catalog.} We study a number of important features as offered by
the contemporary SWEs. This will help catalog and collect the SWEs by their
offerings and identify mechanisms and needs for new and existing features
in one place. This will also help identify current trends and approaches
to computation in scientific communities.

\item \textbf{Application.} We identify the applications and application
classes well adapted to the workflow systems and are benefitting by them.
This will help readers relate to the application they are associated to.
We hope that readers can make a choice of which SWE to use if there is an
adaptation similar in scope to the ones describe in this paper.

\item \textbf{Evolution.} We comment on the evolution of the existing workflow
systems and the possible current and future development directions which
will help readers be aware of the general development direction of the
SWEs.
\end{enumerate}

The rest of this paper is organized as follows: Section
\ref{sec:overview} presents an overview of the key SWEs studied in this work.
Section \ref{sec:scal} discusses the optimization features and resulting
scalable performance as observed in the SWEs studied. Section \ref{sec:data}
studies the data management and description aspects of the SWEs. Section
\ref{sec:inter} presents a study of SWEs approaches to interface the
large-scale computing infrastructures. Section \ref{sec:lang} details the
language features and expressiveness of the SWEs studied. Section
\ref{sec:usable} discusses the usability aspects of the SWEs.  Section
\ref{sec:sum} presents a summary and a classification of all the aspects of
SWEs studied in this work.

Throughout the paper, we also present tables
summarizing the aspects of SWEs considered in respective sections. It was not
possible to obtain an exhaustive information about all the SWEs considered in
this work. We rely on the following sources for our information: 
\begin{enumerate}
    \item \textbf{Publications.} Peer reviewed publications in journals, conferences and workshops.
    \item \textbf{Technical Reports.} Technical reports and theses.
    \item \textbf{Web Presence.} Documentation and other information from the web pages hosting the SWE.
    \item \textbf{Experience.} Where possible, we tried features of the SWEs by installing them on the available systems. 
\end{enumerate}

Some features, such as extendability of an SWE strongly depend on the extension
points designed in the code; consequently, we consider it present in only those
SWEs that have an explicit documented description or demonstration of the given
feature.

### 2014-06-02 text/related.tex
more refactor

\section{Related Work}
Surveys and studies on workflows similar to the present one have been done in
the past. Yu and Buyya \cite{yu-buyya:2005} proposed a detailed taxonomy of
SWEs in the previous decade, studying many existing systems for distributed
computing. This taxonomy presented is based, to a large extent on the
capabilities of the workflow engines. It does provide insights to many of the
SWE aspects considered in the present paper, especially the workflow languages
expressiveness and SWEs interactions with the distributed computing
infrastructures in the early days of the compute Grids. A more recent work by
Goderis \cite{goderis-brooks-etal:2009} studies a topic more closely related to
the workflow language expressiveness: it compares the models of computations of
various workflow directors implemented in Kepler
\cite{ludascher-altintas-etal:2005a} with the perspective of enabling
composition of different models in sub-workflows of a global orchestration.
This work illustrates how different enactors designed for different application
needs have different expressiveness power. Deelman et. al.
\cite{deelman-overview} present a high-level overview of the scientific
workflow features and capabilities including SWE composition, representation
and resource mapping capabilities. A recent survey from
Stratan~\cite{stratan-iosup-etal:2008} about the workflow engine performance
takes into account the latencies involved during the enactment mechanisms. A
small survey and analysis of file-access characteristics for data-intensive
workflows has been done by Shibata~\cite{Shibata}. Early surveys on workflow
languages and suitability of general purpose languages to workflows is
conducted in~\cite{generalpurpose}. These surveys were conducted before a few
years and in the meantime SWEs have come a long way in terms of capabilities
and enactment approaches. New techniques such as the map/reduce framework
\cite{mapreduce} have been introduced as a promising new approach to workflow
enactment over distributed infrastructures. Several issues put forth by such
surveys such as customized service wrapping and enactment have been adequately
addressed over the
years~\cite{glatard-montagnat-etal:2008a,rojasbalderrama-montagnat-etal:2010}.

Another well-studied feature is the workflow languages. Many different
languages have been considered within the SWE community, from raw Directed
Acyclic Graphs (DAGs) of computational processes such as CONDOR
DAGMan~\cite{dagman} to abstractions for parallel computations such as Petri
nets~\cite{alt-hoheisel:2005}, data-driven languages such as
Scufl~\cite{turi-missier-etal:2007} and scripting languages such as Swift
script~ \cite{swift}. Glatard's PhD thesis~\cite{glatard:2007} presents a
taxonomy of SWE languages classifying them into functional, graph,
service-oriented and executable representations. A detailed classification and
taxonomy of parallelism patterns in Grid Workflows has been presented by
Pautasso and Alonso in~\cite{pautasso1}. This classification is more generic
while the one presented in Glatard's thesis has emphasis on service based
workflows.  Each of these approaches can be defended through some aspects well
covered in their design: DAGs are convenient for scheduling
\cite{Malewicz:2007,hall-rosenberg-etal:2007}, Petri nets can be used to detect
properties such as potential deadlocks, data-driven languages ease the
description of application logic for non-expert users and scripting is
extensively used by programmers for prototyping, etc.

### 2014-06-02 paper.tex
refactor

@@
 
-\section{Conceptual Overview}
-In this section we present a conceptual overview of the subjects of this
-study--Workflows and Scientific Workflow Environments.
-
-\subsection{Workflow}
-Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
-Graph (DAG). Execution typically begins with the input data being provided to
-the first task and ends at the results being delivered at the end of the last
-task. Tasks are represented by boxes and data/control flow between the task is
-represented as arrows. Note that in cases where tasks are connected with pure
-control flow, it can be simulated by passing token data between tasks.
-
-For the purpose of this text, we define a workflow as a specification of
-computation with at least two discrete tasks either forming distinct
-computational stages or running for two or more distinct datasets as
-parameters. Consequently, a one-off task or running a task for a single copy of
-data is \emph{not} a workflow.
-
-For the rest of this paper, we will use the above definition of a workflow.
-%
-
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=4cm]{figures/workflow}
-\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
-\label{fig:wf}
-\end{center}
-\end{figure}
-%
-
-\subsection{Architecture}
-Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
-components and their interaction in an SWE. These components are the focus of
-interest for the present work. A \textit{user interface} interacts with the
-\textit{workflow engine} using commands and messages on the inside while with
-user on the outside. The \textit{workflow language} becomes an intermediate
-representation for communication between user interface and the
-\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
-and a high-level scheduler for the processors and interfaces with the
-underlying \textit{distributed resources}. It resolves data dependencies and
-optimizes the flow at runtime and monitors the workflow progress and collects
-the intermediate and final results.
-
-%
-\begin{figure}[htb]
-\begin{center}
-\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
-\caption{A generic SWE scheme showing the relationship and interaction among important components}
-\label{fig:wfms}
-\end{center}
-\end{figure}
-%
-SWEs are used as tools to benefit from the underlying large-scale computing
-infrastructure such as clusters, supercomputers, clouds, and grids. Such
-environments are in use today for production level work in different scientific
-domains worldwide. They do the basic task of linking inter-related processors
-of a multi-staged scientific computation. However, they differ in the features
-they provide based upon the scientific domain or user community they offer
-service for, and the computational model they follow. Often, one of the main
-reasons for such difference is the specialized components and requirements for
-a particular domain. Owing to this fact, a given workflow system dominates a
-particular area of science (eg. astronomy, biology, etc.). Nonetheless, there
-exist a plethora of SWEs suitable for general purpose workflow style
-computations.
-
-DAGs are by far, the most popular abstraction of the way workflows are
-perceived and visualized. The MA DAG~\cite{caron-desprez:2005}, engine is one
-example of a DAG generator engine. MA DAG has evolved from a regular DAG
-generator to a more abstract DAG generator laced with control structures. MA
-DAG benefits from the workflow scheduling mechanisms included in the
-middleware. Due to the impossibility to produce complete DAGs owing to the
-presence of control structures such as conditionals and loops, MA DAG generate
-partial DAGs greedily: a sub-DAG, as complete as possible, is produced as soon
-as possible. Despite the presence of many different approaches to workflow
-systems, new requirements continue to pose workflow challenges. Some of the
-current challenges facing SWEs are:
-
-\begin{enumerate}
-\item \textbf{Multisite execution.} Interfacing to existing computation and
-storate infrastructure is not trivial, not to mention the new and
-unforeseen access and compute modes. SWEs are yet to formulate a
-generalized interface towards the major platforms such that they will
-run on them out of the box.
-
-\item \textbf{Manysite execution.} Even if a SWE has interfaces built for
-more than one infrastructure, it is hard to optimally schedule large
-workflows to many remote sites simultaneously. The problem of limited
-control over remote sites and distributing interdependent data and
-tasks makes it non-trivial to efficiently load-balance work among them.
-
-\item \textbf{Expressivity.} Workflow languages have evolved over the years
-with reasonable expressivity being supported. However, they are often
-compared with those of the traditional programming languages. As of
-current state of the art, the workflow languages expressivity still
-have to catch up to the richness of programming languages across
-different generations such as C and python.  The community is yet to
-realize a truly generic workflow language that can express a wide
-variety of flow patterns encountered in scientific domains.
-
-\item \textbf{Interoperability.} Interoperable workflows will benefit to a
-broad audience of users. Many efforts at this are made in the recent
-past, such as by the consurtium projects and by individual players.
-However, a truly interoperable set of standards for workflows are yet
-to emerge.
-\end{enumerate}
-
+\input{text/concepts}
 \input{text/related}
@@
 \input{text/galaxy}
+\input{text/others}
 % Put this in analysis section. 
@@
 %
-%
+\input{text/scale}
+\input{text/data}
+\input{text/interface}
+\input{text/lang}
+\input{text/usability}
 
-% == SYSTEM SCALABILITY AND OPTIMIZATION ==
-\section{Scalability and Optimization} \label{sec:scal}
-%We have results comparing \MOTEUR and Taverna performance!
-%For Pegasus: Scaling up workflow based applications.
-A simple definition of scalability in the context of SWE is that the
-performance of the workflow engine increases linearly with the increasing
-number of similar tasks. Depending upon the lag or increase in the linearity we
-say a workflow enactor scales up or down.  In this section, we focus on
-the scalability aspects of the workflow runtime. We do not consider the
-specific scalability issues related to a distributed infrastructure such as
-network latencies and protocol related overheads. However, there have been efforts to
-thwart such issues at the workflow level by combining workflow enactment
-techniques with those of distributed computing.  One such example is the
-map/reduce framework \cite{mapreduce}.  Recently, the map/reduce framework,
-that aims at performing distributed processing on large datasets, has gained a
-significant popularity in workflow community. It is based on the master/worker
-metaphor of distributed processing.  It employs the commonly used \textit{map}
-and \textit{reduce} functions from the functional programming languages. The
-map function chops a large chunk of data into smaller pieces and distributes it
-to \textit{workers} which in turn can also chop it to further smaller datasets
-forming a tree-like hierarchy. The worker process the data and return the
-results back to their parents. The reduce step collects the results from the
-workers and combines them to obtain the final result. One of the main reason
-for the popularity of the map/reduce framework for workflows is the recent
-attention to collections based datasets and processing. There exists
-implementations of the map/reduce framework written in different languages with
-enhancements and support services such as a sophisticated query service. Hadoop
-(\textit{http://hadoop.apache.org/}) is one such implementation which has
-already been interfaced by the Kepler SWE~\cite{wang-crawl-etal:2009}.
-
-Large-scale applications with thousands of tasks often pose memory and disk
-management difficulties leading to scalability issues.  However, the
-performance of a workflow system is dependent on several factors including the
-type of tasks and approach for submitting tasks for execution.  There is not
-much research done towards the enactor scalability baring a couple of studies
-carried out by C. Stratan et. al. \cite{stratan-iosup-etal:2008} and Deelman
-et. al. \cite{callaghan-deelman-etal:2010}.  One way to compare the scalability
-of different workflow environments is to experiment with a similar set of
-workflows on a given system. In addition, different SWEs interface to different
-DCIs. This is a time consuming process.  In the current study we do not
-directly compare the workflow system scalability but we study the similarities
-and differences in the techniques employed by workflow engines in order to
-achieve scalability.
-
-The software architecture of the system plays a key role in determining the
-scalability for the workflow applications exhibiting a large number of tasks.
-An architecture that takes into account the properties of the underlying
-infrastructure could scale well in the face of growing number of workflow
-tasks.
-
-In the context of workflow runtime, scalability can be further classified in
-terms of data motion scalability. It can be expressed as the rate of increase
-in number of data movement quantity with respect to the increase in the elapsed
-time.
-
-While the computational scalability of a large scale experiment depends upon
-the underlying computing infrastructure, at the workflow enactor level, it
-becomes important to be able to prepare and dispatch as many processor
-instances as possible without loss in performance. Failing to do so for a
-workflow engine can become a performance bottleneck and hamper the optimal
-utilization of an available computing infrastructure.
-
-Most workflow systems exploit the parallelism at three-levels. First, at the
-task-level wherein the independent branches of a workflow graph are enacted
-concurrently. Second, at the data-level, where a same processor is instantiated
-multiple times based upon the repeated operations on different datasets. Third,
-the low-level parallelism that is encoded within the application components
-independent of the workflow. At this level, the workflow enactor may
-need to make a matching based upon the application type. For instance, a
-workflow enactor may schedule the GPU-based processors on the resources that
-indicate as possessing such capabilities.
-
-In this section we analyze the scalability and runtime optimization measures
-and techniques employed by SWEs. Illustrated below are some of the techniques
-used to achieve scalability in various SWEs.
-
-Taverna version 2.x introduced new measures in order to be able to submit a
-large number of jobs simultaneously. It is able to dispatch large number of
-concurrent instances of processors. An ability to implement
-superpipelined~\cite{hwang:1992} execution of workflow tasks leads to a maximum
-expression of parallelism~\cite{pipeline}.
-
-Within the Taverna engine, all the data involved in the execution is managed by
-a data manager through an external database, and is addressed by the engine
-only by \textit{reference}. This facilitates the management of data-intensive
-workflows in a scalable way, as high volume data is only transferred into the
-engine space when needed by processors that are unable to resolve external data
-references on their own.
-
-Taverna, Swift-Karajan and Pegasus all provide a means to limit the maximum
-number of processor instances that can run concurrently through a builtin
-parameter. In Taverna, this can be configured at a workflow level while in
-Pegasus, this can be configured at the Condor \cite{thain-tannenbaum-etal:2005}
-scheduler level. In Swift-Karajan system this parameter is regulated
-automatically as part of a ``throttling parallel foreach" construct. A similar
-configurable parameter is provided by the Knime thread manager in order to
-limit the size of the thread-pool employed to run a workflow.
-
-Kepler optimizes the workflow execution through parallel and pipelined
-execution. In addition to these, Kepler also provides dataflow analysis
-optimization that involves data bypassing the processors in a workflow that
-does not make use of that data. This is detected using a builtin Kepler
-mechanism called ``static type propagation".
-
-Another alternative is to optimize the enactment at runtime. This approach has
-advantages including a better information about the actual data available. For
-instance, Askalon does runtime branch prediction, parallel loop unrolling and
-loop sequentialization.
-
-The Pegasus workflow manager introduced subworkflows because of scalability
-issues at the localhost for a large number of nodes in a DAG. A hierarchical
-representation helps it scale at the local planning level as well as at the
-execution level on distributed resources.
-
-Static planners in Pegasus translate a workflow graph into a location specific
-DAGMan input file, adding stages for data staging, intersite transfer and data
-registration. These planners can reduce processing of tasks for files that
-already exist, select sites for jobs, and cluster jobs based on various
-criteria. Pegasus performs graph transformation with the knowledge of the whole
-workflow graph. A similar strategy is grouping tasks by application-level
-wrappers written in higher-level programming languages such as bash or Python.
-
-In some systems the structure of a workflow is constructed and expanded
-dynamically as opposed to static planning. A static workflow graph allows
-optimizations based on the global state of the workflow, whereas a dynamic
-workflow structure allows adaptations to changing environments.  Depending upon
-the execution environment and access conditions to the available resources,
-each approach can result in the overall optimization of a given application.
-
-Knime has a different execution strategies for shared memory Symmetric
-Multiprocessing Systems (SMP) and distributed memory systems. In the case of
-SMP, it uses the message passing library or MPI to distribute the independent
-parts of a workflow to available processors. This approach results in efficient
-use of available processors. However, for a large scale, the global memory
-becomes a bottleneck hindering the scalability.
-
-In the case of distributed memory and independent computational nodes connected
-with highspeed network or the internet, Knime employs a Master/Worker metaphor
-of computing. A Knime installation with GUI acts as a master node that
-coordinates the bare Knime installations on the worker nodes. These worker
-nodes are then sent the portions of a workflow by the master node. The master
-node keeps track of the participating nodes by their unique IP addresses. At
-the end of each execution, the master node collects the results from the worker
-nodes.
-
-\begin{table}[ht!]
-%TODO: Caption is not printing well, fix
-%\caption{A summary of SWEs Scalability and optimization Mechanisms}
-\begin{center}
-% use packages: array
-\begin{tabular}{|l|l|l|l|l|l|}
-\hline
-\textbf{SWE} / & \multicolumn{3}{c|}{\textbf{Parallelism}} & \multicolumn{2}{c|}{\textbf{Optimization}}\\
-\cline{2-6}
-\textbf{Language} & \textbf{data} & \textbf{task} & \textbf{pipeline} & \textbf{static} & \textbf{runtime}\\
-\hline
-Taverna/Scufl2 & \checkmark & \checkmark & \checkmark & -- & \checkmark \\
-\hline
-Triana/- & \checkmark & \checkmark & \checkmark & -- & \checkmark \\
-\hline
-MOTEUR/Scufl & \checkmark & \checkmark & \checkmark &-- &\checkmark \\
-\hline
-Kepler/MoML & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-\hline
-GridNexus/GXPL & -- & \checkmark & -- &-- &\checkmark \\
-\hline
-Swift/SwiftScript & \checkmark & \checkmark & \checkmark & \checkmark &\checkmark \\
-\hline
-Askalon/AGWL &--& \checkmark & -- & \checkmark & \checkmark \\
-\hline
-P-GRADE/- &\checkmark &\checkmark &\checkmark &--&--\\
-\hline
-Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark \\
-\hline
-Galaxy/- &-- &-- &\checkmark &-- &-- \\
-\hline
-MOTEUR2/Gwendia~ & \checkmark &\checkmark & \checkmark &-- &\checkmark \\
-\hline
-\end{tabular}
-\label{tbl:scalability}
-\end{center}
-\end{table}
-
-Shown in Table \ref{tbl:scalability} is a summary of the Scalability and
-optimization mechanisms of the studied SWEs. The scalability features
-considered are through three levels of parallelism in the form of data,
-task and pipeline. The optimization features considered are the presence of
-mechanisms by which static and runtime optimizations are achieved.
-
-%
-% == DATA MANAGEMENT ==
-\section{Data Management and Description} \label{sec:data}
-Many aspects of data handling present significant diversity among the workflow
-managers. Data management approaches for scientific workflows can be classified
-as resource and language level. Resource level strategies include data
-management at the application level of abstraction where the data is required
-to be distributed, available and securely remote-accessible. It also determines
-how the data will be delivered to a resource that handles the processing and
-returning the processed data or results to the next processors in the workflow.
-Provenance, staging, and managing heterogeneous data structures (files, arrays
-etc.) are functional requirements at this level.
-
-Whereas the language level data management involves how the actual data items
-are described and handled across a workflow execution. Decisions such as how
-complex data structures are supported, transferred and delivered from workflow
-input source or from one processor to another are determined at language level.
-In most workflow systems, the transfer of data is achieved through `data-ports'
-attached to the processors. These ports act as channels to collect and transfer
-data across a workflow. The association of data description with the workflow
-determines whether the data is loosely or strongly coupled with a workflow. A
-loosely coupled workflow with data has advantage over the strongly coupled ones
-in the sense that the workflow is independent of its data and can be invoked
-with a different set of data.
-
-A source of complexity for the workflow manager is the existence of different
-formats of data for a single scientific experiment. These include the different
-legacy formats of storage as well as the different ways the data is organized
-including database, filesystems and remote links. Further, most remote data is
-stored at locations secured behind different types of secure communication
-protocols involving special access mechanisms.
-
-In the rest of this section, we study the data handling mechanisms of SWEs.
-Taverna's reference scheme makes sure the data produced and consumed is in the
-form of references. This approach of data management is efficient because: 1)
-It takes less time to transport references, and 2) Locality of references
-ensures that the data is processed at a location that is local to multiple
-processors operating on the same dataset.
-
-In MOTEUR and Kepler workflow environments, data is described as an XML grammar
-allowing for nested collections or multi-dimensional arrays of data. Kepler's
-COMAD \cite{keplercomad} workflow enactment model is well suited to collection
-oriented data like arrays and streams. Similar efforts exist elsewhere to model
-the streams for workflow programming
-\cite{blower-haines-etal:2005,herath-plale:2010}.
-
-MOTEUR only uses references to data files while Pegasus groups small files into
-a single compressed file for transfer thus saving transfer time. It optimizes
-the workflow by annotating and reusing the data already available, if
-appropriate. Pegasus is able to generate a workflow from a metadata description
-of the desired data product with the aid of Artificial Intelligence planning
-techniques. Grouping of tasks by Pegasus increases the data usage efficiency
-since a group of tasks that require the same set of files for execution are
-localized. Pegasus also manages the temporary storage space needed for data
-processing at a particular processor location and commands the local Operating
-System to allocate the space. A data provider abstraction allows data to be
-referenced in the Trident SWE from any external entities in the form of typed
-referenced data objects.
-
-A majority of SWE support primitive, MIME based or XSD data types similar to
-traditional programming datatypes. Specific support for complex and structured
-data types are provided by the SWEs based upon their domain and application
-requirements. For instance, Kepler supports a specialized byte data streams and
-Bioflow supports a natively developed life-science database engine. Others such
-as the SwiftScript support extensible structures such as filesystem mappers and
-customized C language style structures.
-
-Remote and secure data access is supported by SWEs by providing secured
-protocols based accession to remote data repositories. Third party protocols
-such as GridFTP, HTTP over TLS (HTTPS) and gLite LFC (LCG File Catalog) access
-are commonly supported.
-
-\begin{table}[ht!]
-\begin{center}
-\hspace*{-2.2cm}
-% use packages: array
-\begin{tabular}{|c|c|c|c|c|c|}
-\hline
-\textbf{SWE /} & \multicolumn{3}{c|}{\textbf{Language Level}} & \multicolumn{2}{c|}{\textbf{Resource Level}}\\
-\cline{2-6}
-\multirow{2}{*}{\textbf{Language}} & \textbf{Data Coupling} & \textbf{Data Types} & \textbf{Data Structures} & \textbf{Type} & \textbf{Access}\\
-& \textbf{with Workflow} & \textbf{Supported} & \textbf{Supported} & & \\
-\hline
-Taverna/Scufl2&Loose&MIME-types&multi-dim Arrays&Local+Remote&By Reference\\\hline
-+/BPEL&Strong&XSD-Based&--&Grids, Clouds& Adhoc\\\hline
-%
-\multirow{2}{*}{Triana/-}&Strong&XML Schema-&Streams,&Local+&Through \\
-&&Types&Java Objects&Remote&Adaptors\\\hline
-%
-\multirow{2}{*}{MOTEUR/Scufl}&Loose&MIME-&Lists&Local+&Services+\\
-&&Types&&Remote&Middleware\\\hline
-% 
-\multirow{2}{*}{Kepler/}&Strong&Primitive&Collections,&Filesystem&Customized\\
-MoML&&Types&User defined&&Actors\\\hline
-% 
-GridNexus/GXPL&Loose&XSD-Based&--&Grid Data&OGSA-DAI \\\hline
-% 
-\multirow{2}{*}{Swift/SwiftScript} &Strong&--&1-Dim Arrays,&Shared& POSIX-style\\
-&&&Customised&FileSystem&\\\hline
-% 
-Askalon/AGWL&Loose&MIME-types&Arrays&Data Repository&SQL-based\\\hline
-%
-\multirow{2}{*}{P-GRADE/-}&Loose&--&--&Grid&Secure\\
-&&&&File Catalogs&Protocols\\\hline
-%
-\multirow{2}{*}{}Pegasus/&Loose&--&--&Distributed&Secure\\
-DAX&&&&Filesystems&Protocols\\\hline
-% 
-Galaxy/-&Loose&--&File Streams&File Systems&--\\\hline
-% 
-\multirow{2}{*}{MOTEUR2/} &Loose&Integer,String,&multi-dim Arrays&Grid&By\\
-Gwendia~ &&Double,URI&&Data&Reference\\\hline
-\end{tabular}
-%\caption{A summary of SWEs Data Management Mechanisms}
-\label{tbl:datadesc}
-\end{center}
-\end{table}
-
-Table \ref{tbl:datadesc} presents a summary of the data management mechanisms
-of the studied SWEs. Language and resource level data management mechanisms are
-highlighted. At the language level, data description as coupled with that of
-the workflow, data types and structures are considered for the SWE studied. At
-the resource level, the major resource types supported and their access
-mechanisms are highlighted. We distinguish the remote form of Grid resource
-types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those
-resource access. Other resource types such as data repositories or file
-catalogues are the resource types specific to a given SWE. For instance,
-Askalon uses a specifically configured data repository to access data and
-application binaries.
-
-%Replication is one of the solution.
-% == INTERFACE TO COMPUTING AND DATA INFRASTRUCTURES ==
-\section{Interface to Distributed Computing Infrastructures (DCIs)} \label{sec:inter}
-Workflow environments are external systems to large scale computing
-infrastructures. They require an interface to connect to and use the
-infrastructures. Provided the complex and distributed nature of grids, it is
-not easy and practical to interact with such an infrastructure directly. Most
-distributed systems have a software service layer through which it interacts
-with the outside world. These interactions include accepting queries for
-information and accepting requests and data for job executions. Such a layer is
-typically called the \textit{middleware}. A workflow engine intending to
-interact with a grid environment must provide interface to link to such
-middleware layer.
-
-Most grid computing systems today are batch oriented execution systems (with a
-notable exception of GridRPC \cite{seymour-nakada-etal:2002} based systems). A
-batch of ``jobs" are submitted to the system which in turn passes through some
-execution queues before actually being scheduled for execution. Such systems
-are prone to high execution latencies and unreliability issues. A workflow
-manager must take these issues into account while interacting with such
-systems.
-
-At the resource management level, most grid computing environments today are
-utilizing one or more components of the Globus Toolkit
-\cite{foster-kesselman:1997} or the Condor distributed computing software
-system~\cite{dagman}.
-
-The Globus Toolkit is a software stack that provides useful functionalities for
-grid computing operations. These functionalities are provided in the form of a
-software stack that ranges from low-level data management tasks to the higher
-level functionalities like authentication and metadata services. Any external
-entity accessing the grid computing system must possess proper authentication
-and sufficient authority to operate upon the infrastructure. The most popular
-secure access system in place for today's grids is the public key
-infrastructure (PKI) system through which a secure access to the grid is
-guaranteed for a finite time period. A workflow manager must be capable of
-handling the security protocols with the underlying execution infrastructure.
-The duration of access to the system is also of critical importance as a
-typical scientific workflow is bound to run for a considerably long time
-period. The Condor software system provides scheduling mechanisms to access the
-low-layer grid resources. A grid workflow enactor must provide interface to
-access the services of these external scheduling and middleware systems.
-
-\begin{table}
-\begin{center}
-%\hspace*{-1cm}
-% use packages: array
-\begin{tabular}{|l|l|l|}
-\hline
-\textbf{SWE/Language} & \textbf{DCI Interfacing Mechanisms} & \textbf{Currently Interfaced DCIs}\\
-\hline
-Taverna/Scufl2 & By Extension & None Inherently \\\hline
-+/BPEL & Ad-hoc & None Inherently \\\hline
-Triana/- & Provides Adapters & P2P Networks, NGS (ngs.org.uk)  \\\hline
-MOTEUR/Scufl & Inherent & EGI (www.eu-egi.eu), g5k (www.grid5000.fr)\\\hline
-Kepler/MoML & Using Directors & EarthGrid (http://seek.ecoinformatics.org) \\\hline
-GridNexus/GXPL & Inherent & Generic Grids \\\hline
-Swift/SwiftScript & Inherent & IBM/bluegene/P HPC Clusters \\\hline
-Askalon/AGWL & Inherent & Austrian Grid (http://www.austriangrid.at/)\\\hline
-P-GRADE/- & Inherent & NGS \\\hline
-Pegasus/DAX & Using Condor & OSG (www.opensciencegrid.org) \\\hline
-Galaxy/- & Extensions & None Inherently \\\hline
-MOTEUR2/Gwendia~ & Inherent & EGI, g5k, HIPernet Cloud \\\hline
-\end{tabular}
-%\vspace*{1cm}
-%\caption{A summary of SWEs DCI Interfacing Mechanisms}
-\label{tbl:interfacedci}
-\end{center}
-\end{table}
-
-Remotely located grid resources are accessed normally over the Internet links.
-In such situations, long network latency or outright link failures are common.
-Workflow managers are likely to face issues such as no response from the job
-submitter or long delays. There must be some mechanism in place to handle such
-scenarios. Depending upon the sophistication of the underlying infrastructure a
-failover mechanism or logging information could be obtained for failures. Often
-times workflow managers have their own failover mechanisms in place.
-
-In the rest of this section, we study the approaches and techniques of SWEs to
-interact with the Computing and Data Infrastructures. We also discuss briefly
-the mechanisms for the mandatory security requirements.
-
-MOTEUR provides seamless access to local and remote grid data resources by
-closely coupling itself with the vBrowser~\cite{olabarriaga-deboer-etal:2006}
-virtual resource browser environment. It has a built-in job management
-mechanism that includes building job-specification, job submission data staging
-and collection of results from the jobmanager. Additionally, MOTEUR2 is
-interfaced with the HiPerNet cloud, a virtual platform manager that takes into
-account, both computing and network resources
-\cite{vicat-blancprimet-roca-etal:2009}.
-
-In Taverna, a processor's invocation sequence can be customized
-programmatically to enhance the processor's performance selectively. This
-feature can be exploited by effectively interfacing a Taverna processor to Grid
-middleware and create jobs and inputs/outputs definitions for that middleware.
-
-In Taverna, a processor's invocation sequence can be customized
-programmatically to enhance the processor's performance selectively. This
-feature can be exploited by effectively interfacing a Taverna processor to Grid
-middleware and create jobs and inputs/outputs definitions for that middleware.
-
-Pegasus interfaces explicitly and transparently with the infrastructure in that
-it maps the simplified workflow DAGs to the grid resources through the Condor
-resource management service. Pegasus interacts with the available replica
-management services in order to optimize the data access cost for the workflow
-executions at runtime. It does cluster--group similar jobs together, in order
-to ease the load on the scheduler and the dispatching of those jobs to the
-worker nodes.
-
-Swift/Karajan and Pegasus interface to a POSIX-style shared file system which
-is one crucial aspect to their performance. %== may be shifted to data desc ==
-
-Askalon does provide advanced resources mapping and reservations. These
-features however are limited to the fact that not all underlying
-infrastructures allow for a resource reservation scheme as offered by ASKALON.
-GridNexus uses the OGSA-DAI (Open Grid Services Architecture Data Access and
-Integration) services to access Globus grid data resources. It uses
-web/grid-service clients in order to interact with the computations wrapped by
-those services.
-
-The data management services of Trident supports local storage as well as the
-remote data storage services such as the Amazon S3
-(\textit{http://aws.amazon.com/s3}) and SQL Server Data Services (SSDS)
-(\textit{http://microsoft.com/sql/dataservices}) clouds.
-
-Table \ref{tbl:interfacedci} presents a summary of the studied SWEs mechanisms for interfacing DCIs.
-% == WORKFLOW LANGUAGE EXPRESSIVENESS ==
-\section{Workflow Language Expressiveness} \label{sec:lang}
-A workflow language is the medium of expression of the task a workflow is
-supposed to accomplish. A workflow language acts as an interface between the
-workflow enactor and an end-user. Apart from correctly expressing the flow of
-execution and data, a workflow language must be able to link with the
-application code in a transparent way.
-
-Historically, low-level, general-purpose scripting languages have been used for
-workflow descriptions. For example, batch files, UNIX shell scripts and
-script-based languages like perl, python \cite{begeman-belikov-etal:2010} etc.
-There also exist some domain specific scripting languages such as bioperl
-(\textit{www.bioperl.org}). However, these languages are less suitable for
-expression of high-level, sophisticated workflow constructs such as iterations,
-while at the same time dealing with parallelism and array processing. They also
-fall short of providing implicit parallelism which must be defined explicitly by
-the user designing workflows. Description of remote data is another important
-aspect that these languages are limited in expressing. Thus, in modern SWEs, it
-is required to have a higher-level, specialized and dedicated workflow
-language.
-
-Applications and in-silico scientific experiments exhibit diverse patterns of
-data and/or control flow. Apart from this, operations and combinations of data
-are required to correctly and efficiently express a scientific workflow. This
-has led to a requirement of highly expressive workflow languages. Most tasks
-could be accomplished by simplified and minimal constructs in a language,
-however, they lead to unintuitive and inefficient workflow design. A
-sophisticated language not only increases the readability and maintainability
-of a workflow, it also increases the chances of its efficient deployment and
-execution in the target computational environment.
-
-Description of data and dataflow in a workflow such that the application code
-is invoked correctly and in accordance with the experiment has been a
-challenging task for large experiments. This is partly because of the large
-amount and different type of parallelization opportunities and partly because
-of the way the components of an application deal with the data.
-
-A workflow language must be able to express the invocation that is required for
-tools to deploy the application components. For example, an experiment may be a
-combination of remotely executable services and deployable legacy application
-binaries. Along with the application binaries, a workflow language must be
-capable of handling the different expressions for different kinds of data
-transfer methods and protocols. A common way of invoking legacy applications is
-through a traditional Command Line Interface. It is thus desirable to have a
-way to wrap such command lines in order to provide a suitable interface to such
-a legacy
-application~\cite{rojasbalderrama-montagnat-etal:2010,maheshwari:2007}.
-
-%
-\begin{table}
-\begin{center}
-% use packages: array
-\begin{tabular}{|l|l|l|l|}
-\hline
-\textbf{SWE/Language} &\textbf{Control Structures}& \textbf{Array Programming} & \textbf{Iterators}\\
-\hline
-Taverna/Scufl2 &Conditionals and Loops& \checkmark & \checkmark \\\hline
-+/BPEL &Conditionals and Loops& -- & -- \\\hline
-Triana/- &Conditionals and Loops& \checkmark & -- \\\hline
-MOTEUR/Scufl &Conditionals& -- & \checkmark \\\hline
-Kepler/MoML &Conditionals and Loops & \checkmark & limited \\\hline
-GridNexus/GXPL &Conditionals and Loops& -- & -- \\\hline
-Swift/SwiftScript & Conditionals and Loops & 1-dim & limited \\\hline
-Askalon/AGWL & Conditionals and Loops & -- & -- \\\hline
-P-GRADE/- & Conditionals & -- & \checkmark \\\hline
-Pegasus/DAX & -- & \checkmark & -- \\\hline
-Galaxy/- & -- & -- & -- \\\hline
-MOTEUR2/Gwendia~ & Conditionals and Loops & \checkmark & \checkmark \\\hline
-\end{tabular}
-%\caption{A summary of SWEs Expressiveness Capabilities}
-\label{tbl:expressive}
-\end{center}
-\end{table}
-
-\paragraph{Dataflow and Functional Languages} Functional and Dataflow languages
-and their derivatives have been a viable choice for workflow languages
-paradigm. In the dataflow paradigm, a workflow processor will `fire' when the
-data sufficient for its firing is available. Dataflow programming paradigm
-particularly suites the workflow approach of programming. There are many
-research works advocating a dataflow oriented approach to solve problems
-involving a flow of data among computing components, predominantly in a
-Directed Acyclic Graph form \cite{dataflow,lee-messerschmitt:1987}. Early
-research on dataflow languages as visual language~\cite{sutherland:1966}
-reinforce the notion of the suitability of dataflow paradigm for workflows.
-
-Functional programming complement the dataflow paradigm by providing
-appropriate expression and semantics to a dataflow. Several features of
-functional programming are shared by the dataflow programming approach. One of
-the key hinderance of functional programming is the `side-effects' produced
-when the values of variables are mutated more than once during a single
-execution. Dataflow and functional programming approach converge when these
-`side-effects' are eliminated and a functional program can express pure
-dataflow under such conditions. 
-% Modify and write ==> Functional languages, such as SCALA and Haskell, offer
-% an implicit way of exposing parallelism, making it easier to identify data
-% dependencies so that the runtime system can implicitly schedule independent
-% work (work that respects the dependency graph).
-In this section we study the features of contemporary workflow languages that
-make them expressive of representing complex data and control flow situations
-and enable them to map these expression onto the underlying computing
-infrastructure.
-
-Pegasus uses DAX which is an XML-based language to describe DAGs. Applications
-using Pegasus use higher level languages such as Python or Java to describe
-workflows and then map them onto DAX specification. Such transformations face
-two issues: 1. DAGs are simple to transform but can get complicated if
-hierarchical DAGs are used. 2. DAGs representations need sophisticated
-translators. All features in low-level languages such as loops are not
-supported by DAGs. This must be taken care of by the translators.
-
-Kepler, GridNexus and Triana support loops, conditionals and sub-workflows as
-advanced constructs of their workflow language. These constructs are manifested
-in the GUI as configurable components that get their mapping in the underlying
-language by a mapper software component.
-
-ASKALON's AGWL (Abstract Grid Workflow Language) provides an abstract XML-based
-representation of dataflows. It does not directly support control flow, however
-control flow can be achieved indirectly using `pseudo datatokens'. It is
-modularized and supports sub-workflows. AGWL provides conditionals, iterators,
-and explicit parallelism. AGWL provides sophisticated workflow constructs
-including sub-workflows, parallel and sequential execution constructs, loops,
-conditionals and iteration strategies.
-
-The GWorkflowDL employed by GWES is an XML-based representation of Workflows
-which are based on the theory of high-level petri nets. This theoreticaly
-well-found language provides basis for high expressiveness to GWES workflows.
-It leads to the expressions of workflows at three different levels of
-abstractions from user-intentions at high-level to the executable services
-instances at the low-level.
-%Swift is dynamic
-%
-%A paragraph saying the most generic language approaches of contemporary
-%workflow environments
-Both Askalon and Pegasus enable users to define abstract workflows to be
-concretized at the later stage of workflow development. This feature
-facilitates users to have the flexibility to generate concrete workflows based
-upon the state and availability of resources at the time of enactment.
-
-The graphical representation is the only workflow language in Knime. It
-provides an interesting feature of metanodes that acts as a self-contained
-subworkflow. A metanode is a processor that act as a container for other
-processors. Thus a branch of a workflow in Knime can be configured into a
-single processor in the form of metanode. A single such processor might be
-enacted multiple times giving a form of loop to a Knime workflow.
-
-Table \ref{tbl:expressive} shows a summary of language expressiveness
-capabilities of the SWEs studied. We consider the control structures:
-conditionals and loops, support for arrays and iterators
-\cite{montagnat-glatard-etal:2006} as the expressive features of SWEs.
-
-\section{Usability} \label{sec:usable}
-A workflow environment influences the end-user comfort and convenience towards
-the goal of efficiently organizing and deploying a scientific experiment
-involving heavy usage of computational resources. However, there is no single
-factor that determines the usability of a SWE. A combination of several
-qualitative and quantitative factors play a role in determining how usable a
-given SWE is.
-
-\begin{table}[ht!]
-\begin{center}
-% use packages: array
-\begin{tabular}{|l|l|l|l|}
-\hline
-\textbf{SWE/Language} & \textbf{Workflow Composition} & \textbf{Reusable} & \textbf{Extendable}\\
-\hline
-Taverna/Scufl2 & GUI & \checkmark & \checkmark\\\hline
-+/BPEL & GUI & \checkmark & \checkmark \\\hline
-Triana/- & GUI & \checkmark & \checkmark \\\hline
-MOTEUR/Scufl & GUI & \checkmark & -- \\\hline
-Kepler/MoML & GUI & \checkmark & \checkmark \\\hline
-GridNexus/GXPL & GUI/Text & -- & -- \\\hline
-Swift/SwiftScript & Text & -- & -- \\\hline
-Askalon/AGWL & GUI & \checkmark & \checkmark \\\hline
-P-GRADE/- & Web Portal & \checkmark & -- \\\hline
-Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
-Galaxy/- & Web Portal & -- & -- \\\hline
-\multirow{2}{*}{MOTEUR2/} & GUI/Text & \checkmark & --\\
-Gwendia~/gscript & & & \\\hline
-\end{tabular}
-%\caption{A summary of SWEs Usability Analysis}
-\label{tbl:usability}
-\end{center}
-\end{table}
-
-In this section, the following aspects of grid workflow usability are mainly
-considered: a) Ease and convenience for workflow composition. This largely
-rests on the available composition interfaces such as a GUI. b) Workflow
-reusability. c) Extensibility by being able to be linked to external
-applications or the ease of customizing the existing codebase of the SWE. Since
-legacy applications are not developed keeping workflow enactment in mind, it
-largely depends upon the extendability of the workflow environment to adapt to
-the application's requirements. We discuss these features in more details in
-the rest of this section and summarize them in table \ref{tbl:usability} for
-the SWEs studied.
-
-A visual workflow development environment is most prominent among the workflow
-community. The visual paradigm of workflow development is convenient as it
-provides an intuitive ability to the user to design workflows by visualizing
-the flow of data and the processors involved in the data processing at each
-stage. It also facilitates users to visualize the opportunities to exploit
-parallelism in the graph.
-
-Another dimension of usability is from the point of view of the computational
-infrastructure and its middleware. A SWE must optimally expose and make
-available all the features of the underlying middleware components.
-
-On the other side of the workflow development paradigm are the script-based
-workflow development capability. A script-based workflow development has its
-own advantages and can lead to the development of highly expressive, compact
-and portable workflows rapidly in the hands of expert or well-experienced
-end-users. An ideal situation is a workflow manager allowing both visual as
-well as scripting development of scientific workflows.
-
-There are two key aspects to highly usable SWE, expressibility and simplicity.
-From the end-user point of view, these are often contradictory and a trade-off
-is required to attain a balance between the degree to which each aspect is
-contained and come in the way of user comfort. User comfort, time to design and
-construct a workflow are of particular importance in the case of SWEs as the
-users are not necessarily technical experts.
-
-This section explores the above mentioned aspects and features seen in the
-SWEs.
-
-The above mentioned aspects also play a role in workflow reusability.
-Scientific procedures are recurring in nature and offer opportunities to reuse
-fully or partially , the existing workflows~\cite{xiaorong-madey:2007}. In
-order to be able to reuse an existing workflow, it must be easily available and
-clearly described as to what are its functions.
-
-Furthermore, other operational factors contribute significantly to a SWE's
-overall usability. An example is as reported by Ramakrishnan et. al. in
-\cite{ramakrishnan-nurmi-etal:2009} wherein factors such as virtual
-reservations, coordinated use of resources and fault tolerance combined
-together considerably enhance workflow usability. Among these, the fault
-tolerance mechanisms are the most challenging to implement and realize at the
-workflow level. This is owing to the fact that workflows interacting with many
-diverse entities have a large number of, and unforeseen fault conditions. A
-small number of SWEs provide some kind of fault tolerance mechanisms. For
-instance MOTEUR provides a `retry' mechanism on job fault while SwiftScript and
-BPEL provide an exception handling mechanism in order to notify users of
-occurrence of known fault conditions.
-
-Some interesting efforts have been made in the past to increase usability of
-grid environments for a particular domain users such as medical imaging
-\cite{olabarriaga-deboer-etal:2006,olabarriaga-glatard-etal:2008}. In the case
-of SWEs, specific usability issues have been addressed relatively late.
-Distributed workflow repositories \cite{stoyanovich-taskar-etal:2010} have been
-especially popular recently. These repositories play an important role in
-making a large number of scientific workflows available and hence reusable. The
-\textit{myexperiment.org} \cite{deroure-goble-etal:2008} is one such popular
-repository of Taverna workflows available publicly. Other examples of such
-repositories are the workflow libraries maintained as a part of the Kepler
-project (library.kepler-project.org/kepler/) and another one with the Vistrails
-project \cite{callahan-freire-etal:2006}. Kepler also integrates with DataStaR
-\footnote{https://wiki.duraspace.org/display/FEDORACREATE/DataStaR+Use+Case} to
-support data sharing among researchers during the research process, and to
-promote publishing or archiving data and metadata to discipline-specific data
-centers. The workflows hosted in these repositories are tagged and categorized
-based upon their features and functionalities. The ``myexperiment" repository
-is especially notable for its adoption of social web paradigm where users not
-only find and reuse the workflows but can actively collaborate on scientific
-experiments over the web.
-
-% -------------------- Fit here as not main content ------------------------
-
-Business Process Execution Language (BPEL) is a workflow
-language pioneered by IBM Corporation and used predominantly to define business
-workflows. It is now a broad standard for business workflow applications. BPEL
-is an open language and commercial enactors are available that provide
-enactment functionality for the BPEL workflows. There has been attempts to use
-BPEL or its extended version in the scientific applications
-\cite{bpel}. However, the language does not suit well to the
-data-intensive, grid workflows and has a limited adoption in the scientific
-community
-\cite{tan-missier-etal:2010,barga-gannon:2007,scherp-hoing-etal:2009,emmerich-butchart-etal:2005}.
-Apart from that, a lack of an open-source workflow engine with graphical
-interface also contributes to its limited acceptance in academic research.
-However, BPEL has been famously unsuitable for data intensive scientific
-applications  BPEL lacks different user-defined strategies to effect multiple
-invocations of each processor according to the application semantics. BPEL also
-lacks data streams and implicit parallelism forcing a user to explicitly define
-parallelism using one of its parallel processing constructs such as
-\texttt{foreach}.
-%
-\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service
-(GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First
-institute. It implements a dynamic workflow composition and enactment model
-based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}.
-GWES enables key workflow development and deployment lifecycle phases including
-visual composition, data-staging, Grid middleware interfacing and execution
-monitoring. Currently, GWES is intensively being used in the field of Medical
-Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany.
-It has been successfully interfaced with the German D-Grid infrastructure.
-%
-
-The GridNexus \cite{GridNexus} SWE, developed at the
-University of North Carolina, Wilmington, facilitates a graphical composition
-and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
-Ptolemy II framework. GridNexus workflows are expressed in an XML-based
-scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
-customizable and extensible scripting language allowing users to write their
-own workflow processor types. GridNexus workflows are grid oriented and can be
-deployed on grid resources accessed through a persistent WS-RF (Web Services
-Resource Framework) interface.
-
-%
-Bioflow \cite{bioflow} is a query based workflow language.
-It is motivated by the fact that bioinformatics applications are query based
-applications that benefit from the complex database queries or saved database
-`views'. A typical Bioflow workflow pipelines a sequence of filter queries in
-order to get the desired results. Bioflow is mainly utilized in the area of
-bioinformatics and querying genome databases. Additionally, a Bioflow workflow
-provides mechanisms for data manipulation and updation in the existing
-databases. It also includes advanced features such as parallel processing of a
-set of mutually independent database queries. The execution model is based on
-the popular database query language SQL (Structured Query Language). Processes
-can be defined in Bioflow which are processor counterparts in workflow paradigm
-and similar to stored-procedures in SQL jargon.
-
-Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical
-representation of bioflow workflows. A Vizbuilder workflow links to an
-underlying bioflow engine and provides an interface to submit the bioflow
-workflows for execution.
-
-\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative
-workflow language that is inspired by the functional programming paradigm. A
-Martlet workflow conveniently enables distributed parameter sweep applications
-wherein the language intuitively defines the computation and a matching with
-the distributed resources can be carried out at workflow execution time. List
-processing applications where a sequence of application components work over an
-array of data in a Multi-data multi-instruction style setup is well-suited to
-the Martlet's model of computation.
-
-%
-\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining
-system for data intensive applications developed at the University of Konstanz.
-It is a java-based platform with a Graphical Interface that is based on eclipse
-(\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists
-of processors belonging to categories such as readers, manipulators, learners,
-predictors, etc. Each processor has a configuration dialog associated with it
-that allows user to set its parameters. Processors are connected through data
-and/or control connections. A highlight of knime is support for shared- and
-distributed-memory workflow enactment and support for parallel workflow loops
-and subworkflows. The loops in Knime manifest themselves in the form of
-subworkflows (called metanodes in Knime) by a repeated enactment of the same
-subworkflow.
-%
-\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based
-script oriented workflow language used to encode parallel pipelines. StarFlow
-handles data as well as control flow using a combination of static, dynamic and
-user annotations analysis. A parallelization engine uses data and control
-dependencies in a workflow in order to build a parallel profile of the
-workflow. StarFlow has been shown to be successfully interface with the SunGrid
-\cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
-The key feature of StarFlow is its capability to execute workflows to maximum
-possible parallelism. While StarFlow provides a script-based compact and
-semiautomatic workflow composition environment, currently it lacks a
-sophisticated GUI for workflow composition. 
-% ---------------------------------------------------------------------------
-
-\section{Conclusions} \label{sec:sum}
+\section{Summary and Conclusions} \label{sec:sum}
 % \begin{sidewaysfigure}

### 2014-06-02 text/concepts.tex
refactor

\section{Conceptual Overview}
In this section we present a conceptual overview of the subjects of this
study--Workflows and Scientific Workflow Environments.

\subsection{Workflow}
Figure~\ref{fig:wf} shows a typical workflow in the form of a Directed Acyclic
Graph (DAG). Execution typically begins with the input data being provided to
the first task and ends at the results being delivered at the end of the last
task. Tasks are represented by boxes and data/control flow between the task is
represented as arrows. Note that in cases where tasks are connected with pure
control flow, it can be simulated by passing token data between tasks.

For the purpose of this text, we define a workflow as a specification of
computation with at least two discrete tasks either forming distinct
computational stages or running for two or more distinct datasets as
parameters. Consequently, a one-off task or running a task for a single copy of
data is \emph{not} a workflow.

For the rest of this paper, we will use the above definition of a workflow.
%

\begin{figure}[htb]
\begin{center}
\includegraphics[width=4cm]{figures/workflow}
\caption{A typical workflow depicted as a Directed Acyclic Graph (DAG) of tasks coordinated by data or control flow and illustrating data and task parallelism}
\label{fig:wf}
\end{center}
\end{figure}
%

\subsection{Architecture}
Figure~\ref{fig:wfms} shows a high-level architecture of a generic SWE. It shows key
components and their interaction in an SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow engine} using commands and messages on the inside while with
user on the outside. The \textit{workflow language} becomes an intermediate
representation for communication between user interface and the
\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
and a high-level scheduler for the processors and interfaces with the
underlying \textit{distributed resources}. It resolves data dependencies and
optimizes the flow at runtime and monitors the workflow progress and collects
the intermediate and final results.

%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing
infrastructure such as clusters, supercomputers, clouds, and grids. Such
environments are in use today for production level work in different scientific
domains worldwide. They do the basic task of linking inter-related processors
of a multi-staged scientific computation. However, they differ in the features
they provide based upon the scientific domain or user community they offer
service for, and the computational model they follow. Often, one of the main
reasons for such difference is the specialized components and requirements for
a particular domain. Owing to this fact, a given workflow system dominates a
particular area of science (eg. astronomy, biology, etc.). Nonetheless, there
exist a plethora of SWEs suitable for general purpose workflow style
computations.

DAGs are by far, the most popular abstraction of the way workflows are
perceived and visualized. The MA DAG~\cite{caron-desprez:2005}, engine is one
example of a DAG generator engine. MA DAG has evolved from a regular DAG
generator to a more abstract DAG generator laced with control structures. MA
DAG benefits from the workflow scheduling mechanisms included in the
middleware. Due to the impossibility to produce complete DAGs owing to the
presence of control structures such as conditionals and loops, MA DAG generate
partial DAGs greedily: a sub-DAG, as complete as possible, is produced as soon
as possible. Despite the presence of many different approaches to workflow
systems, new requirements continue to pose workflow challenges. Some of the
current challenges facing SWEs are:

\begin{enumerate}
\item \textbf{Multisite execution.} Interfacing to existing computation and
storate infrastructure is not trivial, not to mention the new and
unforeseen access and compute modes. SWEs are yet to formulate a
generalized interface towards the major platforms such that they will
run on them out of the box.

\item \textbf{Manysite execution.} Even if a SWE has interfaces built for
more than one infrastructure, it is hard to optimally schedule large
workflows to many remote sites simultaneously. The problem of limited
control over remote sites and distributing interdependent data and
tasks makes it non-trivial to efficiently load-balance work among them.

\item \textbf{Expressivity.} Workflow languages have evolved over the years
with reasonable expressivity being supported. However, they are often
compared with those of the traditional programming languages. As of
current state of the art, the workflow languages expressivity still
have to catch up to the richness of programming languages across
different generations such as C and python.  The community is yet to
realize a truly generic workflow language that can express a wide
variety of flow patterns encountered in scientific domains.

\item \textbf{Interoperability.} Interoperable workflows will benefit to a
broad audience of users. Many efforts at this are made in the recent
past, such as by the consurtium projects and by individual players.
However, a truly interoperable set of standards for workflows are yet
to emerge.
\end{enumerate}

### 2014-06-02 text/data.tex
refactor

% == DATA MANAGEMENT ==
\section{Data Management and Description} \label{sec:data}
Many aspects of data handling present significant diversity among the workflow
managers. Data management approaches for scientific workflows can be classified
as resource and language level. Resource level strategies include data
management at the application level of abstraction where the data is required
to be distributed, available and securely remote-accessible. It also determines
how the data will be delivered to a resource that handles the processing and
returning the processed data or results to the next processors in the workflow.
Provenance, staging, and managing heterogeneous data structures (files, arrays
etc.) are functional requirements at this level.

Whereas the language level data management involves how the actual data items
are described and handled across a workflow execution. Decisions such as how
complex data structures are supported, transferred and delivered from workflow
input source or from one processor to another are determined at language level.
In most workflow systems, the transfer of data is achieved through `data-ports'
attached to the processors. These ports act as channels to collect and transfer
data across a workflow. The association of data description with the workflow
determines whether the data is loosely or strongly coupled with a workflow. A
loosely coupled workflow with data has advantage over the strongly coupled ones
in the sense that the workflow is independent of its data and can be invoked
with a different set of data.

A source of complexity for the workflow manager is the existence of different
formats of data for a single scientific experiment. These include the different
legacy formats of storage as well as the different ways the data is organized
including database, filesystems and remote links. Further, most remote data is
stored at locations secured behind different types of secure communication
protocols involving special access mechanisms.

In the rest of this section, we study the data handling mechanisms of SWEs.
Taverna's reference scheme makes sure the data produced and consumed is in the
form of references. This approach of data management is efficient because: 1)
It takes less time to transport references, and 2) Locality of references
ensures that the data is processed at a location that is local to multiple
processors operating on the same dataset.

In MOTEUR and Kepler workflow environments, data is described as an XML grammar
allowing for nested collections or multi-dimensional arrays of data. Kepler's
COMAD \cite{keplercomad} workflow enactment model is well suited to collection
oriented data like arrays and streams. Similar efforts exist elsewhere to model
the streams for workflow programming
\cite{blower-haines-etal:2005,herath-plale:2010}.

MOTEUR only uses references to data files while Pegasus groups small files into
a single compressed file for transfer thus saving transfer time. It optimizes
the workflow by annotating and reusing the data already available, if
appropriate. Pegasus is able to generate a workflow from a metadata description
of the desired data product with the aid of Artificial Intelligence planning
techniques. Grouping of tasks by Pegasus increases the data usage efficiency
since a group of tasks that require the same set of files for execution are
localized. Pegasus also manages the temporary storage space needed for data
processing at a particular processor location and commands the local Operating
System to allocate the space. A data provider abstraction allows data to be
referenced in the Trident SWE from any external entities in the form of typed
referenced data objects.

A majority of SWE support primitive, MIME based or XSD data types similar to
traditional programming datatypes. Specific support for complex and structured
data types are provided by the SWEs based upon their domain and application
requirements. For instance, Kepler supports a specialized byte data streams and
Bioflow supports a natively developed life-science database engine. Others such
as the SwiftScript support extensible structures such as filesystem mappers and
customized C language style structures.

Remote and secure data access is supported by SWEs by providing secured
protocols based accession to remote data repositories. Third party protocols
such as GridFTP, HTTP over TLS (HTTPS) and gLite LFC (LCG File Catalog) access
are commonly supported.

\begin{table}[ht!]
\begin{center}
\hspace*{-2.2cm}
% use packages: array
\begin{tabular}{|c|c|c|c|c|c|}
\hline
\textbf{SWE /} & \multicolumn{3}{c|}{\textbf{Language Level}} & \multicolumn{2}{c|}{\textbf{Resource Level}}\\
\cline{2-6}
\multirow{2}{*}{\textbf{Language}} & \textbf{Data Coupling} & \textbf{Data Types} & \textbf{Data Structures} & \textbf{Type} & \textbf{Access}\\
& \textbf{with Workflow} & \textbf{Supported} & \textbf{Supported} & & \\
\hline
Taverna/Scufl2&Loose&MIME-types&multi-dim Arrays&Local+Remote&By Reference\\\hline
+/BPEL&Strong&XSD-Based&--&Grids, Clouds& Adhoc\\\hline
%
\multirow{2}{*}{Triana/-}&Strong&XML Schema-&Streams,&Local+&Through \\
&&Types&Java Objects&Remote&Adaptors\\\hline
%
\multirow{2}{*}{MOTEUR/Scufl}&Loose&MIME-&Lists&Local+&Services+\\
&&Types&&Remote&Middleware\\\hline
% 
\multirow{2}{*}{Kepler/}&Strong&Primitive&Collections,&Filesystem&Customized\\
MoML&&Types&User defined&&Actors\\\hline
% 
GridNexus/GXPL&Loose&XSD-Based&--&Grid Data&OGSA-DAI \\\hline
% 
\multirow{2}{*}{Swift/SwiftScript} &Strong&--&1-Dim Arrays,&Shared& POSIX-style\\
&&&Customised&FileSystem&\\\hline
% 
Askalon/AGWL&Loose&MIME-types&Arrays&Data Repository&SQL-based\\\hline
%
\multirow{2}{*}{P-GRADE/-}&Loose&--&--&Grid&Secure\\
&&&&File Catalogs&Protocols\\\hline
%
\multirow{2}{*}{}Pegasus/&Loose&--&--&Distributed&Secure\\
DAX&&&&Filesystems&Protocols\\\hline
% 
Galaxy/-&Loose&--&File Streams&File Systems&--\\\hline
% 
\multirow{2}{*}{MOTEUR2/} &Loose&Integer,String,&multi-dim Arrays&Grid&By\\
Gwendia~ &&Double,URI&&Data&Reference\\\hline
\end{tabular}
%\caption{A summary of SWEs Data Management Mechanisms}
\label{tbl:datadesc}
\end{center}
\end{table}

Table \ref{tbl:datadesc} presents a summary of the data management mechanisms
of the studied SWEs. Language and resource level data management mechanisms are
highlighted. At the language level, data description as coupled with that of
the workflow, data types and structures are considered for the SWE studied. At
the resource level, the major resource types supported and their access
mechanisms are highlighted. We distinguish the remote form of Grid resource
types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those
resource access. Other resource types such as data repositories or file
catalogues are the resource types specific to a given SWE. For instance,
Askalon uses a specifically configured data repository to access data and
application binaries.

%Replication is one of the solution.

### 2014-06-02 text/interface.tex
refactor

% == INTERFACE TO COMPUTING AND DATA INFRASTRUCTURES ==
\section{Interface to Distributed Computing Infrastructures (DCIs)} \label{sec:inter}
Workflow environments are external systems to large scale computing
infrastructures. They require an interface to connect to and use the
infrastructures. Provided the complex and distributed nature of grids, it is
not easy and practical to interact with such an infrastructure directly. Most
distributed systems have a software service layer through which it interacts
with the outside world. These interactions include accepting queries for
information and accepting requests and data for job executions. Such a layer is
typically called the \textit{middleware}. A workflow engine intending to
interact with a grid environment must provide interface to link to such
middleware layer.

Most grid computing systems today are batch oriented execution systems (with a
notable exception of GridRPC \cite{seymour-nakada-etal:2002} based systems). A
batch of ``jobs" are submitted to the system which in turn passes through some
execution queues before actually being scheduled for execution. Such systems
are prone to high execution latencies and unreliability issues. A workflow
manager must take these issues into account while interacting with such
systems.

At the resource management level, most grid computing environments today are
utilizing one or more components of the Globus Toolkit
\cite{foster-kesselman:1997} or the Condor distributed computing software
system~\cite{dagman}.

The Globus Toolkit is a software stack that provides useful functionalities for
grid computing operations. These functionalities are provided in the form of a
software stack that ranges from low-level data management tasks to the higher
level functionalities like authentication and metadata services. Any external
entity accessing the grid computing system must possess proper authentication
and sufficient authority to operate upon the infrastructure. The most popular
secure access system in place for today's grids is the public key
infrastructure (PKI) system through which a secure access to the grid is
guaranteed for a finite time period. A workflow manager must be capable of
handling the security protocols with the underlying execution infrastructure.
The duration of access to the system is also of critical importance as a
typical scientific workflow is bound to run for a considerably long time
period. The Condor software system provides scheduling mechanisms to access the
low-layer grid resources. A grid workflow enactor must provide interface to
access the services of these external scheduling and middleware systems.

\begin{table}
\begin{center}
%\hspace*{-1cm}
% use packages: array
\begin{tabular}{|l|l|l|}
\hline
\textbf{SWE/Language} & \textbf{DCI Interfacing Mechanisms} & \textbf{Currently Interfaced DCIs}\\
\hline
Taverna/Scufl2 & By Extension & None Inherently \\\hline
+/BPEL & Ad-hoc & None Inherently \\\hline
Triana/- & Provides Adapters & P2P Networks, NGS (ngs.org.uk)  \\\hline
MOTEUR/Scufl & Inherent & EGI (www.eu-egi.eu), g5k (www.grid5000.fr)\\\hline
Kepler/MoML & Using Directors & EarthGrid (http://seek.ecoinformatics.org) \\\hline
GridNexus/GXPL & Inherent & Generic Grids \\\hline
Swift/SwiftScript & Inherent & IBM/bluegene/P HPC Clusters \\\hline
Askalon/AGWL & Inherent & Austrian Grid (http://www.austriangrid.at/)\\\hline
P-GRADE/- & Inherent & NGS \\\hline
Pegasus/DAX & Using Condor & OSG (www.opensciencegrid.org) \\\hline
Galaxy/- & Extensions & None Inherently \\\hline
MOTEUR2/Gwendia~ & Inherent & EGI, g5k, HIPernet Cloud \\\hline
\end{tabular}
%\vspace*{1cm}
%\caption{A summary of SWEs DCI Interfacing Mechanisms}
\label{tbl:interfacedci}
\end{center}
\end{table}

Remotely located grid resources are accessed normally over the Internet links.
In such situations, long network latency or outright link failures are common.
Workflow managers are likely to face issues such as no response from the job
submitter or long delays. There must be some mechanism in place to handle such
scenarios. Depending upon the sophistication of the underlying infrastructure a
failover mechanism or logging information could be obtained for failures. Often
times workflow managers have their own failover mechanisms in place.

In the rest of this section, we study the approaches and techniques of SWEs to
interact with the Computing and Data Infrastructures. We also discuss briefly
the mechanisms for the mandatory security requirements.

MOTEUR provides seamless access to local and remote grid data resources by
closely coupling itself with the vBrowser~\cite{olabarriaga-deboer-etal:2006}
virtual resource browser environment. It has a built-in job management
mechanism that includes building job-specification, job submission data staging
and collection of results from the jobmanager. Additionally, MOTEUR2 is
interfaced with the HiPerNet cloud, a virtual platform manager that takes into
account, both computing and network resources
\cite{vicat-blancprimet-roca-etal:2009}.

In Taverna, a processor's invocation sequence can be customized
programmatically to enhance the processor's performance selectively. This
feature can be exploited by effectively interfacing a Taverna processor to Grid
middleware and create jobs and inputs/outputs definitions for that middleware.

In Taverna, a processor's invocation sequence can be customized
programmatically to enhance the processor's performance selectively. This
feature can be exploited by effectively interfacing a Taverna processor to Grid
middleware and create jobs and inputs/outputs definitions for that middleware.

Pegasus interfaces explicitly and transparently with the infrastructure in that
it maps the simplified workflow DAGs to the grid resources through the Condor
resource management service. Pegasus interacts with the available replica
management services in order to optimize the data access cost for the workflow
executions at runtime. It does cluster--group similar jobs together, in order
to ease the load on the scheduler and the dispatching of those jobs to the
worker nodes.

Swift/Karajan and Pegasus interface to a POSIX-style shared file system which
is one crucial aspect to their performance. %== may be shifted to data desc ==

Askalon does provide advanced resources mapping and reservations. These
features however are limited to the fact that not all underlying
infrastructures allow for a resource reservation scheme as offered by ASKALON.
GridNexus uses the OGSA-DAI (Open Grid Services Architecture Data Access and
Integration) services to access Globus grid data resources. It uses
web/grid-service clients in order to interact with the computations wrapped by
those services.

The data management services of Trident supports local storage as well as the
remote data storage services such as the Amazon S3
(\textit{http://aws.amazon.com/s3}) and SQL Server Data Services (SSDS)
(\textit{http://microsoft.com/sql/dataservices}) clouds.

Table \ref{tbl:interfacedci} presents a summary of the studied SWEs mechanisms for interfacing DCIs.

### 2014-06-02 text/lang.tex
refactor

% == WORKFLOW LANGUAGE EXPRESSIVENESS ==
\section{On Workflow Languages} \label{sec:lang}
A workflow language is the medium of expression of the task a workflow is
supposed to accomplish. A workflow language acts as an interface between the
workflow enactor and an end-user. Apart from correctly expressing the flow of
execution and data, a workflow language must be able to link with the
application code in a transparent way.

Historically, low-level, general-purpose scripting languages have been used for
workflow descriptions. For example, batch files, UNIX shell scripts and
script-based languages like perl, python \cite{begeman-belikov-etal:2010} etc.
There also exist some domain specific scripting languages such as bioperl
(\textit{www.bioperl.org}). However, these languages are less suitable for
expression of high-level, sophisticated workflow constructs such as iterations,
while at the same time dealing with parallelism and array processing. They also
fall short of providing implicit parallelism which must be defined explicitly by
the user designing workflows. Description of remote data is another important
aspect that these languages are limited in expressing. Thus, in modern SWEs, it
is required to have a higher-level, specialized and dedicated workflow
language.

Applications and in-silico scientific experiments exhibit diverse patterns of
data and/or control flow. Apart from this, operations and combinations of data
are required to correctly and efficiently express a scientific workflow. This
has led to a requirement of highly expressive workflow languages. Most tasks
could be accomplished by simplified and minimal constructs in a language,
however, they lead to unintuitive and inefficient workflow design. A
sophisticated language not only increases the readability and maintainability
of a workflow, it also increases the chances of its efficient deployment and
execution in the target computational environment.

Description of data and dataflow in a workflow such that the application code
is invoked correctly and in accordance with the experiment has been a
challenging task for large experiments. This is partly because of the large
amount and different type of parallelization opportunities and partly because
of the way the components of an application deal with the data.

A workflow language must be able to express the invocation that is required for
tools to deploy the application components. For example, an experiment may be a
combination of remotely executable services and deployable legacy application
binaries. Along with the application binaries, a workflow language must be
capable of handling the different expressions for different kinds of data
transfer methods and protocols. A common way of invoking legacy applications is
through a traditional Command Line Interface. It is thus desirable to have a
way to wrap such command lines in order to provide a suitable interface to such
a legacy
application~\cite{rojasbalderrama-montagnat-etal:2010,maheshwari:2007}.

%
\begin{table}
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{SWE/Language} &\textbf{Control Structures}& \textbf{Array Programming} & \textbf{Iterators}\\
\hline
Taverna/Scufl2 &Conditionals and Loops& \checkmark & \checkmark \\\hline
+/BPEL &Conditionals and Loops& -- & -- \\\hline
Triana/- &Conditionals and Loops& \checkmark & -- \\\hline
MOTEUR/Scufl &Conditionals& -- & \checkmark \\\hline
Kepler/MoML &Conditionals and Loops & \checkmark & limited \\\hline
GridNexus/GXPL &Conditionals and Loops& -- & -- \\\hline
Swift/SwiftScript & Conditionals and Loops & 1-dim & limited \\\hline
Askalon/AGWL & Conditionals and Loops & -- & -- \\\hline
P-GRADE/- & Conditionals & -- & \checkmark \\\hline
Pegasus/DAX & -- & \checkmark & -- \\\hline
Galaxy/- & -- & -- & -- \\\hline
MOTEUR2/Gwendia~ & Conditionals and Loops & \checkmark & \checkmark \\\hline
\end{tabular}
%\caption{A summary of SWEs Expressiveness Capabilities}
\label{tbl:expressive}
\end{center}
\end{table}

\paragraph{Dataflow and Functional Languages} Functional and Dataflow languages
and their derivatives have been a viable choice for workflow languages
paradigm. In the dataflow paradigm, a workflow processor will `fire' when the
data sufficient for its firing is available. Dataflow programming paradigm
particularly suites the workflow approach of programming. There are many
research works advocating a dataflow oriented approach to solve problems
involving a flow of data among computing components, predominantly in a
Directed Acyclic Graph form \cite{dataflow,lee-messerschmitt:1987}. Early
research on dataflow languages as visual language~\cite{sutherland:1966}
reinforce the notion of the suitability of dataflow paradigm for workflows.

Functional programming complement the dataflow paradigm by providing
appropriate expression and semantics to a dataflow. Several features of
functional programming are shared by the dataflow programming approach. One of
the key hinderance of functional programming is the `side-effects' produced
when the values of variables are mutated more than once during a single
execution. Dataflow and functional programming approach converge when these
`side-effects' are eliminated and a functional program can express pure
dataflow under such conditions. 
% Modify and write ==> Functional languages, such as SCALA and Haskell, offer
% an implicit way of exposing parallelism, making it easier to identify data
% dependencies so that the runtime system can implicitly schedule independent
% work (work that respects the dependency graph).
In this section we study the features of contemporary workflow languages that
make them expressive of representing complex data and control flow situations
and enable them to map these expression onto the underlying computing
infrastructure.

Pegasus uses DAX which is an XML-based language to describe DAGs. Applications
using Pegasus use higher level languages such as Python or Java to describe
workflows and then map them onto DAX specification. Such transformations face
two issues: 1. DAGs are simple to transform but can get complicated if
hierarchical DAGs are used. 2. DAGs representations need sophisticated
translators. All features in low-level languages such as loops are not
supported by DAGs. This must be taken care of by the translators.

Kepler, GridNexus and Triana support loops, conditionals and sub-workflows as
advanced constructs of their workflow language. These constructs are manifested
in the GUI as configurable components that get their mapping in the underlying
language by a mapper software component.

ASKALON's AGWL (Abstract Grid Workflow Language) provides an abstract XML-based
representation of dataflows. It does not directly support control flow, however
control flow can be achieved indirectly using `pseudo datatokens'. It is
modularized and supports sub-workflows. AGWL provides conditionals, iterators,
and explicit parallelism. AGWL provides sophisticated workflow constructs
including sub-workflows, parallel and sequential execution constructs, loops,
conditionals and iteration strategies.

The GWorkflowDL employed by GWES is an XML-based representation of Workflows
which are based on the theory of high-level petri nets. This theoreticaly
well-found language provides basis for high expressiveness to GWES workflows.
It leads to the expressions of workflows at three different levels of
abstractions from user-intentions at high-level to the executable services
instances at the low-level.
%Swift is dynamic
%
%A paragraph saying the most generic language approaches of contemporary
%workflow environments
Both Askalon and Pegasus enable users to define abstract workflows to be
concretized at the later stage of workflow development. This feature
facilitates users to have the flexibility to generate concrete workflows based
upon the state and availability of resources at the time of enactment.

The graphical representation is the only workflow language in Knime. It
provides an interesting feature of metanodes that acts as a self-contained
subworkflow. A metanode is a processor that act as a container for other
processors. Thus a branch of a workflow in Knime can be configured into a
single processor in the form of metanode. A single such processor might be
enacted multiple times giving a form of loop to a Knime workflow.

Table \ref{tbl:expressive} shows a summary of language expressiveness
capabilities of the SWEs studied. We consider the control structures:
conditionals and loops, support for arrays and iterators
\cite{montagnat-glatard-etal:2006} as the expressive features of SWEs.

### 2014-06-02 text/others.tex
refactor

% -------------------- Fit here as not main content ------------------------
\subsection{Other Systems}
In this section we briefly discuss various other systems which are either too
niche or obsoleted.

Business Process Execution Language (BPEL)~\cite{bpel} is a commercil workflow
platform pioneered by IBM Corporation and used predominantly to define business
workflows. It is now a broad standard for business workflow applications. BPEL
is an open language and commercial enactors are available that provide
enactment functionality for the BPEL workflows. There has been attempts to use
BPEL or its extended version in the scientific applications. However, the
language does not suit well to the data-intensive, grid workflows and has a
limited adoption in the scientific
community~\cite{tan-missier-etal:2010,barga-gannon:2007,scherp-hoing-etal:2009}.
Apart from that, a lack of an open-source workflow engine with graphical
interface also contributes to its limited acceptance in academic research.
However, BPEL has been famously unsuitable for data intensive scientific
applications  BPEL lacks different user-defined strategies to effect multiple
invocations of each processor according to the application semantics. BPEL also
lacks data streams and implicit parallelism forcing a user to explicitly define
parallelism using one of its parallel processing constructs such as
\texttt{foreach}.

%
The Grid Workflow Execution Service (GWES)~\cite{Hoheisel2007} is a SWE
developed by the Fraunhofer-First institute. It implements a dynamic workflow
composition and enactment model based on the Grid Workflow Description Language
(GWorkflowDL)~\cite{Alt2006}.  GWES enables key workflow development and
deployment lifecycle phases including visual composition, data-staging, Grid
middleware interfacing and execution monitoring. Currently, GWES is intensively
being used in the field of Medical Image Processing especially at the
Charit\'{e} Hospital at Berlin in Germany.  It has been successfully interfaced
with the German D-Grid infrastructure.

%
The GridNexus \cite{GridNexus} SWE, developed at the University of North
Carolina, Wilmington, facilitates a graphical composition and execution of
DAG-based workflows. Like Kepler, GridNexus is also based on Ptolemy II
framework. GridNexus workflows are expressed in an XML-based scripting language
called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a customizable and extensible
scripting language allowing users to write their own workflow processor types.
GridNexus workflows are grid oriented and can be deployed on grid resources
accessed through a persistent WS-RF (Web Services Resource Framework)
interface.

%
Bioflow~\cite{bioflow} is a query based workflow language.
It is motivated by the fact that bioinformatics applications are query based
applications that benefit from the complex database queries or saved database
`views'. A typical Bioflow workflow pipelines a sequence of filter queries in
order to get the desired results. Bioflow is mainly utilized in the area of
bioinformatics and querying genome databases. Additionally, a Bioflow workflow
provides mechanisms for data manipulation and updation in the existing
databases. It also includes advanced features such as parallel processing of a
set of mutually independent database queries. The execution model is based on
the popular database query language SQL (Structured Query Language). Processes
can be defined in Bioflow which are processor counterparts in workflow paradigm
and similar to stored-procedures in SQL jargon.

%
Vizbuilder~\cite{vizbuilder} is a graphical toolkit to generate the graphical
representation of bioflow workflows. A Vizbuilder workflow links to an
underlying bioflow engine and provides an interface to submit the bioflow
workflows for execution.

%
Martlet~\cite{martlet} is a text-based, declarative
workflow language that is inspired by the functional programming paradigm. A
Martlet workflow conveniently enables distributed parameter sweep applications
wherein the language intuitively defines the computation and a matching with
the distributed resources can be carried out at workflow execution time. List
processing applications where a sequence of application components work over an
array of data in a Multi-data multi-instruction style setup is well-suited to
the Martlet's model of computation.

%
Knime~\cite{knime} is a distributed, parallel pipelining
system for data intensive applications developed at the University of Konstanz.
It is a java-based platform with a Graphical Interface that is based on eclipse
(\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists
of processors belonging to categories such as readers, manipulators, learners,
predictors, etc. Each processor has a configuration dialog associated with it
that allows user to set its parameters. Processors are connected through data
and/or control connections. A highlight of knime is support for shared- and
distributed-memory workflow enactment and support for parallel workflow loops
and subworkflows. The loops in Knime manifest themselves in the form of
subworkflows (called metanodes in Knime) by a repeated enactment of the same
subworkflow.

%
StarFlow~\cite{angelino-yamins-etal:2010} is a Python-based
script oriented workflow language used to encode parallel pipelines. StarFlow
handles data as well as control flow using a combination of static, dynamic and
user annotations analysis. A parallelization engine uses data and control
dependencies in a workflow in order to build a parallel profile of the
workflow. StarFlow has been shown to be successfully interface with the SunGrid
\cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
The key feature of StarFlow is its capability to execute workflows to maximum
possible parallelism. While StarFlow provides a script-based compact and
semiautomatic workflow composition environment, currently it lacks a
sophisticated GUI for workflow composition.

% ---------------------------------------------------------------------------

### 2014-06-02 text/scale.tex
refactor

\section{Scalability and Optimization} \label{sec:scal}
A simple definition of scalability in the context of SWE is that the
performance of the workflow engine increases linearly with the increasing
number of similar tasks. Depending upon the lag or increase in the linearity we
say a workflow enactor scales up or down.  In this section, we focus on
the scalability aspects of the workflow runtime. We do not consider the
specific scalability issues related to a distributed infrastructure such as
network latencies and protocol related overheads. However, there have been efforts to
thwart such issues at the workflow level by combining workflow enactment
techniques with those of distributed computing.  One such example is the
map/reduce framework \cite{mapreduce}.  Recently, the map/reduce framework,
that aims at performing distributed processing on large datasets, has gained a
significant popularity in workflow community. It is based on the master/worker
metaphor of distributed processing.  It employs the commonly used \textit{map}
and \textit{reduce} functions from the functional programming languages. The
map function chops a large chunk of data into smaller pieces and distributes it
to \textit{workers} which in turn can also chop it to further smaller datasets
forming a tree-like hierarchy. The worker process the data and return the
results back to their parents. The reduce step collects the results from the
workers and combines them to obtain the final result. One of the main reason
for the popularity of the map/reduce framework for workflows is the recent
attention to collections based datasets and processing. There exists
implementations of the map/reduce framework written in different languages with
enhancements and support services such as a sophisticated query service. Hadoop
(\textit{http://hadoop.apache.org/}) is one such implementation which has
already been interfaced by the Kepler SWE~\cite{wang-crawl-etal:2009}.

Large-scale applications with thousands of tasks often pose memory and disk
management difficulties leading to scalability issues.  However, the
performance of a workflow system is dependent on several factors including the
type of tasks and approach for submitting tasks for execution.  There is not
much research done towards the enactor scalability baring a couple of studies
carried out by C. Stratan et. al. \cite{stratan-iosup-etal:2008} and Deelman
et. al. \cite{callaghan-deelman-etal:2010}.  One way to compare the scalability
of different workflow environments is to experiment with a similar set of
workflows on a given system. In addition, different SWEs interface to different
DCIs. This is a time consuming process.  In the current study we do not
directly compare the workflow system scalability but we study the similarities
and differences in the techniques employed by workflow engines in order to
achieve scalability.

The software architecture of the system plays a key role in determining the
scalability for the workflow applications exhibiting a large number of tasks.
An architecture that takes into account the properties of the underlying
infrastructure could scale well in the face of growing number of workflow
tasks.

In the context of workflow runtime, scalability can be further classified in
terms of data motion scalability. It can be expressed as the rate of increase
in number of data movement quantity with respect to the increase in the elapsed
time.

While the computational scalability of a large scale experiment depends upon
the underlying computing infrastructure, at the workflow enactor level, it
becomes important to be able to prepare and dispatch as many processor
instances as possible without loss in performance. Failing to do so for a
workflow engine can become a performance bottleneck and hamper the optimal
utilization of an available computing infrastructure.

Most workflow systems exploit the parallelism at three-levels. First, at the
task-level wherein the independent branches of a workflow graph are enacted
concurrently. Second, at the data-level, where a same processor is instantiated
multiple times based upon the repeated operations on different datasets. Third,
the low-level parallelism that is encoded within the application components
independent of the workflow. At this level, the workflow enactor may
need to make a matching based upon the application type. For instance, a
workflow enactor may schedule the GPU-based processors on the resources that
indicate as possessing such capabilities.

In this section we analyze the scalability and runtime optimization measures
and techniques employed by SWEs. Illustrated below are some of the techniques
used to achieve scalability in various SWEs.

Taverna version 2.x introduced new measures in order to be able to submit a
large number of jobs simultaneously. It is able to dispatch large number of
concurrent instances of processors. An ability to implement
superpipelined~\cite{hwang:1992} execution of workflow tasks leads to a maximum
expression of parallelism~\cite{pipeline}.

Within the Taverna engine, all the data involved in the execution is managed by
a data manager through an external database, and is addressed by the engine
only by \textit{reference}. This facilitates the management of data-intensive
workflows in a scalable way, as high volume data is only transferred into the
engine space when needed by processors that are unable to resolve external data
references on their own.

Taverna, Swift-Karajan and Pegasus all provide a means to limit the maximum
number of processor instances that can run concurrently through a builtin
parameter. In Taverna, this can be configured at a workflow level while in
Pegasus, this can be configured at the Condor \cite{thain-tannenbaum-etal:2005}
scheduler level. In Swift-Karajan system this parameter is regulated
automatically as part of a ``throttling parallel foreach" construct. A similar
configurable parameter is provided by the Knime thread manager in order to
limit the size of the thread-pool employed to run a workflow.

Kepler optimizes the workflow execution through parallel and pipelined
execution. In addition to these, Kepler also provides dataflow analysis
optimization that involves data bypassing the processors in a workflow that
does not make use of that data. This is detected using a builtin Kepler
mechanism called ``static type propagation".

Another alternative is to optimize the enactment at runtime. This approach has
advantages including a better information about the actual data available. For
instance, Askalon does runtime branch prediction, parallel loop unrolling and
loop sequentialization.

The Pegasus workflow manager introduced subworkflows because of scalability
issues at the localhost for a large number of nodes in a DAG. A hierarchical
representation helps it scale at the local planning level as well as at the
execution level on distributed resources.

Static planners in Pegasus translate a workflow graph into a location specific
DAGMan input file, adding stages for data staging, intersite transfer and data
registration. These planners can reduce processing of tasks for files that
already exist, select sites for jobs, and cluster jobs based on various
criteria. Pegasus performs graph transformation with the knowledge of the whole
workflow graph. A similar strategy is grouping tasks by application-level
wrappers written in higher-level programming languages such as bash or Python.

In some systems the structure of a workflow is constructed and expanded
dynamically as opposed to static planning. A static workflow graph allows
optimizations based on the global state of the workflow, whereas a dynamic
workflow structure allows adaptations to changing environments.  Depending upon
the execution environment and access conditions to the available resources,
each approach can result in the overall optimization of a given application.

Knime has a different execution strategies for shared memory Symmetric
Multiprocessing Systems (SMP) and distributed memory systems. In the case of
SMP, it uses the message passing library or MPI to distribute the independent
parts of a workflow to available processors. This approach results in efficient
use of available processors. However, for a large scale, the global memory
becomes a bottleneck hindering the scalability.

In the case of distributed memory and independent computational nodes connected
with highspeed network or the internet, Knime employs a Master/Worker metaphor
of computing. A Knime installation with GUI acts as a master node that
coordinates the bare Knime installations on the worker nodes. These worker
nodes are then sent the portions of a workflow by the master node. The master
node keeps track of the participating nodes by their unique IP addresses. At
the end of each execution, the master node collects the results from the worker
nodes.

\begin{table}[ht!]
%TODO: Caption is not printing well, fix
%\caption{A summary of SWEs Scalability and optimization Mechanisms}
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|l|l|}
\hline
\textbf{SWE} / & \multicolumn{3}{c|}{\textbf{Parallelism}} & \multicolumn{2}{c|}{\textbf{Optimization}}\\
\cline{2-6}
\textbf{Language} & \textbf{data} & \textbf{task} & \textbf{pipeline} & \textbf{static} & \textbf{runtime}\\
\hline
Taverna/Scufl2 & \checkmark & \checkmark & \checkmark & -- & \checkmark \\
\hline
Triana/- & \checkmark & \checkmark & \checkmark & -- & \checkmark \\
\hline
MOTEUR/Scufl & \checkmark & \checkmark & \checkmark &-- &\checkmark \\
\hline
Kepler/MoML & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
\hline
GridNexus/GXPL & -- & \checkmark & -- &-- &\checkmark \\
\hline
Swift/SwiftScript & \checkmark & \checkmark & \checkmark & \checkmark &\checkmark \\
\hline
Askalon/AGWL &--& \checkmark & -- & \checkmark & \checkmark \\
\hline
P-GRADE/- &\checkmark &\checkmark &\checkmark &--&--\\
\hline
Pegasus/DAX &\checkmark &\checkmark &\checkmark &\checkmark &\checkmark \\
\hline
Galaxy/- &-- &-- &\checkmark &-- &-- \\
\hline
MOTEUR2/Gwendia~ & \checkmark &\checkmark & \checkmark &-- &\checkmark \\
\hline
\end{tabular}
\label{tbl:scalability}
\end{center}
\end{table}

Shown in Table \ref{tbl:scalability} is a summary of the Scalability and
optimization mechanisms of the studied SWEs. The scalability features
considered are through three levels of parallelism in the form of data,
task and pipeline. The optimization features considered are the presence of
mechanisms by which static and runtime optimizations are achieved.

### 2014-06-02 text/usability.tex
refactor

\section{Usability} \label{sec:usable}
A workflow environment influences the end-user comfort and convenience towards
the goal of efficiently organizing and deploying a scientific experiment
involving heavy usage of computational resources. However, there is no single
factor that determines the usability of a SWE. A combination of several
qualitative and quantitative factors play a role in determining how usable a
given SWE is.

\begin{table}[ht!]
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{SWE/Language} & \textbf{Workflow Composition} & \textbf{Reusable} & \textbf{Extendable}\\
\hline
Taverna/Scufl2 & GUI & \checkmark & \checkmark\\\hline
+/BPEL & GUI & \checkmark & \checkmark \\\hline
Triana/- & GUI & \checkmark & \checkmark \\\hline
MOTEUR/Scufl & GUI & \checkmark & -- \\\hline
Kepler/MoML & GUI & \checkmark & \checkmark \\\hline
GridNexus/GXPL & GUI/Text & -- & -- \\\hline
Swift/SwiftScript & Text & -- & -- \\\hline
Askalon/AGWL & GUI & \checkmark & \checkmark \\\hline
P-GRADE/- & Web Portal & \checkmark & -- \\\hline
Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
Galaxy/- & Web Portal & -- & -- \\\hline
\multirow{2}{*}{MOTEUR2/} & GUI/Text & \checkmark & --\\
Gwendia~/gscript & & & \\\hline
\end{tabular}
%\caption{A summary of SWEs Usability Analysis}
\label{tbl:usability}
\end{center}
\end{table}

In this section, the following aspects of grid workflow usability are mainly
considered: a) Ease and convenience for workflow composition. This largely
rests on the available composition interfaces such as a GUI. b) Workflow
reusability. c) Extensibility by being able to be linked to external
applications or the ease of customizing the existing codebase of the SWE. Since
legacy applications are not developed keeping workflow enactment in mind, it
largely depends upon the extendability of the workflow environment to adapt to
the application's requirements. We discuss these features in more details in
the rest of this section and summarize them in table \ref{tbl:usability} for
the SWEs studied.

A visual workflow development environment is most prominent among the workflow
community. The visual paradigm of workflow development is convenient as it
provides an intuitive ability to the user to design workflows by visualizing
the flow of data and the processors involved in the data processing at each
stage. It also facilitates users to visualize the opportunities to exploit
parallelism in the graph.

Another dimension of usability is from the point of view of the computational
infrastructure and its middleware. A SWE must optimally expose and make
available all the features of the underlying middleware components.

On the other side of the workflow development paradigm are the script-based
workflow development capability. A script-based workflow development has its
own advantages and can lead to the development of highly expressive, compact
and portable workflows rapidly in the hands of expert or well-experienced
end-users. An ideal situation is a workflow manager allowing both visual as
well as scripting development of scientific workflows.

There are two key aspects to highly usable SWE, expressibility and simplicity.
From the end-user point of view, these are often contradictory and a trade-off
is required to attain a balance between the degree to which each aspect is
contained and come in the way of user comfort. User comfort, time to design and
construct a workflow are of particular importance in the case of SWEs as the
users are not necessarily technical experts.

This section explores the above mentioned aspects and features seen in the
SWEs.

The above mentioned aspects also play a role in workflow reusability.
Scientific procedures are recurring in nature and offer opportunities to reuse
fully or partially , the existing workflows~\cite{xiaorong-madey:2007}. In
order to be able to reuse an existing workflow, it must be easily available and
clearly described as to what are its functions.

Furthermore, other operational factors contribute significantly to a SWE's
overall usability. An example is as reported by Ramakrishnan et. al. in
\cite{ramakrishnan-nurmi-etal:2009} wherein factors such as virtual
reservations, coordinated use of resources and fault tolerance combined
together considerably enhance workflow usability. Among these, the fault
tolerance mechanisms are the most challenging to implement and realize at the
workflow level. This is owing to the fact that workflows interacting with many
diverse entities have a large number of, and unforeseen fault conditions. A
small number of SWEs provide some kind of fault tolerance mechanisms. For
instance MOTEUR provides a `retry' mechanism on job fault while SwiftScript and
BPEL provide an exception handling mechanism in order to notify users of
occurrence of known fault conditions.

Some interesting efforts have been made in the past to increase usability of
grid environments for a particular domain users such as medical imaging
\cite{olabarriaga-deboer-etal:2006,olabarriaga-glatard-etal:2008}. In the case
of SWEs, specific usability issues have been addressed relatively late.
Distributed workflow repositories \cite{stoyanovich-taskar-etal:2010} have been
especially popular recently. These repositories play an important role in
making a large number of scientific workflows available and hence reusable. The
\textit{myexperiment.org} \cite{deroure-goble-etal:2008} is one such popular
repository of Taverna workflows available publicly. Other examples of such
repositories are the workflow libraries maintained as a part of the Kepler
project (library.kepler-project.org/kepler/) and another one with the Vistrails
project \cite{callahan-freire-etal:2006}. Kepler also integrates with DataStaR
\footnote{https://wiki.duraspace.org/display/FEDORACREATE/DataStaR+Use+Case} to
support data sharing among researchers during the research process, and to
promote publishing or archiving data and metadata to discipline-specific data
centers. The workflows hosted in these repositories are tagged and categorized
based upon their features and functionalities. The ``myexperiment" repository
is especially notable for its adoption of social web paradigm where users not
only find and reuse the workflows but can actively collaborate on scientific
experiments over the web.

### 2014-06-02 paper.tex
conclusions

@@
 \input{text/usability}
-
-\section{Summary and Conclusions} \label{sec:sum}
-% \begin{sidewaysfigure}
-% \begin{center}
-% \includegraphics[width=\linewidth]{2/figures/pdf/Taxonomy}
-% \caption{A Taxonomy of the Features of Scientific Workflow Environments}
-% \label{fig:taxonomy}
-% \end{center}
-% \end{sidewaysfigure}
-
-%\begin{sidewaysfigure}
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[]
- \path[mindmap,concept color=darkgray,text=white,font=\sf]
- node[concept] {Scientific Workflows}
- [clockwise from=0]
- child[concept color=darkgray,font=\sf] {
- node[concept] {scalability}
- [clockwise from=90]
- child [concept color=gray,font=\sf] {
- node[concept] {\small{parallelism}}
- [clockwise from=120]
- child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{data}}
- child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{pipeline}}}
-	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{task}}}
-% 	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\small{activity}}}
- }
- child[concept color=gray,font=\sf] { node[concept] {\scriptsize{optimization}} }
- }
- child[concept color=darkgray,font=\sf] { node[concept] {\small{data management}}}
- child[concept color=darkgray,font=\sf] {
- node[concept] {\small{expressiveness}}
- [clockwise from=-45]
- child[concept color=gray,font=\sf] { node[concept] {iterators} }
- child[concept color=gray,font=\sf] { node[concept] {\small{array programming}} }
- child[concept color=gray,font=\sf] { node[concept] {\small{control structures}} }
- }
- child[concept color=darkgray,font=\sf] {
- node[concept] {usability}
- [clockwise from=-120]
- child[concept color=gray,font=\sf] { node[concept] {\scriptsize{composition}} }
- child[concept color=gray,font=\sf] { node[concept] {extendible} }
- child[concept color=gray,font=\sf] { node[concept] {reusable} }
- }
- child[concept color=darkgray,font=\sf] { node[concept] {interface to DCIs} };
-\end{tikzpicture}
-\caption{A Taxonomy of the Features of Scientific Workflow Environments}
-\label{fig:taxonomy}
-\end{center}
-\end{figure}
-
-Figure \ref{fig:taxonomy} shows a taxonomy of SWE features studied in this
-paper. We studied twenty SWEs from different scientific fields developed
-worldwide. Some trends as observed in the approaches of the SWEs development
-are noted below:
-
-\begin{itemize}
- \item A recent surge in distributed computing approach of Cloud Computing has
-     given rise to adoption of new enactment technologies.
-
- \item SWEs tend to offer complementary properties which has led to several
-     collaborations among the developing groups. Notable among these are the
-     recent ones between Taverna and Galaxy \cite{mohamed-alaa-etal:2010} and
-     Kepler and Pegasus \cite{mandal-deelman-etal:2007}.
-
- \item Collection oriented SWEs wherein arrays of data are treated as
-     first-class entities has been seen a significant trend among the
-     contemporary SWEs.
-
- \item A trend of increasing number of SWEs providing a script-based workflow
-     composition interface solely or in combination with a visual interface in
-     the form of GUI has been observed.
-
- \item Workflow systems originally developed for business domains have been
-     introduced to scientific domains, such as BPEL and Trident.
-
-\end{itemize}
-
-In the present work we explore different dimensions of SWE paradigm. We study a
-number of SWE implementations. While we see a number of implementations
-dominant in areas suitable to their targeted applications, there is no single
-SWE that could sufficiently fulfill the objectives of this work. There is a
-lack of an SWE addressing the workflow expressibility and parallelism
-requirements while at the same time targeting enactment within the Grid
-Environment. Key features such as effective application data combinations or
-array-based execution coupled with Grid centric execution are either missing or
-partially available in the studied workflows.
-
-However, so far, from our study, we find Taverna as being the closest SWE that
-can be a suitable candidate for our enactments. With this motivation, we get
-into studying the details of the workings of the Taverna Workflow Environment.
-
-While many contemporary SWEs are studied in this survey, we are aware that this is not an exhaustive list.
-Some other SWEs have been missed or omitted, notable among them being: ICENI \cite{furmento-mayer-etal:2002}, Calcium \cite{caromel-henrio-etal:2008} and YML \cite{delannoy-emad-etal:2006}.
-
+%
+\input{text/concl}
 % Bibliography

### 2014-06-02 text/concl.tex
conclusions

\section{Summary and Conclusions} \label{sec:sum}
% \begin{sidewaysfigure}
% \begin{center}
% \includegraphics[width=\linewidth]{2/figures/pdf/Taxonomy}
% \caption{A Taxonomy of the Features of Scientific Workflow Environments}
% \label{fig:taxonomy}
% \end{center}
% \end{sidewaysfigure}

%\begin{sidewaysfigure}
\begin{figure}
\begin{center}
\begin{tikzpicture}[]
 \path[mindmap,concept color=darkgray,text=white,font=\sf]
 node[concept] {Scientific Workflows}
 [clockwise from=0]
 child[concept color=darkgray,font=\sf] {
 node[concept] {scalability}
 [clockwise from=90]
 child [concept color=gray,font=\sf] {
 node[concept] {\small{parallelism}}
 [clockwise from=120]
 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{data}}
 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{pipeline}}}
	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\scriptsize{task}}}
% 	 child[concept color=lightgray,text=darkgray,font=\sf] {node[concept]{\small{activity}}}
 }
 child[concept color=gray,font=\sf] { node[concept] {\scriptsize{optimization}} }
 }
 child[concept color=darkgray,font=\sf] { node[concept] {\small{data management}}}
 child[concept color=darkgray,font=\sf] {
 node[concept] {\small{expressiveness}}
 [clockwise from=-45]
 child[concept color=gray,font=\sf] { node[concept] {iterators} }
 child[concept color=gray,font=\sf] { node[concept] {\small{array programming}} }
 child[concept color=gray,font=\sf] { node[concept] {\small{control structures}} }
 }
 child[concept color=darkgray,font=\sf] {
 node[concept] {usability}
 [clockwise from=-120]
 child[concept color=gray,font=\sf] { node[concept] {\scriptsize{composition}} }
 child[concept color=gray,font=\sf] { node[concept] {extendible} }
 child[concept color=gray,font=\sf] { node[concept] {reusable} }
 }
 child[concept color=darkgray,font=\sf] { node[concept] {interface to DCIs} };
\end{tikzpicture}
\caption{A Taxonomy of the Features of Scientific Workflow Environments}
\label{fig:taxonomy}
\end{center}
\end{figure}

Figure \ref{fig:taxonomy} shows a taxonomy of SWE features studied in this
paper. We studied twenty SWEs from different scientific fields developed
worldwide. Some trends as observed in the approaches of the SWEs development
are noted below:

\begin{itemize}
 \item A recent surge in distributed computing approach of Cloud Computing has
     given rise to adoption of new enactment technologies.

 \item SWEs tend to offer complementary properties which has led to several
     collaborations among the developing groups. Notable among these are the
     recent ones between Taverna and Galaxy \cite{mohamed-alaa-etal:2010} and
     Kepler and Pegasus \cite{mandal-deelman-etal:2007}.

 \item Collection oriented SWEs wherein arrays of data are treated as
     first-class entities has been seen a significant trend among the
     contemporary SWEs.

 \item A trend of increasing number of SWEs providing a script-based workflow
     composition interface solely or in combination with a visual interface in
     the form of GUI has been observed.

 \item Workflow systems originally developed for business domains have been
     introduced to scientific domains, such as BPEL and Trident.
\end{itemize}

In the present work we explore different dimensions of SWE paradigm. We study a
number of SWE implementations. While we see a number of implementations
dominant in areas suitable to their targeted applications, there is no single
SWE that could sufficiently fulfill the objectives of this work. There is a
lack of an SWE addressing the workflow expressibility and parallelism
requirements while at the same time targeting enactment within the Grid
Environment. Key features such as effective application data combinations or
array-based execution coupled with Grid centric execution are either missing or
partially available in the studied workflows.

However, so far, from our study, we find Taverna as being the closest SWE that
can be a suitable candidate for our enactments. With this motivation, we get
into studying the details of the workings of the Taverna Workflow Environment.

While many contemporary SWEs are studied in this survey, we are aware that this
is not an exhaustive list. Some other SWEs have been missed or omitted, notable
among them being: ICENI~\cite{furmento-mayer-etal:2002},
Calcium~\cite{caromel-henrio-etal:2008} and YML~\cite{delannoy-emad-etal:2006}.

### 2014-06-02 paper.tex
hpc vs htc

@@
 \IncMargin{-\parindent}
-
+\setlength\parindent{0pt}
 \usepackage{setspace}
@@
 \end{abstract}
-
 %\category{C.2.2}{Computer-Communication Networks}{Network Protocols}
-
 %\terms{Design, Algorithms, Performance}
-
 %\keywords{Wireless sensor networks, media access control, multi-channel, radio interference, time synchronization}
@@
 
-Author's addresses: Ketan Maheshwari, Argonne National Laboratory; Johan Montagnat, I3S, Sophia Antipolis
+Author's addresses: Ketan Maheshwari, Argonne National Laboratory; Johan Montagnat, I3S, Sophia Antipolis.
 \end{bottomstuff}
@@
 % Pegasus
-% Triana
+% Triana % not sure
 % Galaxy
@@
 % Kepler/Ptolemy
-% Dryad
-% Science pipes
 % Spark
@@
 
-\section{Introduction}
-\label{sec:intro}
-\vspace*{0.5cm}\hspace{1cm} Scientific Workflow Environments (SWE)
-are, in principle similar to the traditional programming environments
-(programming language, interpreter or compiler, and runtime) albeit with a
-larger scale of data and a higher abstraction of computation. The similarity is
-such that scripting languages are sometimes advocated for programming at this
-level. However, traditional scripts do not deal with parallelism and dataflow
-aspects of workflow paradigm which makes them a strong contenders for the
-purpose of high-level application programming. Computation in science is
-dominated by computation for data analysis via one or more mathematical or
-scientific model execution. These models of computation are composed of
-multiple stages where each stage executes a program requiring input parameters
-and data while producing results in the form of data which in turn acts as
-input to the next stage. SWEs can ease these activities by supporting the
-expression, execution and automation.
-% Goals and benefits of a workflow management system
-The goals and benefits of an SWE are as follows:
-
-\begin{enumerate}
-\item \textbf{Bulk Work.}
-Workflows are important because they enable to perform computational work in
-bulk. This could be bulk data processing or running multiple distinct
-processing tasks on a single set of data. In addition to support such
-work, workflow environments provide oppurtunities for giving structure and
-order to a set of tasks and often with sufficient expressivity, choices and
-selections to carry out a bulk of work selectively. Some of the examples of
-this approach are \emph{parameter sweeps} where tasks are run over a series of
-parameter values, \emph{ensemble studies} where tasks are carried out over
-discrete groups of parameters over a space.
-
-\item \textbf{User Interaction.} Adaptation to new methods of computing is not
-easy for domain science users. Such users often need special assistance in
-order to translate their computation requirements into concrete computing
-entities. SWEs are an ideal platform enabling users to perform many tasks
-conveniently. These tasks include visually formulating an analysis, composing
-and marshalling of processes, browsing and selecting resulting data,
-visualizing and sharing final results.
-
-\item \textbf{Resource Access and Utilization.} Computational platforms and
-access modalities change rapidly with upgradation of existing technologies and
-adaptation of new techniques. Clouds are an example of new techniques
-of computational platform access while GPGPUs are example of upgradation of
-existing processing environments. Additionally, with the advent of fast and
-dedicated networks such as ESNet~\cite{esnet} and federated remote
-visualization resources are gaining prominence among scientific users. SWEs
-play important role as platforms to give transparent and hassle free access to
-such novel and multiple environments to users by hiding the technicalities of
-access methods and usage and configuration related nuances.
-
-\item \textbf{Repeatable Execution.} Scientific users often require a
-repeatable execution of same application(with same parameters and input data)
-for reproduction of results and demonstration of methods. SWEs can record and
-save workflows, associated parameters, input data and historical results to
-aid in such activities. Additionally, SWEs facilitate an incremental
-development of these methods leading to augmentation of past efforts to current
-and future developments.
-
-\item \textbf{Provenance and MetaData.} Historical data and metadata has been
-crucial entities in many research fields. They help connecting results from
-a multitude of research points of view. A provenance trace of a computation can
-help checkpoint methods and reproduce and modify computation fully or
-partially. Automated observations of deviations in scientific phenomena from
-provenance data is one of the key triggers for new scientific discoveries.
-Fields of studies involving genomic and protiens are examples of domains where
-provenance and metadata plays key roles. SWEs can be integrated or extended
-with metadata and provenance techniques rendering powerful computing and
-traceability features. In fact, most SWEs of today have separate modules or
-extensions for provenance and metadata management.
-
-\item \textbf{Sharing and Collaboration.} SWEs are often targetted
-towards a particular community of users. Dissemination of methods and results
-to communities and collaborators are some of the activities that SWEs make
-convenient for end-users. For instance, the
-Galaxy~\cite{blankenberg-vonkuster-etal:2010} workflow system is targetted
-specifically for the bioinformatics users encouraging them to collaborate over
-problems and share results. Another such example is Taverna~\cite{taverna2013}
-which is geared towards sharing and collaborative workflow oriented scientific
-computation enabling users to discover new services introduced to a community.
-Sharing and collaboration of applications and related data have become
-ubiquitious with the rise of social web. Initiatives such as
-\emph{myExperiment}~\cite{deroure-goble-etal:2008} and recently
-\emph{gitHub}~\cite{github-web} have enabled easy sharing and collaboration of
-reproducible workflows on the social web.
-
-\end{enumerate}
-
-The above described goals and benefits also form a basis for the need for a
-SWE. The need for the use of SWE can arise as a result of one or more of such
-requirements. Traditional programming platforms are not suitable to address
-these requirements as they operate at a different level of abstraction. SWEs
-differ in many respects among themselves as well as with respect to traditional
-programming environments. Consequently, a choice of the suitable SWE must be
-made by a researcher. The following list identifies the major differences
-between traditional and the SWE programming paradigm:
-
-\begin{enumerate}
-%
-\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
-extensive usage of distributed resources than conventional programs. Owing
-to the large-scale experiments, for which scientific users turn to automation,
-their demands for computational power often exceeds that of a single personal
-computer. These properties make them ideal candidates for a large-scale
-distributed computing environment. The case for suitability of SWEs on
-large-scale distributed infrastructures has been well-argued upon
-in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
-ensure scalable and optimized performance is vital. Additionally, most SWEs are
-capable of executing workflows at small scale with minimal tuning. This
-multi-scalability enables quick and easy conceptualization and prototyping of
-new workflows.
-%
-\item \textbf{Data Management.} The scale of data transfer is much higher in
-SWEs from one execution site to others as well as among storage resources while
-it is lower (ranging from variables to small data structures) in programs.
-Owing to such scales, applications often need access to special data movement
-and access and movement protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
-shipped with capabilities to handle such protocols.
-%
-
-\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
-Supporting an interface to a distributed system is a complex task from the
-point of view of an SWE. The distributed nature of resource utilization and
-asynchronous execution of workflow requires a set of interdependent parameters
-to be configured correctly. Modern computing infrastructures such as Clouds,
-GPUs and supercomputers with hybrid nodes (CPU + GPGPU) pose new challenges of
-interface development.  As SWEs involve running many tasks over an
-infrastructure it becomes crucial to interface with these systems such that the
-tasks are scheduled in an optimal manner. Many SWEs implement
-\emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
-reserve resources via pseudo-tasks that act as placeholders for real tasks.
-Additionally, decision to manage working data storage on disc, memory and cache
-are some of the distinctions between a traditional programming and workflow
-platforms in the context of distributed and scalable computation.
-
-\item \textbf{Expressiveness.} Programming languages are inherently control
-flow oriented where data flow is a secondary aspect. In fact, the
-programmer must explicitly make sure that correct inputs are available
-to instructions and are in the right format. A program proceeds as the
-execution of its statements from beginning to end and completes its
-execution in a predefined sequence. However, in SWEs, the flow is often
-determined by the data availability for execution. This model of computation is
-known as dataflow computing. A workflow thus does not follow a fixed order of
-execution. Different execution sequences are possible which are dominated by
-dataflow. The unit of work in SWEs is at a higher level of granularity. A unit
-of execution is coarse-grained, and often termed as ``processor'', ``task'',
-``service'' or ``actor'' in the SWE community. This notion of high-level
-granularity drives the depth to which a workflow can be expressed in a given
-language form. In the rest of the paper, we refer to this unit of execution of a
-workflow system as \emph{processor}. On the other hand, a unit of execution in
-most programming languages is a much simpler and finer executable statement
-that accomplishes a small task. This presents a whole new set of
-requirements for expressing the processors and how they should interact with
-computational resources and operate upon data.
-
-%
-\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
-experts to utilize the computational power. This makes it essential for
-them to have a friendly, and intuitive user-interface for novice and
-non-expert users. For this reason, SWE user interfaces are dominated by the
-visual paradigm of programming. On the other hand, as the sophistication of
-expression and user expertise increases, the visual environment temds to get
-limited in capabilities it can offer. Thus, most SWEs have a text-based
-composition environment albeit with visually assisting programming environments
-such as IDEs. However, these textual environments are seldom to the extent that
-a non-expert user can graphically compose programs. While a graphical interface
-enhances usability, a text-based workflow environment can arguably give better
-control to an experienced user. Consequently, many workflow systems offer a
-dual interface of graphical and text medium to compose workflows.
-
-\end{enumerate}
-
-We concede that the above is not an exhaustive list of differences in features.
-However, it is reasonably inclusive and we believe it covers the breadth of the
-aspects we discuss in the current work. We do not include other interesting SWE
-features such as application predictability in this analysis as they are out of
-scope of the objectives of the current paper.
-
+\input{text/intro}
 \input{text/concepts}

### 2014-06-02 text/concepts.tex
hpc vs htc

@@
 study--Workflows and Scientific Workflow Environments.
+
+\subsection{Distinction with Traditional Programming Environments}
+SWEs differ in many respects among themselves as well as with respect to traditional
+programming environments. Consequently, a choice of the suitable SWE must be
+made. The following list identifies the major differences
+between traditional and the SWE programming paradigm:
+
+\begin{enumerate}
+%
+\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
+extensive usage of distributed resources than conventional programs. Owing
+to the large-scale experiments, for which scientific users turn to automation,
+their demands for computational power often exceeds that of a single personal
+computer. These properties make them ideal candidates for a large-scale
+distributed computing environment. The case for suitability of SWEs on
+large-scale distributed infrastructures has been well-argued upon
+in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
+ensure scalable and optimized performance is vital. Additionally, most SWEs are
+capable of executing workflows at small scale with minimal tuning. This
+multi-scalability enables quick and easy conceptualization and prototyping of
+new workflows.
+%
+
+\item \textbf{Performance and Throughput.} While programming platforms for
+large-scale computing exist mostly as HPC (high-performance computing)
+domain systems such as MPI, OpenMP, CUDA, etc. the dominant area for SWEs in
+this field is HTC (high-throughput computing). The main underlying difference
+here is the granularity of a task. HPC platforms run tasks at a finer
+granularity such as a CUDA thread or an MPI process where as SWEs run a task at
+a coarser granularity such as an invocation of a legacy executable.
+
+\item \textbf{Data Management.} The scale of data transfer is much higher in
+SWEs from one execution site to others as well as among storage resources while
+it is lower (ranging from variables to small data structures) in programs.
+Owing to such scales, applications often need access to special data movement
+and access and movement protocols such as GridFTP, Amazon S3, NAS, etc. SWEs are often
+shipped with capabilities to handle such protocols.
+%
+
+\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
+Supporting an interface to a distributed system is a complex task from the
+point of view of an SWE. The distributed nature of resource utilization and
+asynchronous execution of workflow requires a set of interdependent parameters
+to be configured correctly. Modern computing infrastructures such as Clouds,
+GPUs and supercomputers with hybrid nodes (CPU + GPGPU) pose new challenges of
+interface development.  As SWEs involve running many tasks over an
+infrastructure it becomes crucial to interface with these systems such that the
+tasks are scheduled in an optimal manner. Many SWEs implement
+\emph{pilot-jobs}~\cite{montagnat-glatard-etal:2010,ucc2011} in order to
+reserve resources via pseudo-tasks that act as placeholders for real tasks.
+Additionally, decision to manage working data storage on disc, memory and cache
+are some of the distinctions between a traditional programming and workflow
+platforms in the context of distributed and scalable computation.
+
+\item \textbf{Expressiveness.} Programming languages are inherently control
+flow oriented where data flow is a secondary aspect. Often, the
+programmer must explicitly make sure that correct inputs are available
+to instructions and are in the right format. A program proceeds as the
+execution of its statements from beginning to end and completes its
+execution in a predefined sequence. However, in SWEs, the flow is often
+determined by the data availability for execution. This model of computation is
+known as dataflow computing. A workflow thus does not follow a fixed order of
+execution. Different execution sequences are possible which are dominated by
+dataflow. The unit of work in SWEs is at a higher level of granularity. A unit
+of execution is coarse-grained, and often termed as ``processor'', ``task'',
+``service'' or ``actor'' in the SWE community. This notion of high-level
+granularity drives the depth to which a workflow can be expressed in a given
+language form. In the rest of the paper, we refer to this unit of execution of a
+workflow system as \emph{processor}. On the other hand, a unit of execution in
+most programming languages is a much simpler and finer executable statement
+that accomplishes a small task. This presents a whole new set of
+requirements for expressing the processors and how they should interact with
+computational resources and operate upon data.
+
+%
+\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
+experts to utilize the computational power. This makes it essential for
+them to have a friendly, and intuitive user-interface for novice and
+non-expert users. For this reason, SWE user interfaces are dominated by the
+visual paradigm of programming. On the other hand, as the sophistication of
+expression and user expertise increases, the visual environment temds to get
+limited in capabilities it can offer. Thus, most SWEs have a text-based
+composition environment albeit with visually assisting programming environments
+such as IDEs. However, these textual environments are seldom to the extent that
+a non-expert user can graphically compose programs. While a graphical interface
+enhances usability, a text-based workflow environment can arguably give better
+control to an experienced user. Consequently, many workflow systems offer a
+dual interface of graphical and text medium to compose workflows.
+
+\end{enumerate}
+
+We concede that the above is not an exhaustive list of differences in features.
+However, it is reasonably inclusive and we believe it covers the breadth of the
+aspects we discuss in the current work. We do not include other interesting SWE
+features such as application predictability in this analysis as they are out of
+scope of the objectives of the current paper.
 

### 2014-06-03 text/concepts.tex
hpc vs htc

@@
 interest for the present work. A \textit{user interface} interacts with the
-\textit{workflow engine} using commands and messages on the inside while with
+\textit{workflow system} using commands and messages on the inside while with
 user on the outside. The \textit{workflow language} becomes an intermediate
 representation for communication between user interface and the
-\textit{workflow engine}. The workflow engine acts as a compiler/interpreter
+\textit{workflow system}. The workflow system acts as a compiler/interpreter
 and a high-level scheduler for the processors and interfaces with the
-underlying \textit{distributed resources}. It resolves data dependencies and
+underlying \textit{compute and storage resources}. It resolves data dependencies and
 optimizes the flow at runtime and monitors the workflow progress and collects
@@
 perceived and visualized. The MA DAG~\cite{caron-desprez:2005}, engine is one
-example of a DAG generator engine. MA DAG has evolved from a regular DAG
+example of a DAG generator. MA DAG has evolved from a regular DAG
 generator to a more abstract DAG generator laced with control structures. MA
@@
 \begin{enumerate}
-\item \textbf{Multisite execution.} Interfacing to existing computation and
-storate infrastructure is not trivial, not to mention the new and
-unforeseen access and compute modes. SWEs are yet to formulate a
-generalized interface towards the major platforms such that they will
-run on them out of the box.
+\item \textbf{Multisite and Manysite execution.} Interfacing to existing computation and
+storate infrastructure is not trivial, not to mention the new and unforeseen
+access and compute modes. SWEs are yet to formulate a generalized interface
+towards the major platforms such that they will run on them out of the box.
+Despite having interfaces built for more than one infrastructure, it is hard to
+optimally schedule large workflows to many remote sites simultaneously. The
+problem of limited control over remote sites and distributing interdependent
+data and tasks makes it non-trivial to efficiently load-balance work among
+them.
 
-\item \textbf{Manysite execution.} Even if a SWE has interfaces built for
-more than one infrastructure, it is hard to optimally schedule large
-workflows to many remote sites simultaneously. The problem of limited
-control over remote sites and distributing interdependent data and
-tasks makes it non-trivial to efficiently load-balance work among them.
-
-\item \textbf{Expressivity.} Workflow languages have evolved over the years
+\item \textbf{Language.} Workflow languages have evolved over the years
 with reasonable expressivity being supported. However, they are often

### 2014-06-04 text/galaxy.tex
dded a sentence about galaxy

@@
 \paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
-SWE developed at the Penn State University, Pennsilvania. It provides a
-webportal based environment to build complex pipelines from an available
-repository of tools (\textit{http://g2.bx.psu.edu}).  Galaxy is currently used
-mainly by bioinformatics community. Users can extend the tool repository
-provided by Galaxy using their own Galaxy server. One of the major use of
-Galaxy is to achieve a mix and match of data analysis tools in order to analyze
-a vast repository of genome based database. Galaxy offers a web-based,
-interactive platform for data analysis via reusable and reproducible workflows.
-While excelling in the usability aspects, the Galaxy environment has a limited
-parallelism capabilities and a small number of existing implementations for
-interfacing with diverse, scalable computing infrastructures. The user can draw
-workflows in which multiple copies of a task are created to operate on multiple
-inputs at once, but this is not a scalable approach.
+SWE developed at the Penn State University, Pennsylvania. It provides a
+web-based portal environment to build complex pipelines from an available
+repository of processors called tools within Galaxy project
+(\textit{http://g2.bx.psu.edu}).  Galaxy is currently used mainly by
+bioinformatics community. Users can contribute and extend the tool repository
+provided by Galaxy using a hosting server. One of the major use of Galaxy is to
+achieve a mix and match of data analysis tools in order to analyze a vast
+repository of genome based database. Galaxy offers a web-based, interactive
+platform for data analysis via reusable and reproducible workflows.  While
+excelling in the usability aspects, the Galaxy environment has a limited
+parallelism expression capabilities and a small number of existing
+implementations for interfacing with diverse, scalable computing
+infrastructures. For instance, it is hard to express the commonly found
+\texttt{foreach} parallel expression using Galaxy's visual interface. The user
+can draw workflows in which multiple copies of a task are created to operate on
+multiple inputs at once. However, workflows tend to get cluttered quickly and
+is not scalable to large-scale computations.
